---
title: 'Atomic Vectors II'
author: "Nicholas J. Gotelli"
date: "30 January 2025"
output:
  html_document:
    highlight: tango
    keep_md: no
    theme: united
  pdf_document: default
---

### Creating Vectors

#### Creating an empty vector and adding to it (avoid)

```{r, eval=FALSE}
# create an empty vector, specify mode & length
z <- vector(mode="numeric",length=0)  
print(z)

# now add elements to it

z <- c(z,5)
print(z)
# This "dynamic sizing" is very SLOW!

```


#### Creating a vector of predefined length (preferred)

#### Better to create a vector of pre-defined length

```{r, eval=FALSE}
# has 100 0s in a numeric vector
z <- rep(0,100) 
head(z)

# but would be better to start with NAs
z <- rep(NA,100)
head(z)

# but what is its type?
typeof(z)

# take advantage of coercion!
z[1] <- "Washington"
head(z)
typeof(z)
```

### Efficiently creating a vector with many names
```{r, eval=FALSE}
my_vector <- runif(100)
my_names <- paste("Species",seq(1:length(my_vector)),sep="")
names(my_vector) <- my_names
head(my_vector)
str(my_vector)
```

### Using rep to repeat elements and create vectors

```{r, eval=FALSE}
# rep for repeating elements
rep(0.5,6) # give the element (or vector!) and number of times to repeat
rep(x=0.5,times=6) # using the argument names is always prudent
rep(times=6,x=0.5) # with argument names, order is not important
my_vec <- c(1,2,3)
rep(x=my_vec,times=2) # applies to entire vectors
rep(x=my_vec,each=2) # repeat each element individually
rep(x=my_vec,times=my_vec) # what does this do?
rep(x=my_vec,each=my_vec) # and this?
```

### Using seq to create regular sequences

```{r, eval=FALSE}
seq(from=2, to=4) # set limits for integer sequences
2:4 # very common short cut with no explicit function wrappers
seq(from=2,to=4,by=0.5) # use a by function can generate real numbers
x <- seq(from=2, to=4,length=7) # sometimes easier to just specify the length
my_vec <- 1:length(x) # commonly used, but actually slow
print(my_vec)
seq_along(my_vec) # much faster for models and big data!
seq_len(5) # this is also faster than 1:5
```

### Using rnorm and runif to create vectors of random numbers
```{r, eval=FALSE}
runif(5) # 5 random uniform values betweeon 0 and 1
runif(n=3,min=100, max=101) # 3 random uniform values between 100 and 101

rnorm(6) # 6 random normal values with mean 0 and standard deviation 1
rnorm(n=5, mean=100, sd=30) # 5 random normal values with mean 100 and sd 30

## Explore distributions by sampling and plotting
library(ggplot2) # do this at the very start
z <- runif(1000) #default uniform (0,1)
qplot(x=z)
z <- rnorm(1000) #default normal (0,1)
qplot(x=z)
```

### Using sample to draw random values from an existing vector

```{r, eval=FALSE}
long_vec <- seq_len(10)
typeof(long_vec)
str(long_vec)

sample(x=long_vec) # with no other params, this reorders the vector
sample(x=long_vec, size=3) # specify a number (sampling without replacement)
sample(x=long_vec,size=16,replace=TRUE) # can generate duplicates
my_weights <- c(rep(20,5),rep(100,5)) # create a set of non-zero positive weights (integer or real)
print(my_weights)
sample(x=long_vec,replace=TRUE,prob=my_weights) # sampling with replacement and weights
sample(x=long_vec,replace=FALSE,prob=my_weights) # sampling without replacement and weights
sample(x=long_vec,replace=TRUE,prob=long_vec) # what does this do?
```

### Techniques for subsetting atomic vectors

```{r,eval=FALSE}
z <- c(3.1, 9.2, 1.3, 0.4, 7.5)

# positive index values
z[c(2,3)]

# negative index values to exclude elements
z[-c(2,3)]

# create a vector of logical elements to select conditions

z[z<3]

# equivalent to the following
tester <- z<3
print(tester)
z[tester]

# also use which() function to find subscript indicators
which(z<3)

# this works, but is overkill; just use the boolean operator
z[which(z<3)]

# can also use length() for relative positioning to last element
z[-(length(z):(length(z)-2))]

# also can subset using named vector elements
names(z) <- letters[1:5]
z[c("b","c")]


```

### Relational operators in R

```{r eval=FALSE}
# <   less than
# >   greater than
# <=  less than or equal to
# >=  greater than or equal to
# ==  equal to
```

### Logical operators
```{r eval=FALSE}
# ! not
# & and (vector)
# | or (vector)
# xor(x,y)
# && and (only for first element)
# || or (only for first element)

x <- 1:5
y <- c(1:3,7,7)
x == 2
x != 2
x == 1 & y == 7
x == 1 | y == 7
x == 3 | y == 3
xor(x==3, y==3)
x == 3 && y == 3
```

### Subscripting with missing values

```{r, eval=FALSE}
set.seed(90)
z <- runif(10) # simple integer sequence
print(z)

z < 0.5 # create logical vector
z[z < 0.5] # use as index call
which(z < 0.5) # use to get indices for logical
z[which(z < 0.5)] # does same as above

zD <- c(z,NA,NA) # contaminate it
zD[zD < 0.5] # NA values carried along!
zD[which(zD < 0.5)] # NA values dropped

```

