---
title: "dplyr_SQL_Lecture"
author: "George Ni"
date: "2023-02-27"
output:
  html_document:
    theme: cerulean
---


#SQL stands for Structured Query Language, and it's useful for storing and processing datasets. This is often used in a lot of other applications, for example I use it a lot when running GIS applications like ArcGIS Pro or ArcMaps, which usually doesn't have a dedicated way to interact with datasets except with SQL.

#We're going to be using the site_by_species.csv and site_by_variables.csv. The first has every row being a site, and the presence/absence of each species as a separate column. The second has every row being a site, and the columns being various environmental conditions/predictors.

```{r message=FALSE}
#Installing Packages
library(sqldf)
library(dplyr)
library(tidyverse)
```

```{r echo=FALSE}
#Read in datafiles
species_clean<-read.csv("site_by_species.csv") 

var_clean<-read.csv("site_by_variables.csv")
```

# Take a look at the datasets first

```{r, eval=FALSE}
head(species_clean)

head(var_clean)
```
# Start with operations using 1 file

## How to subset rows

dplyr Method:
```{r}
#Dplyr method-uses the filter() method
species<-filter(species_clean, Site<30)
var<-filter(var_clean, Site<30)
```

SQL Method
```{r}
#SQL Method
query="
SELECT Site, Sp1, Sp2, Sp3 
FROM species
WHERE Site<'30'
"
sqldf(query)
```

# How to subset columns
## Dplyr method-uses select() function, which can use either column # or name

dplyr method
```{r}
edit_species<-species%>%
  select(Site, Sp1, Sp2, Sp3)
edit_species_2<-species%>% #Using column #s doesn't require any indicators
  select(1, 2, 3, 4)

```

SQL Method
```{r eval=FALSE}
#SQL method-
# query entire table
query ="
SELECT *
FROM species
"
a=sqldf(query) # save results to data frame

sqldf(query) # dump to console
```

# Specify columns
```{r}
query ="
SELECT Site, Sp1, Sp2, Sp3
FROM species
"
sqldf(query)
```


```{r}
# reorder columns
query ="
SELECT Sp1, Sp2, Sp3, Site
FROM species
"
sqldf(query)

```


If we want to rename columns, we can just use the rename() function


Dplyr method

```{r}
species<-rename(species, Long=Longitude.x., Lat=Latitude.y.)
```

# Pull out all the numerical columns

dplyr method
```{r}
num_species<-species%>%
  mutate(letters=rep(letters, length.out=length(species$Site)))
num_species<-select(num_species, Site, Long, Lat, Sp1,letters)
num_species_edit<-select(num_species, where(is.numeric))

```



# Pivot_Longer

Pivot_longer lengthens the data, decreasing the number of columns, and increases the number of rows. Can use either pivot_longer or gather, but gather is outdated

```{r}
species_long<-pivot_longer(edit_species, cols=c(Sp1, Sp2, Sp3), names_to="ID")
```


# Pivot_wide

Pivot_wide goes from long to wide, widens the data, increasing the number of columns, decreases the number of rows. Can use pivot_wider or spread

```{r}
species_wide<-pivot_wider(species_long, names_from=ID)
```



Aggregating
SQL Method


# SQL Method
## aggregate and give counts of ObjectType
```{r}
query="
SELECT SUM(Sp1+Sp2+Sp3)
FROM species_wide
GROUP BY SITE
"
sqldf(query)
```

Aggregate and give counts with new variable name 
```{r}
query="
SELECT SUM(Sp1+Sp2+Sp3) AS Occurence
FROM species_wide
GROUP BY SITE
"
sqldf(query)
```


Mutating/Adding columns
Dplyr method-uses mutate function()-we've already covered this!


SQL Method
```{r}
#SQL method
query="
ALTER TABLE species_wide
ADD new_column VARCHAR
"
sqldf(query)

```


# Re-ordering columns 
## Dplyr method-using the select() method
```{r}
reorder<-select(species, Sp1, Sp2, Site)
```


SQL Method
reorder columns
```{r}
query ="
SELECT Sp1, Sp2, Sp3, Site
FROM species
"
sqldf(query)
```



# 2 File Operations

Here, we are going to interact with 2 data files (.csv's), and try to gather them into a usable form. Often times people will store different variables/types of data into different files, just to maintain organization-sometimes this is good, but others, we may need want to combine them, in which case we "join" them together. 

Joins are going to be different from binding-joining requires the data files to have at least one similar column so that it knows how to combine them, and will order it accordingly,while binding just immediately pastes the rows/columns together, and by position, not by anything meaningful. So, use those with that warning-there is a high chance for error when using bind functions, as any minor error in that could throw off the entire dataset.


# Left/Right/Union Joins

Let's first start with clean files-Reset the species and var variables, and let's filter them to a manageable size.

```{r}
edit_species<-species_clean%>%
  filter(Site<30)%>%
  select(Site, Sp1, Sp2, Sp3, Sp4, Longitude.x., Latitude.y.)

edit_var<-var_clean%>%
  filter(Site<30)%>%
    select(Site, Longitude.x., Latitude.y., BIO1_Annual_mean_temperature, BIO12_Annual_precipitation)
```


```{r}
#Dplyr method
#Left_join-a left join basically means you are stitching the matching rows of file B to file A-this does require that there be some matching/marker column to actually link. 
left<-left_join(edit_species, edit_var, by="Site")
head(left)

```


```{r}
#Right_join-does the opposite-joins the matching rows of file A to B. The difference here is what is lost when you match them together
right<-right_join(edit_species, edit_var, by="Site")
head(right)
```
```{r}
#Inner_join-retains rows that match between both files A and B. This one loses a lot of information if they aren't matching very well.
inner<-inner_join(edit_species, edit_var, by="Site")
```
```{r}
#Full_join-the opposite of an inner join, just retains all values, all rows, so instead of losing a lot of rows, you have a trade-off where you instead get a bunch of NA's when they're missing.
full<-full_join(edit_species, edit_var, by="Site")
```


SQL method
```{r}
query="
SELECT *
FROM edit_var
INNER JOIN edit_species ON edit_var.Site = edit_species.Site
"
x<-sqldf(query)
```

```{r}
# ------------------------------
# left join keeps all rows from the first table and adds
# data from second table for matches and NAs for mismatches (n = 500 rows = row number in Orders table)
query="
SELECT *
FROM edit_var
LEFT JOIN edit_species ON edit_var.Site = edit_species.Site
"
x<-sqldf(query)
```
