---
title: "Manipulating data using `dplyr`"
author: "Lauren V. Ash"
date: "March 1, 2023"
output: 
  html_document: 
    highlight: tango
    theme: united
editor_options: 
  chunk_output_type: console
---
<font size="3">   



### What is `dplyr`?   
+ a new(ish) package that provides a set of tools for manipulating datasets in R   
+ part of the `tidyverse`: an opinionated collection of R packages which share philosophy, grammar, and data structure   
+ specifically written to be fast!   
+ has individual functions that correspond to the most common operations    
+ makes it easier for you to figure out what to do with your data   

### The core 'verbs'
1. `filter()`   
2. `arrange()`   
3. `select()`   
4. `summarize()` and `group_by()`     
5. `mutate()`   


### Start with a built-in dataset   
```{r warning=FALSE, message=FALSE}
library(tidyverse)
data(starwars)
class(starwars)

```
### What is a tibble?    
+ "a modern take on data frames"   
+ said to keep the great aspects of data frames and drops the frustrating ones (i.e. changing variable names, changing an input type)   

```{r}
glimpse(starwars) #more effective than str() in this case   
head(starwars) #looks a little different since it's a tibble (mentions dimensions and also states variable type)
```

### Cleaning up our data   
```{r}
# complete.cases is not part of dplyr, but it is very useful!
#dataClean <- starwars[complete.cases(starwars),] # removes all rows with NA values but doesn't work with lists

# to get complete cases for just certain columns 
starwarsClean <-starwars[complete.cases(starwars[,1:3]),]


# Check for NAs
is.na(starwarsClean[1,1]) #useful for only a few observations since it returns a list of True/False
anyNA(starwarsClean)
anyNA(starwars[,1:10]) # compared to our original dataset

# What does our data look like now?
glimpse(starwarsClean) # the list columns are still there but no NAs in columns 1-10
head(starwarsClean)
```


### `filter()`: Pick/subset observations by their values 
```{r}
### uses >, >=, <, <=,!=, == (not just one!) for comparisons   
### Logical operators: & | !

# filter automatically excludes NAs, have to ask for them specifically 

filter(starwarsClean, gender == "masculine" & height < 180) # you can also use commas in place of &   

filter(starwars, gender == "masculine", height <180, height > 100) # can add multiple conditions for the same variable   

filter(starwars, eye_color %in% c("blue", "brown")) # we can use %in% for a few conditions, similar to ==
```


### `arrange()`: Reorder rows   
```{r}
arrange(starwarsClean, by = height) #default is ascending order   

arrange(starwarsClean, by = desc(height)) #we can use desc() to change that

arrange(starwarsClean, height, desc(mass)) # each additional column used to break ties with the preceding column

starwars1<- arrange(starwars, height) # missing values are at the end; note we haven't been assigning anything to a variable, just printing (until now) 
tail(starwars1)

```



### `select()`: Choose variables by their names    
```{r} 
## All of these do the same thing (subset)
starwarsClean[1:10,] # using base R
select(starwarsClean, 1:10) # you can use numbers to subset
select(starwarsClean, name:species) # you can use variables names too
select(starwarsClean, -(films:starships)) # you can subset everything except particular variables

## Rearrange columns
select(starwarsClean, name, gender, species, everything()) # using the everything() helper function is useful if you have a few variables to move to the beginning
select(starwarsClean, contains("color")) ## other helpers include: ends_with, starts_with, matches (reg ex), num_range

## Renaming columns
select(starwars, haircolor = hair_color) # select can rename but only keeps specified variable; actual name of the column comes after stating new name
rename(starwarsClean,  haircolor = hair_color) # keeps all the variables
```


### `mutate()`: Create new variables with functions of existing variables   
```{r}
# Let's create a new column that is a height divided by mass
mutate(starwarsClean, ratio = height/mass) # note we use arithmetic operators 

starwars_lbs <- mutate(starwarsClean, mass_lbs=mass*2.2) # now let's convert kg to lbs
head(starwars_lbs) # mass_lbs is now the last variable of the tibble

select(starwars_lbs, 1:3, mass_lbs, everything()) # we can bring it to the beginning using select

# If we only wanted the new variable
transmute(starwarsClean, mass_lbs=mass*2.2) # just keep the new variable

transmute(starwarsClean, mass, mass_lbs=mass*2.2) # you can mention variables you want to keep in the new dataset
```


### `summarize()` and `group_by()`: Collapse many values down to a single summary 
```{r}
summarize(starwarsClean, meanHeight = mean(height)) # gives summary statistics for entire tibble

# working with NAs
summarize(starwars, meanHeight = mean(height)) # does not calculate mean if NAs are present
summarize(starwars, meanHeight = mean(height, na.rm=TRUE), TotalNumber = n()) # n() is a useful function in determining sample size
summarize(starwarsClean, meanHeight = mean(height), number = n() ) #na.rm not necessary since we are working with a clean dataset

# use group_by for maximum usefulness
starwarsGenders<- group_by(starwars, gender)
head(starwarsGenders) # now we see the groups mentioned at the top of the tibble
summarize(starwarsGenders, meanHeight = mean(height, na.rm = TRUE), number = n() ) # now we can compare height and sample size between groups
```


### What is piping? 
+ used to emphasize a sequence of actions   
+ it lets you pass an intermediate result onto the next function (it takes the output of one statement and makes it the input of the next statement)
+ avoid when you need to manipulate more than one object at a time or there are meaningful intermediate objects   
+ formatting: should always have a space before it and usually followed by a new line (usually automatic indent)

```{r}
starwarsClean %>%
    group_by(gender) %>%
    summarize(meanHeight = mean(height, na.rm=T), number=n()) # so much cleaner! notice we lost a bunch of groups with the clean dataset
```

```{r}
## case_when() is useful for multiple ifelse statements


starwarsClean %>%
  mutate(sp = case_when(species == "Human" ~ "Human", TRUE ~ "Non-Human")) %>%
  select(name, sp, everything())

starwarsClean %>%
  group_by(films) %>%
  mutate(
    sp = case_when(species == "Human" ~ "Human", TRUE ~ "Non-Human"),
    status = case_when(str_detect(films, "A New Hope") ~ "OG", TRUE ~ "Later")) %>%
  select(name,sp, status, everything()) %>%
  arrange(status) %>%
  {starwarsOGstatus <<- .}



tail(starwarsOGstatus)
```

```{r}
## long to wide format and vice-versa
glimpse(starwarsClean)

# make data set wider
wideSW <- starwarsClean %>% 
  select(name,sex, height) %>%
  pivot_wider(names_from = sex, values_from = height, values_fill = NA)
wideSW

starwars %>%
  select(name, homeworld) %>%
  group_by(homeworld) %>%
  mutate(rn = row_number()) %>%
  ungroup() %>%
  pivot_wider(names_from = homeworld, values_from = name) %>%
  select(-rn)

## make data set longer
glimpse(wideSW)
wideSW %>%
  pivot_longer(cols= male:female,
               names_to = "sex",
               values_to= "height",
               values_drop_na = T)
  
```

</font>
