---
title: "File Batch Processing"
author: "Nicholas J. Gotelli"
date: "17 22 24 April"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval=FALSE, echo = TRUE)
```

## Set Up Project
- create a new project
- using the `upscaler` package create folders from the console in project with `add_folder()`
- inside the `CleanedData` folder, manually create a new folder called `ToyDataFiles`

## Build Function `create_toy_data_files`
- from the console, run `build_function("create_toy_data_files")
- in the Function folder, open the script `CreateToyDataFiles.R` and add code:

```{r}
# --------------------------------------
# FUNCTION create_toy_data_files
# required packages: none
# description: build set of toy .csv files
# inputs: number of files,max rows,max col
# outputs: a set of .csv files in a new subfolder
########################################
create_toy_data_files <- function(nrow=NULL,
                               ncol=NULL,
                               nfiles=NULL){
  
  # assign parameter defaults
  if (is.null(nrow) | is.null(ncol) | is.null(nfiles)){
    nrow=10
    ncol=9
    nfiles=6
  }
  
  # build file labels
  file_labels <- upscaler::create_padded_labels(n=nfiles,
                                      string="Toy_Data",
                                      suffix=".csv") 
  
  # run for loop
  
  for (i in 1:nfiles) {
    df <- as.data.frame(matrix(runif(nrow*ncol),
                               nrow=nrow,
                               ncol=ncol))
    write.table(df,file=paste("CleanedData/ToyDataFiles/",
                              file_labels[i],sep=""),
                              sep=",")
    
  }
  
  
  
  
  return()
  
} # end of function CreateToyDataFiles
# --------------------------------------
# create_toy_data_files()

```

## Build Function `crunch_data()`
- from the console, run `build_function("crunch_data")
- in the Function folder, open the script `CrunchData.R` and add code:

```{r}
# --------------------------------------
# FUNCTION crunch_data
# required packages: none
# description: operate on a single data frame
# inputs: df = dataframe
#         crunch_cols = numeric list of columns to use
#.        param_names = character list of output labels
# outputs: results = list of output values 
########################################
crunch_data <- function(df=NULL,
                        crunch_cols=NULL,
                        param_names=NULL){

# assign parameter defaults
if (is.null(df) | is.null(crunch_cols) | is.null(param_names)) {
  df <- data.frame(runif(10),runif(10),runif(10))
  crunch_cols <- list(2,3)
  param_names <- list("avg","skew","weird")
}

# function body

avg <- mean(df[,crunch_cols[[1]]])
skew <- sum(df[,crunch_cols[[1]]] - avg)^3/nrow(df)
weird <- sum(df[,crunch_cols[[1]]]) + 
  sum(df[,crunch_cols[[2]]]) + pi

results <- list(avg,skew,weird)
names(results) <- param_names

return(results)

} # end of function crunch_data
# --------------------------------------
   # crunch_data()
```

## Build Function `filebatchr()`
- from the console, run `build_function("filebatchr")
- in the Function folder, open the script `Filebatchr.R` and add code:
```{r}
# --------------------------------------
# FUNCTION filebatchr
# required packages: none
# description: use for loop for batch processing
# inputs: file_names = list of file names
#         crunch_cols = list of columns to analyze in df
#         param_names = list of names of output objects
#.        fun = name of operations function
# outputs: output_df with rows for each data file
########################################
filebatchr <- function(file_names=NULL,
                          fun=NULL,
                          crunch_cols=NULL,
                          param_names=NULL){

  if(is.null(file_names)) return(print("no input"))
  
output_df <- as.data.frame(matrix(rep(NA,length(file_names)*length(param_names)),nrow=length(file_names),ncol=length(param_names)))
names(output_df)=param_names
  
  nobs <- rep(NA,length(file_names))
  for (i in 1:length(file_names)) {
    df <- read.table(file=file_names[[i]],
                     header=TRUE,
                     sep=",")
    . <- fun(df=df,
             crunch_cols=unlist(crunch_cols),
             param_names=param_names)
    
    output_df[i,] <- .
    nobs[i] <- nrow(df)
  }
  output_df
  # add initial metadata columns (ID,filename,nobs)
  output_df <- cbind(ID=1:length(file_names),file=basename(unlist(file_names)),nobs=nobs,output_df)
  




return(output_df)

} # end of function filebatchr
# --------------------------------------
 # filebatchr(file_names=file_names,
               # fun=crunch_data,
               # crunch_cols=crunch_cols,
               # param_names=param_names)
```

## Create script `MainScript.R`
- from the root of your project directory, create a new script `MainScript.R` and save it at the root.
- fill this script with the following code:

```{r}
# filebatchr() for operating on batch files
# NJG and Spring 2025 Bio 6100 class
# 15 April 2025

library(upscaler)
library(ggplot2)
set_up_log()
source_batch(folder="Functions")
# create_toy_data_files(nrow=15,
#                       ncol=10,
#                       nfiles=8)

# create global variables
file_names <- as.list(list.files(pattern="\\.csv$",
           path="CleanedData/ToyDataFiles",
           full.names=TRUE))
crunch_cols <- list(4,5)
param_names <- list("avg","skew","weird")



# use lapply to run it all with two lines of code!
z <- lapply(file_names,read.table,sep=",")
final <- lapply(z,crunch_data)

# do the work in a for loop
output_df <- as.data.frame(matrix(rep(NA,length(file_names)*length(param_names)),nrow=length(file_names),ncol=length(param_names)))
names(output_df)=param_names
nobs <- rep(NA,length(file_names)) # empty vector for row counts
for (i in 1:length(file_names)) {
  df <- read.table(file=file_names[[i]],
                   header=TRUE,
                   sep=",")
  . <- crunch_data(df=df,
                   crunch_cols=unlist(crunch_cols),
                   param_names=param_names)

  output_df[i,] <- .
  nobs[i] <- nrow(df)
}
output_df
# add initial metadata columns (ID,filename,nobs)
output_df <- cbind(ID=1:length(file_names),file=basename(unlist(file_names)),nobs=nobs,output_df)

output_df



# filebatchr(file_names=file_names,
#                fun=crunch_data,
#                crunch_cols=crunch_cols,
#                param_names=param_names)
```

Once the toy data files have been built, and the global variables assigned at the top of the script the code can be executed in 3 different ways, with the following lines of code from the MainScript:

- with the lapply function (uses only 2 lines of code)
- with the for loop (which allows you to see the components)
- with the batchr function (which puts that for loop in a function)