---
title: 'Lecture #21: Markov Models'
author: "Nicholas J. Gotelli"
date: "April 13, 2017"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---
### Assumption of Markov Models
- set of $s$ discrete "states" for system at each time step
- discrete constant time steps
- constant probability transitions between states
- states are mutually exclusive and exhaustive
- at each time step, system stays in its current state or changes to another state
- behavior of system depends only on
    - current state
    - rules governing state changes
- system described by a vector $\textbf{n}$ of counts or frequencies of different $s$ states at time $t$
- change in $\textbf{n} at time $t + 1$ controlled by a $s \times s$ transition matrix $\textbf{A}$, which has all of the pairwise transition elements.
- change follows rules of matrix algebra:

$$\textbf{n}_{t + 1} = \textbf{An}_t$$


### Simple (Algebraic) Example

```
states of a patch: disturbed, grassland, forest
landscape of 20 patches described by vector: n <- c(10,7,3)

transition rules described by matrix A:

(state at t + 1)
disturbed               0.3    0.4  0.3
grassland               0.0    0.1  0.2
forest                  0.7    0.5  0.5
```
### Calculation Of Changes

```
How many patches are in each state at time (t + 1)?

(0.3)*10 + (0.4)*7 + (0.3)*3 = 6.7 disturbed patches
(0.7)*10 + (0.1)*7 + (0.2)*3 = 8.3 grassland patches
(0.0)*10 + (0.5)*7 + (0.5)*3 = 5.0 forest patches
```
### Using the `popbio` package
```{r}
library(popbio)
stages <- c("Disturbed", "Grassland", "Forest")
A <- matrix(c(0.3, 0.4, 0.3,
              0.0, 0.1, 0.2,
              0.7, 0.5, 0.5),
            nrow=3,
            byrow=TRUE, dimnames=list(stages,stages))
n <- c(10,7,3)
timesteps=10
print(A)
MarkovOut <- pop.projection(A=A,n=n,iterations=timesteps)
stage.vector.plot(MarkovOut$stage.vectors,col=2:4)
print(MarkovOut)
##################################################
# function: plotCounts
# plot raw counts of states at each time step
# input: matrix (rows=stages, cols=times)
# output: plot
#------------------------------------------------- 
plotCounts <- function(m=matrix) {
par(mfrow=c(1,2))

plot(x=1:ncol(m),
     y=m[1,],
     type="o",
     ylab="Frequency",
     xlab="Time",
     ylim=c(0,max(m)),
     col="red",
     lwd=2)
  for (i in 2:nrow(m)){
lines(x=1:ncol(m),
      y=m[i,],
      type="o",
      col=i+1,
      lty=i,
      lwd=2)
  }
stage.vector.plot(MarkovOut$stage.vectors,col=2:(nrow(m)+1))
par(mfrow=c(1,1))
}
plotCounts(MarkovOut$stage.vectors)

# hypothesis: proportions in patches are constant regardless of starting point
# hypothesis: constant numbers only result when column sums are all 1.0
```

### Real Example #1: Sarracenia stage-structured population growth model

```{r}
# data from Gotelli N.J. and A.M. Ellison. 2002. Nitrogen deposition and extinction risk in the northern pitcher plant, Sarracenia purpurea. Ecology 83: 2758-2765.
stages <- c("Recruit","Juvenile","NFAdult","FAdult")
timesteps=100
n=c(50,50,50,50)
A <- matrix(c(0,0,0,4,
              0.1, 0.854, 0.177, 0,
              0, 0.131, 0.708, 0.660,
              0,0,0.1,0.307),nrow=4,byrow=TRUE,dimnames=list(stages,stages))
MarkovOut <- pop.projection(A=A,n=n,iterations=timesteps)
plotCounts(MarkovOut$stage.vectors)
print(MarkovOut$lambda)              
print(MarkovOut$stable.stage)              
print(head(MarkovOut$pop.sizes))
print(tail(MarkovOut$pop.sizes))

eigen.analysis(A)
```


### Real Example #2: Ground squirrel age-structured population growth
```{r}
# data simplified from Slade N.A. and D.F. Balph. 1974. Population ecology of Uinta ground squirrels. Ecology 55: 989-1003
stages <- as.character(0:8)
timesteps=150
n= rep(20,9)
A <- matrix(c(0,0,0.646,3,2.17,0,1.054,0,2.214,
              0.662,0,0,0,0,0,0,0,0,
              0,0.501,0,0,0,0,0,0,0,
              0,0,0.756,0,0,0,0,0,0,
              0,0,0,0.142,0,0,0,0,0,
              0,0,0,0,0.566,0,0,0,0,
              0,0,0,0,0,0.026,0,0,0,
              0,0,0,0,0,0,0.103,0,0,
              0,0,0,0,0,0,0,0,0.000),nrow=9,byrow=TRUE,dimnames=list(stages,stages))
print(A)
MarkovOut <- pop.projection(A=A,n=n,iterations=timesteps)
plotCounts(MarkovOut$stage.vectors)
print(MarkovOut$lambda)              
print(MarkovOut$stable.stage)              
print(head(MarkovOut$pop.sizes))
print(tail(MarkovOut$pop.sizes))

eigen.analysis(A)
```

### Real Example #3 Ant community assembly with climate warming
```{r}
# Data from Diamond, S. E., L. M. Nichols, S. L. Pelini, C. A. Penick, G. W. Barber, S. H. Cahan, R. R. Dunn, A. M. Ellison, N. J. Sanders, and N. J. Gotelli. 2016. Climatic warming destabilizes forest ant communities. Science Advances 2.

# transition matrix from control 5C heated chamber
stages <- c("Empty",paste("Species",LETTERS[1:3]))
timesteps=70
n= rep(20,4)
A <- matrix(c(245,14,8,0,
              18,17,0,2,
              8,1,0,5,
              1,0,0,1) ,nrow=length(stages),byrow=TRUE,dimnames=list(stages,stages)) 
print(A)
A <- apply(A,2, function(x) x/sum(x)) # convert raw counts to column proportions
print(A)
MarkovOut <- pop.projection(A=A,n=n,iterations=timesteps)
plotCounts(MarkovOut$stage.vectors)
print(MarkovOut$stable.stage)              
eigen.analysis(A)

# transition matrix from control (=unheated) chamber
stages <- c("Empty",paste("Species",LETTERS[1:3]))
timesteps=70
n= rep(20,4)
A <- matrix(c(202,14,13,0,
              18,9,2,0,
              13,1,35,1,
              1,0,0,11) ,nrow=length(stages),byrow=TRUE,dimnames=list(stages,stages)) 
print(A)
A <- apply(A,2, function(x) x/sum(x)) # convert raw counts to column proportions
print(A)
MarkovOut <- pop.projection(A=A,n=n,iterations=timesteps)
plotCounts(MarkovOut$stage.vectors)
print(MarkovOut$stable.stage)              
eigen.analysis(A)



```

