---
title: 'Lecture #23 ggplot IV'
author: "Nicholas J. Gotelli"
date: "April 17, 2018"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---
### ggplot (continued)

1. Aesthetics
  a. Colors that are attractive
    - large geoms (fills) - pale colors
    - small geoms(lines,points) - bright colors
  b. Colors that highlight elements
    - pale, grey to dee-emphasize
    - bright or saturated colors to emphasize

2. Information content
  a. Discrete scale
   - colors to group similar treatments
   - neutral colors (black,grey,white) to indicate control groups
   - Symbolic colors (heat=red, cool = blue, photosynthesis/growth=green, oligotrophic=blue, eutrophic=brown, infected=red)
   - Colors that map to chemical stains or gels, or even colors of organisms
 b. Continuous scale
  - monochromatic (differing shades of 1 color)
  - 2 tone chromatic scale (from color x to color y)
  - 3 tone divergent scale (from color x through color y to color z)
 c. Use color information with and between graphs
    



#### Preliminaries
```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(ggthemes)
library(patchwork)
library(TeachingDemos)
library(wesanderson)
char2seed("Dark Star")
```
#### Working with color and grey

```{r,warning=FALSE,message=FALSE}
d <- mpg # use built in mpg data frame
p1 <- ggplot(data=d,mapping=aes(x=fl,y=hwy,group=fl))
p1 + geom_boxplot() # no color
p1 + geom_boxplot(fill="red") # single color (no info)

# create a palette
myColors <- c("red","green","pink","blue","orange") #no info
p1 + geom_boxplot(fill=myColors)

myColors <- c("red","green","green","blue","orange") # info
p1 + geom_boxplot(fill=myColors)


```

#### Color specifications in R
```{r}
# color specifications
myGrey <- gray(0.7)
p1 + geom_boxplot(fill=myGrey) # single color (no info)
print(myGrey)
# explain rgb color system in r
# convert to an rgb scale (intensities of red,green,blue)
. <- col2rgb("red") # converts named colors to rgb
print(.)

. <- ./255 # rescale to 0,1 scale
print(.)

. <- rgb(t(.)) # convert 0,1 scale back to hexadecimal number
print(.) # color representation of red


. <- col2rgb("red") # converts named colors to rgb
print(.)
. <- ./255 # rescale to 0,1 scale
. <- rgb(t(.),alpha=0.5) # add transparency alpha
print(.) # color representation of red

# add a fourth element, which is alpha or transparency (0=transparent, 1=opaque)
myPaleRed <- .
print(myPaleRed)

# can also specify color in "RRGGBB" hexadecimal 0(00) to 255(FF)
p1 + geom_boxplot(fill=myPaleRed)
p1 + geom_boxplot(fill=gray(seq(from=0.1,to=0.9,length=5)))


# using alpha transparency for histograms
x1 <- rnorm(n=100,mean=0)
x2 <- rnorm(n=100,mean=3)
dFrame <- data.frame(v1=c(x1,x2))
lab <- rep(c("Control","Treatment"),each=100)
dFrame <- cbind(dFrame,lab)
str(dFrame)
h1 <- ggplot(data=dFrame,mapping=aes(x=v1,fill=lab))
h1 + geom_histogram(position="identity",alpha=0.5,color="black") 
# now go out to web and show them the palette page
p1 + geom_boxplot(fill=wes_palettes[["Royal2"]])
p1 + geom_boxplot(fill=c(grey(0.5),canva_palettes[[1]]))

# use scale fill manual

p2 <- ggplot(data=d, mapping=aes(x=fl,y=hwy,fill=fl)) + geom_boxplot()+ scale_fill_manual(values=wes_palettes[["Darjeeling"]])
print(p2)


myCanvas <- canva_palettes[["Fresh and bright"]]
myCanvas <- c(gray(0.6),myCanvas)
p2 + geom_boxplot() + scale_fill_manual(values=myCanvas)
```

#### Colorbrewer page
```{r}
# visit colorbrewer page for demo
p2 + geom_boxplot() + scale_fill_brewer(palette="Blues")

# slightly different scheme for points, which have color

p3 <- ggplot(data=d, mapping=aes(x=displ,y=hwy,color=fl)) +
  geom_point()   +
   scale_color_brewer(palette="Spectral")
print(p3)
```

#### Continuous variables
```{r}
p3 <- ggplot(data=d, mapping=aes(x=displ,y=hwy,color=cty)) +
  geom_point()    
print(p3)

# use scale_color_gradient to change low and high colors
  # scale_color_gradient(
p3 +
  scale_color_gradient(low="red",high="blue")

# use scale_color_gradient2 for a 3 color gradient
z=mean(d$cty)
p3 +
  scale_color_gradient2(midpoint=z,low="red",mid="seagreen",high="cyan",space="Lab")


# use scale_color_gradientn for multicolored changes
p3+scale_color_gradientn(colours = rainbow(5))
p3+scale_color_gradientn(colours = myCanvas)
```

#### Making a heat map
```{r}
xVar <- 1:30
yVar <- 1:5
myData <- expand.grid(xVar=xVar,yVar=yVar)
head(myData)
zVar <- myData$xVar + myData$yVar + 2*rnorm(n=150)
myData <- cbind(myData,zVar)
head(myData)
p4 <- ggplot(data=myData,mapping=aes(x=xVar,y=yVar,fill=zVar))
p4 + geom_tile()

p4 <- ggplot(data=myData,mapping=aes(x=xVar,y=yVar,fill=zVar))
p4 + geom_tile() +
scale_fill_gradient2(midpoint=19,low="brown",mid=grey(0.8),high="darkblue")
```

