---
title: 'Probability Distributions'
author: "Nicholas J. Gotelli"
date: "February 18, 2020"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

### Probability distributions in R
#### Discrete distributions

- Poisson
    * Range: [0,$\infty$]
    * Parameters: size = number of trials, rate = $\lambda$
    * Interpretation: Distribution of events that occur during a fixed time interval or sampling effort with a constant rate of independent events
    
- Binomial
    * Range: [0, # of trials]
    * Parameters: size= number of trials; p = probability of positive outcome
    * Interpretation: Distribution of number of successful independent dichotomous trials, with constant p; resembles normal with large $\lambda$, or exponential with small $\lambda$
    
- Negative Binomial
    * Range: [0, $\infty$]
    * Parameters: size=number of successes; p = probability of success
    * Interpretation: Distribution of number of failures in a series of independent Bernouli trials, each with p = probability of a success. Generates a discrete distribution that is more heterogeneous ("overdispersed") than Poisson
    
#### Continuous distributions

- Uniform
    * Range: [min,max]
    * Parameters: min = minimum boundary; max = maximum boundary
    * Interpretation: Distribution of a value that is equally likely within a specified range
    
- Normal
    * Range: [$-\infty,\infty$]
    * Parameters: mean = central tendency; SD = standard deviation
    * Interpretation: Symmetric bell-shaped curve with unbounded tails
    
- Gamma $\Gamma$
    * Range: [0,$\infty$]
    * Parameters: shape, scale
    * Interpretation: mean=$shape*scale$, variance=$shape*scale^2$; generates a variety of shapes (including normal and exponential) for positive continuous variables
    
- Beta $\beta$
    * Range: [0,1] (can be rescaled to any range by simple multiplication and addition)
    * Paramters: shape1, shape2
    * Interpretation: if shape1 and shape 2 are integers, interpret as a coin toss, with shape1 = # of successes + 1, shape2 = # of failures + 1. Gives distribution of value of p, estimated from data, which can range from exponential through uniform through normal (but all are bounded). Setting shape1 and shape2 <1 yields u-shaped distributions.

### The "grammar" of probability distributions in R
- `d` gives probability density function
- `p` gives cumulative distribution function
- `q` gives quantile function (the inverse of `p`)
- `r` gives random number generation

Combine these with the base name of the function. For example `rbinom` gives a set of random values drawn from a binomial, whereas `dnorm` gives the density function for a normal distribution. There are many probability distributions available in R, but we will discuss only 7 of them.

#### Poisson distribution

```{r}
library(ggplot2)
library(MASS)
#-------------------------------------------------
# Poisson distribution
# Discrete X >= 0
# Random events with a constant rate lambda
# (observations per time or per unit area)
# Parameter lambda > 0

# "d" function generates probability density
hits <- 0:10
myVec <- dpois(x=hits,lambda=1)
qplot(x=hits,y=myVec,geom="col",color=I("black"),fill=I("goldenrod"))

myVec <- dpois(x=hits,lambda=2)
qplot(x=hits,y=myVec,geom="col",color=I("black"),fill=I("goldenrod"))

hits <- 0:15
myVec <- dpois(x=hits,lambda=6)
qplot(x=hits,y=myVec,geom="col",color=I("black"),fill=I("goldenrod"))


hits <- 0:15
myVec <- dpois(x=hits,lambda=0.2)
qplot(x=hits,y=myVec,geom="col",color=I("black"),fill=I("goldenrod"))

sum(myVec)  # sum of density function = 1.0 (total area under curve)

# for a Poisson distribution with lambda=2, 
# what is the probability that a single draw will yield X=0?

dpois(x=0,lambda=2)

# "p" function generates cumulative probability density; gives the 
# "lower tail" cumulative area of the distribution

hits <- 0:10
myVec <- ppois(q=hits,lambda=2)
qplot(x=hits,y=myVec,geom="col",color=I("black"),fill=I("goldenrod"))


# for a Poisson distribution with lambda=2, 
# what is the probability that a single random draw will yield X <= 1?

ppois(q=1, lambda=2)


# We could also get this through dpois
p1 <- dpois(x=0,lambda=2)
p1
p2 <- dpois(x=1,lambda=2)
p2
p1 + p2


# The q function is the inverse of p
# What is the number of hits corresponding to 50% of the probability mass
qpois(p=0.5,lambda=2.5)
qplot(x=0:10,y=dpois(x=0:10,lambda=2.5),geom="col",color=I("black"),fill=I("goldenrod"))

# but distribution is discrete, so this is not exact
ppois(q=2,lambda=2.5)

# finally, we can simulate individual values from a poisson
ranPois <- rpois(n=1000,lambda=2.5)
qplot(x=ranPois,color=I("black"),fill=I("goldenrod"))


# for real or simulated data, we can use the quantile
# function to find the empirical  95% confidence interval on the data

quantile(x=ranPois,probs=c(0.025,0.975))
```

#### Binomial distribution

```{r}
#-------------------------------------------------
# Binomial distribution
# p = probability of a dichotomous outcome
# size = number of trials
# x = possible outcomes
# outcome x is bounded between 0 and number of trials

# use "d" binom for density function
hits <- 0:10
myVec <- dbinom(x=hits,size=10,prob=0.5)
qplot(x=0:10,y=myVec,geom="col",color=I("black"),fill=I("goldenrod"))

# what is probability of getting 5 heads out of 10 tosses?
dbinom(x=5,size=10,prob=0.5)

# of course, the coin can be a biased one
myVec <- dbinom(x=hits,size=10,prob=0.95)
qplot(x=0:10,y=myVec,geom="col",color=I("black"),fill=I("goldenrod"))

# use "p" binom for cumulative distribution


# what is the probability of getting 5 
# or fewer heads out of 10 tosses?
pbinom(q=5,size=10,prob=0.5)

# symmetry requires odd number of tosses
pbinom(q=4,size=9,prob=0.5)



# what is a 95% confidence interval for 100 trials 
# of a coin with p = 0.7 for heads?
qbinom(p=c(0.025,0.975),size=100,prob=0.7)

# and how does this compare to the interval for an actual simlation of 50 tosses of 100 coins?

myCoins <- rbinom(n=50,size=100,prob=0.5)
qplot(x=myCoins,color=I("black"),fill=I("goldenrod"))
quantile(x=myCoins,probs=c(0.025,0.975))

```

#### Negative Binomial distribution

```{r}
#-------------------------------------------------
# negative binomial: number of failures (values of MyVec)
# in a series of (Bernouli) with p=probability of success 
# before a target number of successes (= size)
# generates a discrete distribution that is more 
# heterogeneous ("overdispersed") than Poisson
hits <- 0:40
myVec <- dnbinom(x=hits, size=5, prob=0.5)
qplot(x=hits,y=myVec,geom="col",color=I("black"),fill=I("goldenrod"))

# geometric series is a special case where N= 1 success
# each bar is a constant fraction 1 - "prob" of the bar before it
myVec <- dnbinom(x=hits, size=1, prob=0.1)
qplot(x=hits,y=myVec,geom="col",color=I("black"),fill=I("goldenrod"))


# alternatively specify mean = mu of distribution and size, 
# the dispersion parameter (small is more dispersed)
# this gives us a poisson with a lambda value that varies
# the dispersion parameter is the shape parameter in the gamma
# as it increases, the distribution has a smaller variance
# just simulate it directly

nbiRan <- rnbinom(n=1000,size=10,mu=5)
qplot(nbiRan,color=I("black"),fill=I("goldenrod"))

nbiRan <- rnbinom(n=1000,size=0.1,mu=5)
qplot(nbiRan,color=I("black"),fill=I("goldenrod"))
```
#### Uniform distribution

```{r}
#-------------------------------------------------
# uniform
# params specify minimum and maximum

#runif for random data
qplot(x=runif(n=100,min=0,max=5),color=I("black"),fill=I("goldenrod"))
qplot(runif(n=1000,min=0,max=5),color=I("black"),fill=I("goldenrod"))
#-------------------------------------------------
```
#### Normal distribution

```{r}
# normal 
myNorm <- rnorm(n=100,mean=100,sd=2)
qplot(myNorm,color=I("black"),fill=I("goldenrod"))

# problems with uniform when mean is small but zero is not allowed.
myNorm <- rnorm(n=100,mean=2,sd=2)
qplot(myNorm,color=I("black"),fill=I("goldenrod"))
summary(myNorm)
tossZeroes <- myNorm[myNorm>0]
qplot(tossZeroes,color=I("black"),fill=I("goldenrod"))
summary(tossZeroes)
```

#### Gamma distribution

``` {r}
#-------------------------------------------------
# gamma distribution, continuous positive values, but bounded at 0

myGamma <- rgamma(n=100,shape=1,scale=10)
qplot(myGamma,color=I("black"),fill=I("goldenrod"))

# gamma with shape= 1 is an exponential with scale = mean

# shape <=1 gives a mode near zero; very small shape rounds to zero
myGamma <- rgamma(n=100,shape=0.1,scale=1)
qplot(myGamma,color=I("black"),fill=I("goldenrod"))

# large shape parameters moves towards a normal
myGamma <- rgamma(n=100,shape=20,scale=1)
qplot(myGamma,color=I("black"),fill=I("goldenrod"))

# scale parameter changes mean- and the variance!
qplot(rgamma(n=100,shape=2,scale=100),color=I("black"),fill=I("goldenrod"))
qplot(rgamma(n=100,shape=2,scale=10),color=I("black"),fill=I("goldenrod"))
qplot(rgamma(n=100,shape=2,scale=1),color=I("black"),fill=I("goldenrod"))
qplot(rgamma(n=100,shape=2,scale=0.1),color=I("black"),fill=I("goldenrod"))


# unlike the normal, the two parameters affect both mean and variance

# mean = shape*scale
# variance= shape*scale^2
```

#### Beta distribution 

```{r}
#-------------------------------------------------

# beta distribution 
# bounded at 0 and 1
# analagous to a binomial, but result is a continuous distribution of probabilities
# parameter shape1 = number of successes + 1
# parameter shape2 = number of failures + 1
# interpret these in terms of a coin you are tossing

# shape1 = 1, shape2 = 1 = "no data"
myBeta <- rbeta(n=1000,shape1=1,shape2=1)
qplot(myBeta,xlim=c(0,1),color=I("black"),fill=I("goldenrod"))


# shape1 = 2, shape1 = 1 = "1 coin toss, comes up heads!"
myBeta <- rbeta(n=1000,shape1=2,shape2=1)
qplot(myBeta,xlim=c(0,1),color=I("black"),fill=I("goldenrod"))

# two tosses, 1 head and 1 tail
myBeta <- rbeta(n=1000,shape1=2,shape2=2)
qplot(myBeta,xlim=c(0,1),color=I("black"),fill=I("goldenrod"))

# two tosses, both heads
myBeta <- rbeta(n=1000,shape1=2,shape2=1)
qplot(myBeta,xlim=c(0,1),color=I("black"),fill=I("goldenrod"))

# let's get more data
myBeta <- rbeta(n=1000,shape1=20,shape2=20)
qplot(myBeta,xlim=c(0,1),color=I("black"),fill=I("goldenrod"))

myBeta <- rbeta(n=1000,shape1=500,shape2=500)
qplot(myBeta,xlim=c(0,1),color=I("black"),fill=I("goldenrod"))

# if the coin is biased
myBeta <- rbeta(n=1000,shape1=1000,shape2=500)
qplot(myBeta,xlim=c(0,1),color=I("black"),fill=I("goldenrod"))
myBeta <- rbeta(n=1000,shape1=10,shape2=5)
qplot(myBeta,xlim=c(0,1),color=I("black"),fill=I("goldenrod"))


# shape parameters less than 1.0 give us a u-shaped distribution
myBeta <- rbeta(n=1000,shape1=0.1,shape2=0.1)
qplot(myBeta,xlim=c(0,1),color=I("black"),fill=I("goldenrod"))
myBeta <- rbeta(n=1000,shape1=0.5,shape2=0.2)
qplot(myBeta,xlim=c(0,1),color=I("black"),fill=I("goldenrod"))
```

### Estimating paramaters from data

```{r}
#-------------------------------------------------
# estimating parameters from data
# maximum likelihood estimator theta versus P(data|theta)

# use fitdistr function, feeding it data and a distribution type)
x <- rnorm(1000,mean=92.5,sd=2.5)
qplot(x,color=I("black"),fill=I("goldenrod"))
fitdistr(x,"normal")

# compare to true parameters
# compare to parameters estimated from simple means and standard deviations
mean(x)
sd(x)

# but how do we "know" what distribution to fit?
fitdistr(x,"gamma")
z <- fitdistr(x,"gamma")

# find components of z
str(z)
# rate = 1/scale
# so here is the estimate of the mean
z$estimate[1]/z$estimate[2]

# and here is the estimate of the variance
z$estimate[1]/z$estimate[2]^2
```




### Archetype Experimental Designs
- independent versus dependent variables
- discrete versus continuous variables
- continuous variables (integer and real)
- direction of cause and effect, x axis is independent
- continuous versus discrete (natural or arbitrary or statistical bins)

### Regression (dependent: continuous, independent: continuous)
- linear model of $y = a + bx$
- statistical tests for null of hypothesis of slope and/or intercept = 0
- confidence and prediction intervals of uncertainty
- goodness of fit tests for linearity

### Set-up
```{r}
library(tidyverse)

```


### Data Frame construction for Regression Data

```{r}
n = 50  # number of observations (rows)


varA <- runif(n) # random uniform values (independent)
varB <- runif(n) # a second random column (dependent)
varC <- 5.5 + varA*10 # a noisy linear relationship with varA
ID <- seq_len(n) # creates a sequence from 1:n (if n > 0!)
regData <- data.frame(ID,varA,varB,varC)
head(regData)
str(regData)
```

# Basic regression analysis in R
```{r}
# model
regModel <- lm(varB~varA,data=regData)

# model output
regModel # printed output is sparse
str(regModel) # complicated, but has "coefficients"
head(regModel$residuals) # contains residuals

# 'summary' of model has elements
summary(regModel) # 
summary(regModel)$coefficients
str(summary(regModel))

# best to examine entire matrix of coefficients:
summary(regModel)$coefficients[] #shows all

# can pull results from this, but a little wordy
summary(regModel)$coefficients[1,4]   #p value for intercept
summary(regModel)$coefficients["(Intercept)","Pr(>|t|)"] # uggh


# alternatively unfurl this into a 1D atomic vector with names
z <- unlist(summary(regModel))
str(z)
z
z$coefficients7

# grab what we need and put into a tidy  list

regSum <- list(intercept=z$coefficients1,
               slope=z$coefficients2,
               interceptP=z$coefficients7,
               slopeP=z$coefficients8,
               r2=z$r.squared)

# much easier to query and use
print(regSum)
regSum$r2
regSum[[5]]

```

### Basic ggplot of regression model

```{r}
regPlot <- ggplot(data=regData,aes(x=varA,y=varB)) +
           geom_point() +
           stat_smooth(method=lm,se=0.99) # default se=0.95 
print(regPlot)
# ggsave(filename="Plot1.pdf",plot=regPlot,device="pdf")
```

### Data frame construction for one-way ANOVA

```{r}
nGroup <- 3 # number of treatment groups
nName <- c("Control","Treat1", "Treat2") # names of groups
nSize <- c(12,17,9) # number of observations in each group
nMean <- c(40,41,60) # mean of each group
nSD <- c(5,5,5) # standardd deviation of each group

ID <- 1:(sum(nSize)) # id vector for each row
resVar <- c(rnorm(n=nSize[1],mean=nMean[1],sd=nSD[1]),
            rnorm(n=nSize[2],mean=nMean[2],sd=nSD[2]),
            rnorm(n=nSize[3],mean=nMean[3],sd=nSD[3]))
TGroup <- rep(nName,nSize)
ANOdata <- data.frame(ID,TGroup,resVar)
str(ANOdata)
```

### Basic ANOVA in R

```{r}
ANOmodel <- aov(resVar~TGroup,data=ANOdata)
print(ANOmodel)
print(summary(ANOmodel))
z <- summary(ANOmodel)
str(z)
aggregate(resVar~TGroup,data=ANOdata,FUN=mean)
unlist(z)
unlist(z)[7]
ANOsum <- list(Fval=unlist(z)[7],probF=unlist(z)[9])
ANOsum


```


### Basic ggplot of ANOVA data

```{r}
ANOPlot <- ggplot(data=ANOdata,aes(x=TGroup,y=resVar,fill=TGroup)) +
           geom_boxplot()
print(ANOPlot)
# ggsave(filename="Plot2.pdf",plot=ANOPlot,device="pdf")
```


### Data frame construction for logistic regression
```{r}

xVar <- sort(rgamma(n=200,shape=5,scale=5))
yVar <- sample(rep(c(1,0),each=100),prob=seq_len(200))
lRegData <- data.frame(xVar,yVar)

```
### Logistic regression analysis in R

```{r}
lRegModel <- glm(yVar ~ xVar,
                 data=lRegData,
                 family=binomial(link=logit))
summary(lRegModel)
summary(lRegModel)$coefficients

```

### Basic ggplot of logistic regression
```{r}
lRegPlot <- ggplot(data=lRegData, aes(x=xVar,y=yVar)) +
            geom_point() +
            stat_smooth(method=glm, method.args=list(family=binomial))
print(lRegPlot)
   
```

### Data for contingency table analysis
```{r}
# integer counts of different data groups
vec1 <- c(50,66,22)
vec2 <- c(120,22,30)
dataMatrix <- rbind(vec1,vec2)
rownames(dataMatrix) <- c("Cold","Warm")
colnames(dataMatrix) <-c("Aphaenogaster",
                         "Camponotus",
                         "Crematogaster")
str(dataMatrix)
```

### Basic contingency table analysis in R

```{r}
print(chisq.test(dataMatrix))
```

### Plotting contingency table analyses

```{r}
# some simple plots using baseR
mosaicplot(x=dataMatrix,
           col=c("goldenrod","grey","black"),
           shade=FALSE)
barplot(height=dataMatrix,
        beside=TRUE,
        col=c("cornflowerblue","tomato"))


dFrame <- as.data.frame(dataMatrix)
dFrame <- cbind(dFrame,list(Treatment=c("Cold","Warm")))
dFrame <- gather(dFrame,key=Species,Aphaenogaster:Crematogaster,value=Counts) 

p <- ggplot(data=dFrame,aes(x=Species,y=Counts,fill=Treatment)) + geom_bar(stat="identity",position="dodge",color=I("black")) +
  scale_fill_manual(values=c("cornflowerblue","coral"))
print(p)
```
