---
title: 'Lecture #11: Complete Cases, Tribbles, Long/Wide Data, and Archetypal Stats'
author: "Nicholas J. Gotelli"
date: "October 2, 2018"
output:
  html_document:
    highlight: tango
    keep_md: yes
    theme: united
  pdf_document: default
---
### Eliminating missing values with `complete.cases()`

```{r, eval=TRUE}

# preliminaries
library(tidyverse)
set.seed(99)


z <- 1:10 # simple integer sequence
z <- sample(z) # reshuffle the ordering
print(z)

z < 4 # create logical vector
z[z < 4] # use as index call
which(z < 4) # use to get indices for logical
z[which(z < 4)] # does same as above

zD <- c(z,NA,NA) # contaminate it
zD[zD < 4] # NA values carried along!
zD[which(zD < 4)] # NA values dropped

# use complete.cases with atomic vector
print(zD)

complete.cases(zD)

zD[complete.cases(zD)] # clean them out

which(!complete.cases(zD)) # find NA slots

# use with a matrix

m <- matrix(1:20,nrow=5)
m[1,1] <- NA
m[5,4] <- NA
print(m)

m[complete.cases(m),] 

# now get complete cases for only certain columns!
m[complete.cases(m[,c(1,2)]),] # drops row 1
m[complete.cases(m[,c(2,3)]),] # no drops
m[complete.cases(m[,c(3,4)]),] # drops row 4
m[complete.cases(m[,c(1,4)]),] # drops 1&4

```



### Converting between long and wide forms of data
```{r}
# Introducing the "tribble", a transposed tibble
# that is easy to type directly in to your own code

repMeasDat <- tribble(
  ~Subject, ~Treatment, ~Time1, ~Time2, ~Time3,
  #-----------------------------
  1, "Control", 1,1,2,
  2, "Control", 1,4,1,
  3, "Control", 2,2,2,
  4, "Heated", 5,6,6,
  5, "Heated", 7,7,5,
  6, "Heated", 7,5,5
  #------------------------------
)

print(repMeasDat)
```
This is a very common data format, in which each row is a subject. There are two treatments, and the three columns give the measurements on each subject at 3 different times. 

Unfortunately, this is not in the "long" format, which requires that each row represents a different observation. In this case, the observations should be the individual measurements, not the subjects.

Here is how we convert this to a correct format:



```{r}
repMeasDat %>%
gather(Time1:Time3, key="Time",value="Response")  
```

We first list the set of variables that need to be gathered (Time1:Time3). Next, we provide a `key`, which is the name of the new variable (in this case "Time"). Finally, we add a value, which is the measure that is associated with each row (in this case "Response"). Notice that we did not reference "Subject" or "Treatment". Their values are properly repeated in each of the appropriate rows. Thus, each row corresponds to a unique observation in the experiment. In this case, there were 3 subjects in each of 2 treatments, and each were measured 3 times. Thus, there are 3 x 3 x 2 rows in the new data frame.

Let's add one other refinement, using dplyr's `arrange` function to re-order the rows so that the 3 measurements from a single subject appear on successive lines in the data frame:

```{r}
longData <- repMeasDat %>%
gather(Time1:Time3, key="Time",value="Response") %>%
arrange(Subject)
print(longData)
```
The reason the long format is important, is that most statistical models and summaries in R operate this way. For example, to calculate averages for different times and treatments, we would use

```{r}
longData %>%
group_by(Time,Treatment) %>%
summarize(AvgRes = mean(Response)) %>%
arrange(Treatment,Time)
```


Now let's convert this back to wide format, using the `spread` function which is roughly the inverse of `gather`

```{r}
wideData <- longData %>%
spread(key=Time,value=Response)
print(wideData)
```

### Archetype Data Designs
- independent versus dependent variables
- discrete versus continuous variables
- continuous variables (integer and real)
- direction of cause and effect, x axis is independent
- continuous versus discrete (natural or arbitrary or statistical bins)

### Regression (dependent: continuous, independent: continuous)
- linear model of $y = a + bx$
- statistical tests for null of hypothesis of slope and/or intercept = 0
- confidence and prediction intervals of uncertainty
- goodness of fit tests for linearity

### Set-up
```{r}
library(ggplot2)
library(TeachingDemos)
char2seed('espresso')
```


### Data Frame construction for Regression Data

```{r}
n = 50  # number of observations (rows)


varA <- runif(n) # random uniform values (independent)
varB <- runif(n) # a second random column (dependent)
varC <- 5.5 + varA*10 # a noisy linear relationship with varA
ID <- seq_len(n) # creates a sequence from 1:n (if n > 0!)
regData <- data.frame(ID,varA,varB,varC)
head(regData)
str(regData)
```

# Basic regression analysis in R
```{r}
# model
regModel <- lm(varB~varA,data=regData)

# model output
regModel # printed output is sparse
str(regModel) # complicated, but has "coefficients"
head(regModel$residuals) # contains residuals

# 'summary' of model has elements
summary(regModel) # 
summary(regModel)$coefficients
str(summary(regModel))

# best to examine entire matrix of coefficients:
summary(regModel)$coefficients[] #shows all

# can pull results from this, but a little wordy
summary(regModel)$coefficients[1,4]   #p value for intercept
summary(regModel)$coefficients["(Intercept)","Pr(>|t|)"] # uggh


# alternatively unfurl this into a 1D atomic vector with names
z <- unlist(summary(regModel))
str(z)
z
z$coefficients7

# grab what we need and put into a tidy  list

regSum <- list(intercept=z$coefficients1,
               slope=z$coefficients2,
               interceptP=z$coefficients7,
               slopeP=z$coefficients8,
               r2=z$r.squared)

# much easier to query and use
print(regSum)
regSum$r2
regSum[[5]]

```

### Basic ggplot of regression model

```{r}
regPlot <- ggplot(data=regData,aes(x=varA,y=varB)) +
           geom_point() +
           stat_smooth(method=lm,se=0.99) # default se=0.95 
print(regPlot)
# ggsave(filename="Plot1.pdf",plot=regPlot,device="pdf")
```

### Data frame construction for one-way ANOVA

```{r}
nGroup <- 3 # number of treatment groups
nName <- c("Control","Treat1", "Treat2") # names of groups
nSize <- c(12,17,9) # number of observations in each group
nMean <- c(40,41,60) # mean of each group
nSD <- c(5,5,5) # standardd deviation of each group

ID <- 1:(sum(nSize)) # id vector for each row
resVar <- c(rnorm(n=nSize[1],mean=nMean[1],sd=nSD[1]),
            rnorm(n=nSize[2],mean=nMean[2],sd=nSD[2]),
            rnorm(n=nSize[3],mean=nMean[3],sd=nSD[3]))
TGroup <- rep(nName,nSize)
ANOdata <- data.frame(ID,TGroup,resVar)
str(ANOdata)
```

### Basic ANOVA in R

```{r}
ANOmodel <- aov(resVar~TGroup,data=ANOdata)
print(ANOmodel)
print(summary(ANOmodel))
z <- summary(ANOmodel)
str(z)
aggregate(resVar~TGroup,data=ANOdata,FUN=mean)
unlist(z)
unlist(z)[7]
ANOsum <- list(Fval=unlist(z)[7],probF=unlist(z)[9])
ANOsum


```


### Basic ggplot of ANOVA data

```{r}
ANOPlot <- ggplot(data=ANOdata,aes(x=TGroup,y=resVar,fill=TGroup)) +
           geom_boxplot()
print(ANOPlot)
# ggsave(filename="Plot2.pdf",plot=ANOPlot,device="pdf")
```


### Data frame construction for logistic regression
```{r}

xVar <- sort(rgamma(n=200,shape=5,scale=5))
yVar <- sample(rep(c(1,0),each=100),prob=seq_len(200))
lRegData <- data.frame(xVar,yVar)

```
### Logistic regression analysis in R

```{r}
lRegModel <- glm(yVar ~ xVar,
                 data=lRegData,
                 family=binomial(link=logit))
summary(lRegModel)
summary(lRegModel)$coefficients

```

### Basic ggplot of logistic regression
```{r}
lRegPlot <- ggplot(data=lRegData, aes(x=xVar,y=yVar)) +
            geom_point() +
            stat_smooth(method=glm, method.args=list(family=binomial))
print(lRegPlot)
   
```

### Data for contingency table analysis
```{r}
# integer counts of different data groups
vec1 <- c(50,66,22)
vec2 <- c(120,22,30)
dataMatrix <- rbind(vec1,vec2)
rownames(dataMatrix) <- c("Cold","Warm")
colnames(dataMatrix) <-c("Aphaenogaster",
                         "Camponotus",
                         "Crematogaster")
str(dataMatrix)
```

### Basic contingency table analysis in R

```{r}
print(chisq.test(dataMatrix))
```

### Plotting contingency table analyses

```{r}
# some simple plots using baseR
mosaicplot(x=dataMatrix,
           col=c("goldenrod","grey","black"),
           shade=FALSE)
barplot(height=dataMatrix,
        beside=TRUE,
        col=c("cornflowerblue","tomato"))
```

