---
title: 'Lecture #13: ANOVA, Regression Designs'
author: "Nicholas J. Gotelli"
date: "February 22, 2018"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval=FALSE)
```



### Archetype Data Designs
- independent versus dependent variables
- discrete versus continuous variables
- continuous variables (integer and real)
- direction of cause and effect, x axis is independent
- continuous versus discrete (natural or arbitrary or statistical bins)

### Regression (dependent: continuous, independent: continuous)
- linear model of $y = a + bx$
- statistical tests for null of hypothesis of slope and/or intercept = 0
- confidence and prediction intervals of uncertainty
- goodness of fit tests for linearity

### Set-up
```{r}
library(ggplot2)
library(TeachingDemos)
char2seed('espresso')
```


### Data Frame construction for Regression Data

```{r}
n = 50  # number of observations (rows)


varA <- runif(n) # random uniform values (independent)
varB <- runif(n) # a second random column (dependent)
varC <- 5.5 + varA*10 # a noisy linear relationship with varA
ID <- seq_len(n) # creates a sequence from 1:n (if n > 0!)
regData <- data.frame(ID,varA,varB,varC)
head(regData)
str(regData)
```

# Basic regression analysis in R
```{r}
# model
regModel <- lm(varB~varA,data=regData)

# model output
regModel # printed output is sparse
str(regModel) # complicated, but has "coefficients"
head(regModel$residuals) # contains residuals

# 'summary' of model has elements
summary(regModel) # 
summary(regModel)$coefficients
str(summary(regModel))

# best to examine entire matrix of coefficients:
summary(regModel)$coefficients[] #shows all

# can pull results from this, but a little wordy
summary(regModel)$coefficients[1,4]   #p value for intercept
summary(regModel)$coefficients["(Intercept)","Pr(>|t|)"] # uggh


# alternatively unfurl this into a 1D atomic vector with names
z <- unlist(summary(regModel))
str(z)
z
z$coefficients7

# grab what we need and put into a tidy  list

regSum <- list(intercept=z$coefficients1,
               slope=z$coefficients2,
               interceptP=z$coefficients7,
               slopeP=z$coefficients8,
               r2=z$r.squared)

# much easier to query and use
print(regSum)
regSum$r2
regSum[[5]]

```

### Basic ggplot of regression model

```{r}
regPlot <- ggplot(data=regData,aes(x=varA,y=varB)) +
           geom_point() +
           stat_smooth(method=lm,se=0.99) # default se=0.95 
print(regPlot)
# ggsave(filename="Plot1.pdf",plot=regPlot,device="pdf")
```

### Data frame construction for one-way ANOVA

```{r}
nGroup <- 3 # number of treatment groups
nName <- c("Control","Treat1", "Treat2") # names of groups
nSize <- c(12,17,9) # number of observations in each group
nMean <- c(40,41,60) # mean of each group
nSD <- c(5,5,5) # standardd deviation of each group

ID <- 1:(sum(nSize)) # id vector for each row
resVar <- c(rnorm(n=nSize[1],mean=nMean[1],sd=nSD[1]),
            rnorm(n=nSize[2],mean=nMean[2],sd=nSD[2]),
            rnorm(n=nSize[3],mean=nMean[3],sd=nSD[3]))
TGroup <- rep(nName,nSize)
ANOdata <- data.frame(ID,TGroup,resVar)
str(ANOdata)
```

### Basic ANOVA in R

```{r}
ANOmodel <- aov(resVar~TGroup,data=ANOdata)
print(ANOmodel)
print(summary(ANOmodel))
z <- summary(ANOmodel)
str(z)
aggregate(resVar~TGroup,data=ANOdata,FUN=mean)
unlist(z)
unlist(z)[7]
ANOsum <- list(Fval=unlist(z)[7],probF=unlist(z)[9])
ANOsum


```


### Basic ggplot of ANOVA data

```{r}
ANOPlot <- ggplot(data=ANOdata,aes(x=TGroup,y=resVar,fill=TGroup)) +
           geom_boxplot()
print(ANOPlot)
# ggsave(filename="Plot2.pdf",plot=ANOPlot,device="pdf")
```


### Data frame construction for logistic regression
```{r}

xVar <- sort(rgamma(n=200,shape=5,scale=5))
yVar <- sample(rep(c(1,0),each=100),prob=seq_len(200))
lRegData <- data.frame(xVar,yVar)

```
### Logistic regression analysis in R

```{r}
lRegModel <- glm(yVar ~ xVar,
                 data=lRegData,
                 family=binomial(link=logit))
summary(lRegModel)
summary(lRegModel)$coefficients

```

### Basic ggplot of logistic regression
```{r}
lRegPlot <- ggplot(data=lRegData, aes(x=xVar,y=yVar)) +
            geom_point() +
            stat_smooth(method=glm, method.args=list(family=binomial))
print(lRegPlot)
   
```

### Data for contingency table analysis
```{r}
# integer counts of different data groups
vec1 <- c(50,66,22)
vec2 <- c(120,22,30)
dataMatrix <- rbind(vec1,vec2)
rownames(dataMatrix) <- c("Cold","Warm")
colnames(dataMatrix) <-c("Aphaenogaster",
                         "Camponotus",
                         "Crematogaster")
str(dataMatrix)
```

### Basic contingency table analysis in R

```{r}
print(chisq.test(dataMatrix))
```

### Plotting contingency table analyses

```{r}
# some simple plots using baseR
mosaicplot(x=dataMatrix,
           col=c("goldenrod","grey","black"),
           shade=FALSE)
barplot(height=dataMatrix,
        beside=TRUE,
        col=c("cornflowerblue","tomato"))
```

