---
title: 'Lecture #21: ggplot I'
author: "Nicholas J. Gotelli"
date: "13 April 2023"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---

#### Preliminaries
```{r, eval=FALSE}
library(ggplot2)
library(ggthemes)
library(patchwork)

d <- mpg # use built in mpg data frame
str(d)
hist(mpg$hwy)

```

#### Use base R for simple, fast plotting while coding

```{r,warning=FALSE,message=FALSE, eval=FALSE}
# basic histogram (1 variable, continuous)
hist(d$hwy)

# basic density plot (1 variable, continuous)
plot(density(d$hwy))

# counts of a discrete variable (1 (or two) variable discrete)
table(d$fl)
table(d$fl,d$drv)

# basic bar plt (1 variable, discrete, counts)
barplot(table(d$fl))

# alternative using plot function
plot(table(d$fl),type="h",lwd=10)

# basic bar plot (1 variable, continuous)
mass <- c(con=10.5, trt=12.5, trt2=2)
barplot(mass)

# basic boxplot (1 continuous response variable, 1 discrete predictor variable)
 boxplot(d$cty~d$fl)

# basic scatter plot (2 continuous variables)
 cause <- runif(20)
 effect <- runif(20)
 plot(effect~cause)

# basic bubble plot (3 continuous variables)
 covariate <- 1:20
 cause <- sort(cause)
 plot(effect~cause,cex=0.5*covariate)
print(data.frame(cause,effect,covariate))

# basic line graph
plot(runif(20)~runif(20),type="l")

# basic line graph with points
plot(runif(20)~runif(20),type="b")



# basic curves and functions (use small line segments)
x_var <- seq(1,3,by=0.5)
my_fun <- function(x) sin(x) + 0.1*x
y_var <- my_fun(x_var)

plot(y_var~x_var,type="l")

# now use many tiny segments
x_var <- seq(1,100,by=0.1)
y_var <- my_fun(x_var)
plot(y_var~x_var,type="l")

# built in stats density functions
my_vec <- seq(1,100,by=0.1)
plot(dgamma(my_vec,shape=5,scale=3),type="l")

```
### Summary of quick plots in base r

#### 1 variable
- `hist(x)` # histogram of 1 continuous variable
- `plot(density(x))` # smoothed density plot of 1 continuous variable
- `barplot(x)` # simple bar plot with 1 continuous variable of bar heights
- `barplot(table(x))` # simple bar plot of frequency counts in x 
- `plot(table(x),type="h",lwd=10)` # histogram "bars" from plot function
- `plot(d<stat(param1=a,param2=b), type="l")` # use for stats prob density

#### 2 variables
- `boxplot(y~x)` # continuous y variable, categorical x variable
- `plot(y~x)` # simple scatter plot
- `plot(y~x,type="l")` # use for evaluating functions
- `plot(y~x,type="b")` # use for time series data

#### 3 variables
- `plot(y~x,cex=0.5*z)` # bubble plot of 3 continuous variables

### Introduction to ggplot

#### Components of a ggplot layer
1. data (in a data frame)
2. aesthetic mappings (variables are mapped to aesthetics)
3. geom (the geometric used to draw the layer; has specific aesthetics)
4. stat (takes raw data and transforms it for useful plotting)
5. position (adjustment to prevent overplotting)

---

```
p1 <- ggplot(data= <DATA>) +
      aes(<MAPPINGS>) + 
      <GEOM_FUNCTION>(aes(<MAPPINGS>),
                stat=<STAT>,
                position=<POSITION>) +
      <COORDINATE_FUNCTION> +
      <FACET_FUNCTION>

print(p1)
p1 + geom_annotate # will print contents of p1 with geom_annotate() added


ggsave(plot=p1,
       filename="MyPlot.pdf",
       width=5,height=3,units="in",
       dpi=300)
```

#### Template for ggplot components
### Themes and fonts

```{r,warning=FALSE,message=FALSE, eval=FALSE}
library(ggthemes)
d <- mpg
 p1 <- ggplot(data=d, mapping=aes(x=displ,y=cty)) + geom_point()
 print(p1)
 
 p1 + theme_bw() # good with grid lines
 p1 + theme_classic() # no grid lines
 p1 + theme_linedraw() # black frame
 p1 + theme_dark() # good for brightly colored points
 p1 + theme_base() # mimics base R
 p1 + theme_par() # matches current par settings in base
 p1 + theme_void() # shows data only
 p1 + theme_solarized() # good for web pages
 p1 + theme_economist() # many specialized themes
 p1 + theme_grey() # ggplots default theme
``` 

### Major theme modifications 

```{r, warning=FALSE,message=FALSE, eval=FALSE}
d <- mpg
 p1 <- ggplot(data=d, mapping=aes(x=displ,y=cty)) + geom_point()
 print(p1)
 # use theme parameters to modify font and font size
 p1 + theme_classic(base_size=20,base_family="serif")
 # defaults: theme_grey, base_size=16,base_family="Helvetica")
 # font families (Mac): Times, Ariel, Monaco, Courier, Helvetica, serif,sans
 
# use coordinate_flip to invert entire plot
 p2 <- ggplot(data=d) +
       aes(x=fl,fill=fl) + 
       geom_bar()
 print(p2)
 p2 + coord_flip() + theme_grey(base_size=25,base_family="sans")
 
 # avoid this common mistake!
  p2 <- ggplot(data=d) +
       aes(x=fl,fill="goldenrod") + 
       geom_bar()
 print(p2)
 
 # instead, apply fill and color to the geom, not to the original aes
  p2 <- ggplot(data=d) +
       aes(x=fl) + 
       geom_bar(fill="goldenrod",color="black")
 print(p2)
 
```

### Minor theme modifications
```{r, warning=FALSE, message=FALSE, eval=FALSE}
d <- mpg
 p1 <- ggplot(data=d) +
       aes(x=displ,y=cty) + 
       geom_point(size=7,
                  shape=21,
                  color="black",
                  fill="steelblue") +
       labs(title="My graph title here",
            subtitle="An extended subtitle that will print below the main title",
            x="My x axis label",
            y="My y axis label") +
            xlim(0,4) + ylim(0,20)
 print(p1)
 
```

