---
title: 'Lecture #21: Markov Models'
author: "Nicholas J. Gotelli"
date: "April 13, 2017"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---
### Pseudocode outline for randomization tests
1. Define univariate metric X
2. Calculate X for observed data X~obs~
3. Randomize observed data, subject to constraints
4. Calculate X for simulated data X~sim~
5. Repeat (4) many times to generate distribution of X~sim~, which is H~0~
6. Estimate P(X~obs~|H~0~)
7. Draw biological inference

### Compare Species Richness Among New England Ecotypes
```{r, eval=FALSE}
data <- read.table(file="antcountydata.csv",
                   header=TRUE,
                   sep=",",
                   stringsAsFactors=FALSE)

data$ecoregion <- as.factor(data$ecoregion) # create model factor
boxplot(data$n.species~data$ecoregion,col="indianred") # plot data
myModel <- aov(n.species~ecoregion,data=data)
summary(myModel)

```
### Define response metric
Use the variance among treatment means as the factor

```{r, eval=FALSE}

##################################################
# function: VarMeans
# Calculates variance among treatment means
# input: vector of groups, vector of values
# output: among group variance of means
#------------------------------------------------- 
VarMeans <- function(fac=NULL,res=runif(25)) {
if(is.null(fac)){
  fac <- as.factor(rep(LETTERS[1:5],each=5))
}
temp <- aggregate(x=res,by=list(fac),FUN=mean)
return(var(temp$x))
# return(var(temp[,2]))
}
```

### Specify treatment randomization


```{r, eval=FALSE}
##################################################
# function: TreatRan
# Randomize assignment of replicates to treatments
# input: vector of treatment assignments, vector of response variable
# output: vector of randomized treatment assignments
#------------------------------------------------- 
TreatRan <- function(fac=NULL) {
if(is.null(fac)){
  fac <- as.factor(rep(LETTERS[1:5],each=5))
}
fac <- sample(fac, replace=FALSE) # reshuffle factor labels 


return(fac)
}
```
### Summarize output
```{r, eval=FALSE}
##################################################
# function: randomizeSummary
# give summary stats and histogram plot from randomization test
# input: vector of simulated values, observed value
# output: list of null model statistics, histogram, box plots
#------------------------------------------------- 
randomizeSummary <- function(obsX=runif(1),
                             simX=runif(1000), 
                             fac=NULL,
                             res=runif(25)) {
  
if(is.null(fac)){
  fac <- as.factor(rep(LETTERS[1:5],each=5))
} 
  pLow <- mean(simX <= obsX)
  pHigh <- mean(simX >= obsX)
  meanSimX <- mean(simX)
  varSimX <- var(simX)
  SES <- (obsX - mean(simX))/sd(simX)
  
  
  
  outList <- list(stat=list(obsX=obsX,
                  meanSimX=meanSimX,
                  varSimX=varSimX,
                  pLow=pLow,
                  pHigh=pHigh,
                  SES=SES,
                  reps=length(simX)),
                  raw=simX)
par(mfrow=c(1,3))  

# Show boxplot of real data
boxplot(res~fac,col="indianred")
mtext("Observed Data",side=3,font=2,cex=0.75)

# Show boxplot of simulated data
boxplot(res~sample(fac),col="goldenrod")
mtext("Simulated Data",side=3,font=2,cex=0.75)



# Show histogram of simulated and observed values
hist(simX, breaks=25, 
           col="goldenrod",
           main="Null Distribution",
           xlab="Simulated Response",
           xlim=c(0,max(c(simX,obsX))))  # make sure histogram includes observed

Interval95 <- quantile(x=simX, probs=c(0.05,0.95))
Interval975 <- quantile(x=simX, probs=c(0.025,0.975))
abline(v = obsX,col="indianred",lwd=2)
abline(v = Interval95,col="black",lwd=2,lty="dotted")
abline(v = Interval975,col="black",lwd=2,lty="dashed")
par(mfrow=c(1,1))
return(outList)
}
# z <- randomizeSummary()
```

### Create program body to use these functions

```{r, eval=FALSE}
# global variables
filename <- "antcountydata.csv"
data <- read.table(file=filename,
                   header=TRUE,
                   sep=",",
                   stringsAsFactors=FALSE)

data$ecoregion <- as.factor(data$ecoregion) # create model factor
grps <- data$ecoregion # grouping factor (ecoregions)
richness <- data$n.species # response variable
nreps <- 1000 # number of repetitions for randomization
outVec <- vector(mode="numeric",length=nreps) # storage variable for simulated
obsX <- VarMeans(fac=grps,res=richness) # observed index (variance among means)


# for (i in 1:nreps) {
#  outVec[i] <- VarMeans(fac=TreatRan(grps),res=richness)
# }

# runif(1)
# runif(5)
# replicate(5,runif(1))
# replicate(3,runif(5))

outVec <- replicate(nreps,VarMeans(fac=TreatRan(grps),res=richness))

z <- randomizeSummary(obsX=obsX,simX=outVec,fac=grps,res=richness)
z$stat #print statistics
```


