---
title: 'Lecture #25 ggplot III'
author: "Nicholas J. Gotelli"
date: "April 16th, 2020"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      warning=FALSE)
```

### ggplot (continued)

#### Preliminaries
```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(ggthemes)
library(patchwork)
library(colorblindr)
library(cowplot)
library(colorspace)
library(wesanderson)
library(ggsci)
library(TeachingDemos)

char2seed("Dark Star")
d <- mpg
```

#### Bar plots
```{r,message=FALSE, warning=FALSE}
# use to plot the counts of rows for a categorical variable
table(d$drv)
p1 <- ggplot(d,aes(x=drv)) + geom_bar(color="black",fill="cornsilk")
print(p1)

# aesthetic mapping gives multiple groups for each bar
p1 <- ggplot(d,aes(x=drv,fill=fl)) + geom_bar()
print(p1)


# stacked, but need to adjust color transparency, which is "alpha"
p1 <- ggplot(d,aes(x=drv,fill=fl)) + geom_bar(alpha = 0.3, position="identity")
print(p1)

# better to use position = fill for stacking, but with equivalent height
p1 <- ggplot(d,aes(x=drv,fill=fl)) + geom_bar(position="fill")
print(p1)

# best to use position = dodge for multiple bars
p1 <- ggplot(d,aes(x=drv,fill=fl)) + geom_bar(position="dodge",color="black",size=1)
print(p1)

# more typical "bar plot" has heights as the values themselves
dTiny <- tapply(X=d$hwy,INDEX=as.factor(d$fl),FUN=mean) #calculate the means
dTiny <- data.frame(hwy=dTiny) # create a single-column data frame
dTiny <- cbind(fl=row.names(dTiny),dTiny) # 

p2 <- ggplot(dTiny, aes(x=fl,y=hwy,fill=fl)) +
  geom_col()
print(p2)



#### Use a box plot instead of standard "means" bars!


```{r, warning=FALSE, message=FALSE}
# basic boxplot is simple and informative
p1 <- ggplot(d,aes(x=fl,y=hwy,fill=fl)) +
               geom_boxplot()
print(p1)

# now overlay the raw data
p1 <- ggplot(d,aes(x=fl,y=hwy)) +
               geom_boxplot(fill="thistle",outlier.shape=NA) + 
  # geom_point()
  geom_point(position=position_jitter(width=0.1,height=0.7),color="grey60",size=2)

print(p1)
```
### Using Colors

1. Aesthetics
  a. Colors that are attractive
    - large geoms (fills) - pale colors
    - small geoms(lines,points) - bright colors
  b. Colors that highlight elements
    - pale, grey to de-emphasize
    - bright or saturated colors to emphasize
  c. Colors that are visible to the color blind
  d. Colors that convert well to black and white

2. Information content
  a. Discrete scale
   - colors to group similar treatments
   - neutral colors (black,grey,white) to indicate control groups
   - Symbolic colors (heat=red, cool = blue, photosynthesis/growth=green, oligotrophic=blue, eutrophic=brown, infected=red)
   - Colors that map to chemical stains or gels, or even colors of organisms
 b. Continuous scale
  - monochromatic (differing shades of 1 color)
  - 2 tone chromatic scale (from color x to color y)
  - 3 tone divergent scale (from color x through color y to color z)
 c. Use color information within and between graphs
 
 3. show color names, hex in base R
 4. show color schemes in colorbrewer
 
#### Color visualizations
```{r}
my_cols <- c("thistle","tomato","cornsilk","cyan","chocolate")
demoplot(my_cols,"map")
demoplot(my_cols,"bar")
demoplot(my_cols,"scatter")
demoplot(my_cols,"heatmap")
demoplot(my_cols,"spine")
demoplot(my_cols,"perspective")
```

#### Working with black and white
```{r}
# gray function versus gray colors

# built in greys (0 = black, 100 = white
my_greys <- c("grey20","grey50","grey80")
demoplot(my_greys,"bar")

my_greys2 <- grey(seq(from=0.1,to=0.9,length.out=10))         
demoplot(my_greys2,"heatmap")

# converting color plots to black and white
p1 <- ggplot(d,aes(x=as.factor(cyl),y=cty,fill=as.factor(cyl))) + geom_boxplot() 
plot(p1)

# default colors look identical in black white
p1_des<- colorblindr::edit_colors(p1, desaturate)
plot(p1_des)

# custom colors not pretty, but convert ok to bw
p2 <- p1 + scale_fill_manual(values=c("red","blue","green","yellow"))
plot(p2)
p2_des<- colorblindr::edit_colors(p2, desaturate)
plot(p2_des)

```

#### using alpha transparency for histograms
```{r}
x1 <- rnorm(n=100,mean=0)
x2 <- rnorm(n=100,mean=2.7)
dFrame <- data.frame(v1=c(x1,x2))
lab <- rep(c("Control","Treatment"),each=100)
dFrame <- cbind(dFrame,lab)
str(dFrame)
h1 <- ggplot(dFrame,aes(x=v1,fill=lab))
h1 + geom_histogram(position="identity",alpha=0.5,color="black") 
```

#### Color customizations
```{r}
d <- mpg
# --------- discrete classification
# scale_fill_manual for boxplots,bars
# scale_color_manual for points, lines

# boxplot no color
p_fil <- ggplot(d,aes(x=as.factor(cyl),y=cty))
p_fil + geom_boxplot()

# boxplot default ggplot fill
p_fil <- ggplot(d,aes(x=as.factor(cyl),y=cty,fill=as.factor(cyl))) + geom_boxplot()
plot(p_fil)

# create custom color palette
my_cols <- c("red","brown","blue","orange")


# boxplot with custom colors for fill
p_fil + scale_fill_manual(values=my_cols)

# scatterplot with no color
p_col <- ggplot(d,aes(x=displ,y=cty))
p_col + geom_point(size=3)

# scatterplot default ggplot colors
p_col <- ggplot(d,aes(x=displ,y=cty,col=as.factor(cyl))) + geom_point(size=3)
plot(p_col)

# scatterplot with custom colors for point color
p_col + scale_color_manual(values=my_cols)

# ------- continuous classification (color gradient)

# default color gradient
p_grad <- ggplot(d,aes(x=displ,y=cty,col=hwy)) + geom_point(size=3)
plot(p_grad)

# custom sequential gradient (2-colors)
p_grad + scale_color_gradient(low="green", high="red")

# custom diverging gradient (3-colors)
mid <- median(d$cty)
p_grad + scale_color_gradient2(midpoint=mid,
                               low="blue",
                               mid="white",
                               high="red")

# custom diverging gradient (n-colors
p_grad + scale_color_gradientn(colors=c("blue","green","yellow","purple","orange"))
```













#### Working with color and grey

```{r,warning=FALSE,message=FALSE}
d <- mpg # use built in mpg data frame
p1 <- ggplot(d,aes(x=fl,y=hwy,group=fl)) 
p1 + geom_boxplot() # no color
p1 + geom_boxplot(fill="red") # single color (no info)

# create a palette
myColors <- c("red","green","pink","blue","orange") #no info
p1 + geom_boxplot(fill=myColors)

myColors <- c("red","green","green","blue","orange") # info
p1 + geom_boxplot(fill=myColors)



p1 + geom_boxplot(fill=gray(seq(from=0.1,to=0.9,length=5)))


# using alpha transparency for histograms
x1 <- rnorm(n=100,mean=0)
x2 <- rnorm(n=100,mean=3)
dFrame <- data.frame(v1=c(x1,x2))
lab <- rep(c("Control","Treatment"),each=100)
dFrame <- cbind(dFrame,lab)
str(dFrame)
h1 <- ggplot(dFrame,aes(x=v1,fill=lab))
h1 + geom_histogram(position="identity",alpha=0.5,color="black") 

# now go out to web and show them the palette page
p1 + geom_boxplot(fill=wes_palettes[["Royal2"]])
p1 + geom_boxplot(fill=c(grey(0.5),canva_palettes[[1]]))

# use scale fill manual

p2 <- ggplot(d,aes(x=fl,y=hwy,fill=fl)) + geom_boxplot()+ scale_fill_manual(values=wes_palettes[["Darjeeling1"]])
print(p2)


myCanvas <- canva_palettes[["Fresh and bright"]]
myCanvas <- c(gray(0.6),myCanvas)
p2 + geom_boxplot() + scale_fill_manual(values=myCanvas)
```

#### Colorbrewer page
```{r}
# visit colorbrewer page for demo
p2 + geom_boxplot() + scale_fill_brewer(palette="Blues")

# slightly different scheme for points, which have color

p3 <- ggplot(d,aes(x=displ,y=hwy,color=fl)) +
  geom_point()   +
   scale_color_brewer(palette="Spectral")
print(p3)
```

#### Continuous variables
```{r}
p3 <- ggplot(d,aes(x=displ,y=hwy,color=cty)) +
  geom_point()    
print(p3)

# use scale_color_gradient to change low and high colors
  # scale_color_gradient(
p3 +
  scale_color_gradient(low="red",high="blue")

# use scale_color_gradient2 for a 3 color gradient
z=mean(d$cty)
p3 +
  scale_color_gradient2(midpoint=z,low="red",mid="seagreen",high="cyan",space="Lab")


# use scale_color_gradientn for multicolored changes
p3+scale_color_gradientn(colours = rainbow(5))
p3+scale_color_gradientn(colours = myCanvas)
```

#### Making a heat map
```{r}
xVar <- 1:30
yVar <- 1:5
myData <- expand.grid(xVar=xVar,yVar=yVar)
head(myData)
zVar <- myData$xVar + myData$yVar + 2*rnorm(n=150)
myData <- cbind(myData,zVar)
head(myData)
p4 <- ggplot(myData,aes(x=xVar,y=yVar,fill=zVar))
p4 + geom_tile()

p4 <- ggplot(myData,aes(x=xVar,y=yVar,fill=zVar))
p4 + geom_tile() +
scale_fill_gradient2(midpoint=19,low="brown",mid=grey(0.8),high="darkblue")
```
#### New viridis color scale
```{r}
p4 + geom_tile() + scale_fill_viridis_c()
# options viridis, cividis, magma, inferno, plasma
p4 + geom_tile() + scale_fill_viridis_c(option="inferno")
```

```{r}
p4 <- p4 + geom_tile() + scale_fill_viridis_c() 
p4des<-edit_colors(p4, desaturate)
ggdraw(p4des)


```

