---
title: 'Lecture #8: Lists and Subsetting'
author: "Nicholas J. Gotelli"
date: "February 6, 2020"
output:
  html_document:
    highlight: tango
    keep_md: yes
    theme: united
  pdf_document: default
---

### Creating Vectors

#### Creating an empty vector and adding to it (avoid)

```{r, eval=FALSE}
# create an empty vector, specify mode & length
z <- vector(mode="numeric",length=0)  
print(z)

# now add elements to it

z <- c(z,5)
print(z)
# This "dynamic sizing" is very SLOW!

```


#### Creating a vector of predefined length (preferred)

#### Better to create a vector of pre-defined length

```{r, eval=FALSE}
# has 100 0s in a numeric vector
z <- rep(0,100) 
head(z)

# but would be better to start with NAs
z <- rep(NA,100)
head(z)

# but what is its type?
typeof(z)

# take advantage of coercion!
z[1] <- "Washington"
head(z)
typeof(z)
```

### Efficiently creating a vector with many names
```{r, eval=FALSE}
my_vector <- runif(100)
my_names <- paste("Species",seq(1:length(my_vector)),sep="")
names(my_vector) <- my_names
head(my_vector)
str(my_vector)
```

### Using rep to repeat elements and create vectors

```{r, eval=FALSE}
# rep for repeating elements
rep(0.5,6) # give the element (or vector!) and number of times to repeat
rep(x=0.5,times=6) # using the argument names is always prudent
rep(times=6,x=0.5) # with argument names, order is not important
my_vec <- c(1,2,3)
rep(x=my_vec,times=2) # applies to entire vectors
rep(x=my_vec,each=2) # repeat each element individually
rep(x=my_vec,times=my_vec) # what does this do?
rep(x=my_vec,each=my_vec) # and this?
```

### Using seq to create regular sequences

```{r, eval=FALSE}
seq(from=2, to=4) # set limits for integer sequences
2:4 # very common short cut with no explicit function wrappers
seq(from=2,to=4,by=0.5) # use a by function can generate real numbers
x <- seq(from=2, to=4,length=7) # sometimes easier to just specify the length
my_vec <- 1:length(x) # commonly used, but actually slow
print(my_vec)
seq_along(my_vec) # much faster for models and big data!
seq_len(5) # this is also faster than 1:5
```

### Using rnorm and runif to create vectors of random numbers
```{r, eval=FALSE}
runif(5) # 5 random uniform values betweeon 0 and 1
runif(n=3,min=100, max=101) # 3 random uniform values between 100 and 101

rnorm(6) # 6 random normal values with mean 0 and standard deviation 1
rnorm(n=5, mean=100, sd=30) # 5 random normal values with mean 100 and sd 30

## Explore distributions by sampling and plotting
library(ggplot2) # do this at the very start
z <- runif(1000) #default uniform (0,1)
qplot(x=z)
z <- rnorm(1000) #default normal (0,1)
qplot(x=z)
```

### Using sample to draw random values from an existing vector

```{r, eval=FALSE}
long_vec <- seq_len(10)
typeof(long_vec)
str(long_vec)

sample(x=long_vec) # with no other params, this reorders the vector
sample(x=long_vec, size=3) # specify a number (sampling without replacement)
sample(x=long_vec,size=16,replace=TRUE) # can generate duplicates
my_weights <- c(rep(20,5),rep(100,5)) # create a set of non-zero positive weights (integer or real)
print(my_weights)
sample(x=long_vec,replace=TRUE,prob=my_weights) # sampling with replacement and weights
sample(x=long_vec,replace=FALSE,prob=my_weights) # sampling without replacement and weights
sample(x=long_vec,replace=TRUE,prob=long_vec) # what does this do?
```

### Techniques for subsetting atomic vectors

```{r,eval=FALSE}
z <- c(3.1, 9.2, 1.3, 0.4, 7.5)

# positive index values
z[c(2,3)]

# negative index values to exclude elements
z[-c(2,3)]

# create a vector of logical elements to select conditions

z[z<3]

# equivalent to the following
tester <- z<3
print(tester)
z[tester]

# also use which() function to find subscript indicators
which(z<3)

# this works, but is overkill; just use the boolean operator
z[which(z<3)]

# can also use length() for relative positioning to last element
z[-(length(z):(length(z)-2))]

# also can subset using named vector elements
names(z) <- letters[1:5]
z[c("b","c")]


```

### Relational operators in R

```{r eval=FALSE}
# <   less than
# >   greater than
# <=  less than or equal to
# >=  greater than or equal to
# ==  equal to
```

### Logical operators
```{r eval=FALSE}
# ! not
# & and (vector)
# | or (vector)
# xor(x,y)
# && and (only for first element)
# || or (only for first element)

x <- 1:5
y <- c(1:3,7,7)
x == 2
x != 2
x == 1 & y == 7
x == 1 | y == 7
x == 3 | y == 3
xor(x==3, y==3)
x == 3 && y == 3
```

### Subscripting with missing values

```{r, eval=FALSE}
set.seed(90)
z <- runif(10) # simple integer sequence
print(z)

z < 0.5 # create logical vector
z[z < 0.5] # use as index call
which(z < 0.5) # use to get indices for logical
z[which(z < 0.5)] # does same as above

zD <- c(z,NA,NA) # contaminate it
zD[zD < 0.5] # NA values carried along!
zD[which(zD < 0.5)] # NA values dropped

```

### Matrices

```{r, eval=FALSE}
# a matrix is an atomic vector that is organized into rows and columns

my_vec <- 1:12

m <- matrix(

```
### Lists

```{r,eval=FALSE}
# Lists are atomic vectors but each element 
# can hold things of different types and different sizes

myList <- list(1:10, matrix(1:8,nrow=4,byrow=TRUE), letters[1:3],pi)
str(myList)
print(myList)


# using [] gives you a single item, which is of type list
myList[4]
myList[4] - 3 # no, can't subtract a number from a list!


# single brackets gives you only the element in that slot, which is always of type list

# to grab the object itself, use [[]]
myList[[4]]
myList[[4]] - 3 # now we have the contents

# if a list has 10 elements it is like a train with 10 cars
# [[5]] gives you the contents of car #5
##[c(4,5,6)] gives you a little train with cars 4, 5, and 6

# once the double bracket is called, you can access individual elements as before

myList[[2]]
myList[[2]][4,1]

# name list items when they are created

myList2 <- list(Tester=FALSE,littleM=matrix(1:9,nrow=3))

# named elements can be accessed with dollar sign

myList2$littleM[2,3] # get row 2, column3
myList2$littleM # show whole matrix
myList2$littleM[2,] # show second row, all columns
myList2$littleM[2] # what does this give you?

# The "unlist" strings everything back into
# a single atomic vector; coercion is used if there are mixed data types
unRolled <- unlist(myList2)
print(unRolled)


# The most common use of list: output from a linear model is a list

library(ggplot2)
yVar <- runif(10)
xVar <- runif(10)
myModel <- lm(yVar~xVar)
qplot(x=xVar,y=yVar)

# look at output in myModel
print(myModel)

# full results are in summary
print(summary(myModel))

# drill down into summary to get what we want
str(summary(myModel))
summary(myModel)$coefficients
summary(myModel)$coefficients["xVar","Pr(>|t|)"]
summary(myModel)$coefficients[2,4]

# use unlist instead

u <- unlist(summary(myModel))
print(u)

mySlope <- u$coefficients2
myPval <- u$coefficients8

```

### Data Frames

```{r,eval=FALSE}
# a data frame is a list of equal-lengthed vectors, each of which is a column

varA <- 1:12
varB <- rep(c("Con","LowN","HighN"),each=4)
varC <- runif(12)
dFrame <- data.frame(varA,varB,varC,stringsAsFactors=FALSE)
print(dFrame)
str(dFrame)

# add another row with rbind
# make sure you add a list, with each item corresponding to a column

# newData <- data.frame(list(varA=13,varB="HighN",varC=0.668),stringsAsFactors=FALSE)
newData <- list(varA=13,varB="HighN",varC=0.668)
print(newData)
str(newData)

# now bind them
dFrame <- rbind(dFrame,newData)
str(dFrame)
tail(dFrame)


# adding another column is a little easier

#newVar <- data.frame(varD=runif(13))
newVar <- runif(13)
dFrame <- cbind(dFrame,newVar)
head(dFrame)

```

### Important Distinctions Between Lists and Matrices

```{r, eval=FALSE}

# create a matrix and data frame with same structures
zMat <- matrix(data=1:30,ncol=3,byrow=TRUE)
zDframe <- as.data.frame(zMat) # coerce it

str(zMat)    # an atomic vector with 2 dimensions
str(zDframe) # note horizontal layout of variabes!

head(zDframe) # note automatic variable names
head(zMat) # note identical layout

# element referencing is the same in both
zMat[3,3]
zDframe[3,3]

# so is column referencing

zMat[,3]
zDframe[,3]
zDframe$V3 # note use of $ and named variable column
# and row referencing
zMat[3,]
zDframe[3,] # note variable names and row number shown

# what happens if we reference only one dimension?

zMat[2] # takes the second element of atomic vector (column fill)
zDframe[2] # takes second atomic vector (= column) from list
zDframe["V2"]
zDframe$V2
```

### Eliminating missing values
```{r, eval=FALSE}
# use complete.cases with atomic vector
print(zD)

complete.cases(zD)

zD[complete.cases(zD)] # clean them out

which(!complete.cases(zD)) # find NA slots

# use with a matrix

m <- matrix(1:20,nrow=5)
m[1,1] <- NA
m[5,4] <- NA
print(m)

m[complete.cases(m),] 

# now get complete cases for only certain columns!
m[complete.cases(m[,c(1,2)]),] # drops row 1
m[complete.cases(m[,c(2,3)]),] # no drops
m[complete.cases(m[,c(3,4)]),] # drops row 4
m[complete.cases(m[,c(1,4)]),] # drops 1&4

```



### Techniques for assignments and subsetting matrices and data frames
```{r,eval=FALSE}
# same principle applied to both dimensions of a matrix
m <- matrix(data=1:12,nrow=3)
dimnames(m) <- list(paste("Species",LETTERS[1:nrow(m)],sep=""),paste("Site",1:ncol(m),sep=""))
print(m)

# subsetting based on elements
m[1:2,3:4]
# same subsetting based on character strings (but no negative elements)
m[c("SpeciesA","SpeciesB"), c("Site3","Site4")]

# use blanks before or after comma to indicate full rows or columns
m[1:2, ]    

m[ ,3:4]

# use logicals for more complex subsetting

# e.g. select all columns for which the totals are > 15

# first try this logical
colSums(m) > 15
m[ , colSums(m) > 15]


# e.g. select all rows for which the row total is 22
m[rowSums(m)==22, ]

# note == for logical equal and != for logical NOT equal
m[rowSums(m)!=22, ]

# e.g., choose all rows for which numbers for site 1 are less than 3
# AND choose all columns for which the numbers for species A are less than 5

# first, try out this logical for rows
m[ ,"Site1"]<3

# add this in and select with all columns
m[m[ ,"Site1"]<3, ]

# and try this logical for columns
m["SpeciesA", ]<5

# add this in and select with all rows
m[ ,m["SpeciesA", ]<5]

# now combine both
m[m[ ,"Site1"]<3,m["SpeciesA", ]<5]

# and compare with full m
print(m)


# caution! simple subscripting to a vector changes the data type!
z <- m[1, ]
print(z)
str(z)

# to keep this as a matrix, must add the drop=FALSE option

z2 <- m[1, ,drop=FALSE]
print(z2)
str(z2)

# caution #2, always use both dimensions, or you will select a single matrix element

m2 <- matrix(data=runif(9),nrow=3)
print(m2)
m2[2, ]

# but now this will just pull the second element
m2[2]

# probably should specify row and column indicators
m2[2,1]
# also use logicals for assignments, not just subsetting
m2[m2>0.6] <- NA
print(m2)


# A few changes for working with data frames:

data <-read.csv(file="antcountydata.csv",header=TRUE,sep=",",stringsAsFactors=FALSE)
str(data)

# the data frame is a list of vectors, so it is set up like a matrix
data[3,2]

# you can specify just the column names

dataNames <- data[c("state","county")]
str(dataNames)

# or in matrix style
dataNames <- data[ ,c("county", "ecoregion")]
str(dataNames)


# as before, with matrices, selecting only a single column changes it
# from a data frame to a vector
dataNames <- data[ ,"county"]
str(dataNames)

```
