---
title: 'Lecture #17: Control Structures II'
author: "Nicholas J. Gotelli"
date: "March 30, 2017"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---
### `for` loops
* The workhorse function for doing repetitive tasks
* Universal in all computer languages
* Controversial in R
    - often not necessary (use vectorized operations!)
    - very slow with binding operations (c,rbind,cbind,list) 
    - many operations can be handled by special faimly of `apply` functions
    
#### Anatomy of a `for` loop
```
for (var in seq) { # start of for loop

# body of for loop 

} # end of for loop

var is a counter variable that will hold the current value of the loop
seq is an integer vector (or a vector of character strings) that defines the starting and ending values of the loop
```
It is traditional in the statistics literature to use variables `i,j,k` to indicate counters. e.g.

$$A = \sum_{i=1}^S n_i$$

You can also use `t` for time, but remember that `t()` is also an R function that transposes a matrix. R is smart enough to recognize the difference, but it could cause problems for others.

There are many ways to indicate the sequence, and the length of our loops will almost always be related to the size of a vector that we are working with. The most efficient structure to to create the sequence is `seq_along(x)` where x is the vector whose length defines the number of repetitions

Let's see a simple example:

```{r, eval=FALSE}
myDat <- signif(runif(10),digits=2)
for (i in seq_along(myDat)) {
  cat("loop number = ",i,"vector element = ", myDat[i],"\n")
}
```

#### Tip #1: Don't do things in the loop if you do not need to!
```{r, eval=FALSE}
myDat <- vector(mode="numeric",length=10)
for (i in seq_along(myDat)) {
  myDat[i] <- signif(runif(1),digits=2) # do this outside of the loop!
  cat("loop number = ",i,"vector element = ", myDat[i],"\n")
}
```

#### Tip #2: Do not change object dimensions (`cbind,rbind,c,list`) in the loop!
```{r, eval=FALSE}
myDat <- runif(1)
for (i in 2:10) {
  temp <- signif(runif(1),digits=2) # do this outside of the loop!
  myDat <- c(myDat,temp) # do not change vector size in the loop!
  cat("loop number = ",i,"vector element = ", myDat[i],"\n")
}

```

#### Tip #3: Do not write a loop if you can vectorize an operation
```{r, eval=FALSE}
myDat <- 1:10
for (i in seq_along(myDat)) {
  myDat[i] <-  myDat[i] + myDat[i]^2
  cat("loop number = ",i,"vector element = ", myDat[i],"\n")
}

# No loop is needed here!
z <- 1:10
z <- z + z^2
print(z)


```
#### Tip #4: Always be alert to the distinction between the counter variable `i` and the vector element `z[i]`
```{r, eval=FALSE}
z <- c(10,2,4)
for (i in seq_along(z)) {
  cat("i =",i,"z[i] = ",z[i],"\n")
}
# What is value of i at this point?
print(i)
```

#### Tip #5: Use `next` to skip certain elements in the loop

```{r, eval=FALSE}
z <- 1:20
# What if we want to work with only the odd-numbered elements?

for (i in seq_along(z)) {
  if(i %% 2==0) next
  print(i)
}

# Another method, probably faster (why?)
z <- 1:20
z <- z[z %% 2!=0] # contrast with logical expression in previous if statement!
length(z)
for (i in seq_along(z)) {
  cat("i = ",i,"z[i] = ",z[i],"\n")
}
```

# Tip #6: Use `break` to set up a conditional to break out of loop early


