---
title: 'Lecture #3: Working With R Studio'
author: "Nicholas J. Gotelli"
date: "January 24, 2017"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---

___

### Window appearance
- sizing of window boundaries
- window screens

### Options to set
- `general` set default directory
- restore project and documents
- `code` do not insert matching parentheses and quotes
- default keybindings (not Vim or eMacs)
- `pane layout` - as set up
- `appearance` - choices for color, font, size

### File Names and Types
- `Plain Text` human-readable characters (none are unprinted)
    - not proprietary
    - readable by any computer system
    - readable wtih any `Plain Text Editor`
- `WYSIWYG` what you see is what you get
- `WYSIWYW` what you see is what you want
    - plain text codes to indicate formatting
    - no time lost using the mouse
    - separates writing from formatting, allowing you to concentrate on the former
- `Markdown` a language of simple plain text codes for rendering (= "knitting" a file into a different output format (`.html`,`.pdf`, and `.doc`)

- `.Rmd` is Rstudio's special blend of Markdown (a text rendering language)
- `.md` is the more generic Markdown file type. In RStudio, the intermediate `.md` files are not (in the default state) preserved.

### Languages
- `LaTeX` Typesetting language invented in the 1970s by computer programmer Donald Knuth for proper rendering of text and equations. Hundreds of commands. Precise, but somewhat complicated to use.
- `Markdown` Greatly simplified version of LaTeX with a small number of commands to master. Easy to use, text looks great.
- `yaml`^[Originally "Yet Another Markdown Language", but more recently "YAML Ain't no Markdown Language", which is a self-referencing acronym]. Header material for `.Rmd` files that can be changed manually or by selecting options.
- `R` Computer language for manipulating numbers, text, graphics, statistics.

### Using R-markdown
- pull up cheatsheet
- new `.Rmd` file

### Basic text formatting
- headers
- text
- markup
    * italic
    * bold-face
    * subscript
    * superscript
    * strikethrough
    * quotations
    
### Spacing, manual breaks, lines
- line spacing effects
- two extra spaces at the end of a line for a manual break
- lists
    * indented
    * numbered
    
### Links  
- links [linking text](website address)
- footnotes^[Footnoted text, consecutively at bottom of page]
- images 
```
![image link](website address)
```

### Tables

```
First Header  | Second Header
------------- | -------------
Content Cell  | Content Cell
Content Cell  | Content Cell
```

- also the `kable` package for later
```
{r, echo=TRUE,results='asis'}
library(knitr)
kable(head(iris))
```

```{r, echo=TRUE, results='asis'}
library(knitr)
kable(head(iris))
```


### Fencing
- in line fencing `to shade`
- in line R code for calculations `r 3 + pi`
- simple fencing to set off large chunks of plain text


### Equations

* in-line `$`
* centered `$$`

- basic math and text spacing handled by LateX

```
$$y = a + b$$
```

$$y = a + b$$

#### Subcripts

```
$$H_0 = Z_{a + b}$$
```

$$H_0 = Z_{a + b}$$

####Superscripts

```
$$S = cA^z$$

```

$$S = cA^z$$


Contrast this with the construction of superscripts in Markdown

```

```
```
Markdown z^x^
```

z^x^

```
LaTeX: $z^x$
```
$z^x$

```
Markdown: z^x + 1^
```

z^x + 1^

```
LaTeX: $z^{x + 1}$
```

$z^{x + 1}$

```
Markdown: z^x+1^
```
z^x+1^

```
LaTeX: $z^{x+1}$
```
$z^{x+1}$


- elements can be coupled and nested

$$S=cA^z_1 + z_{2 + x}$$


```
$$S=cA^z_1 + z_{2 + x}$$

```

#### Fractions and Greek Symbols

$$\alpha = \frac{\beta}{\delta + \gamma_x}$$
```
$$\alpha = \frac{\beta}{\delta + \gamma_x}$$

```

#### Summation signs

$$z = \sum_{i=1}^X{K}$$

```
$$z = \sum_{i=1}^X{K}$$

```

#### "Escaping" the \ special character in LaTeX

Use `\backslash`

```
$$\backslash \alpha \le b \backslash$$

```

$$\backslash \alpha \le b \backslash$$

#### Rendering plain text in a LaTex Equation

$$P(Occurrence Of Species A) = Z$$
```
$$P(Occurrence Of Species A) = Z$$
```

$$P(\mbox{Occurrence Of Species A}) = Z$$

```
$$P(\mbox{Occurrence Of Species A}) = Z$$
```

### R Code in Chunks

* `CTRL-ALT-I` inserts a new "chunk" of R code
```{r}
# Use comments extensively in ALL of your coding!
Pred <- seq(1,10)     # make a vector of integers from 1 to 10
Res <- runif(10)      # generate 10 draws from a random uniform (0,1) distribution

# print the random numbers
print(Res)

# plot the graph
plot(x=Pred,y=Res,type="b")
```

### Using chunk options to control printing of code and output
```{r}
plot(runif(50),cex=10*runif(50),col="goldenrod")
print(rnorm(n=10, mean=10, sd=2))
```

### Writing R scripts

- new script
- code executes and prints in console
- graphs in output window
- `CTRL - ENTER` to run a single line of code
- `CTRL - SHIFT - ENTER` to source the entire script
- code to illustrate:
```
# Use comments extensively throughout your code
Time <- seq(1,10) # Create a vector of integers, 1 to 10
Resp <- runif(10) # Create 10 random uniform (0,1) variates
#
print(Time)
print(Resp)
#
# Plot the data
plot(x=Time,y=Resp,type="b")
```
### Converting `.R` files to `.html` (creating an R notebook)
- use notebook script
- roxygen comments ``#`


### `Purling` to strip and consolidate R chunks from a Markdown file
- commands executed from console

```
library(knitr)
purl("FileName.Rmd")
```
These commands will create a file `FileName.R` that has all of the R code and none of the markdown text. R comments are still retained.

### Bonus: A Cool New Markdown Editor!

- Demo of `Typora` a WYSIWYG text editor at https://typora.io/

