---
title: 'Lecture #23 ggplot III'
author: "Nicholas J. Gotelli"
date: "November 13, 2018"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---
### ggplot (continued)

#### Preliminaries
```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(ggthemes)
library(patchwork)
library(TeachingDemos)
char2seed("Dark Star")
d <- mpg

# instead of subsetting, just map an aesthetic
p1 <- ggplot(data=d,mapping=aes(x=displ,y=hwy)) +
  geom_point(mapping=aes(color=drv)) + geom_smooth()
print(p1)

# Conversely, map the smoother, but not the points
p1 <- ggplot(data=d,mapping=aes(x=displ,y=hwy)) +
  geom_point() + geom_smooth(mapping=aes(color=drv))
print(p1)

# also, subset in the first layer to eliminate some data entirely
# instead of subsetting, just map an aesthetic
p1 <- ggplot(data=d[d$drv!="4",],mapping=aes(x=displ,y=hwy)) +
  geom_point(mapping=aes(color=drv)) + geom_smooth()
print(p1)

```


#### Bar plots
```{r,message=FALSE, warning=FALSE}
# use to plot the counts of rows for a categorical variable
table(d$drv)
p1 <- ggplot(data=d,mapping=aes(x=drv)) + geom_bar(color="black",fill="goldenrod")
print(p1)

# aesthetic mapping gives multiple groups for each bar
p1 <- ggplot(data=d,mapping=aes(x=drv,fill=fl)) + geom_bar()
print(p1)


# stacked, but need to adjust color transparency, which is "alpha"
p1 <- ggplot(data=d,mapping=aes(x=drv,fill=fl)) + geom_bar(alpha = 1/4, position="identity")
print(p1)

# better to use position = fill for stacking, but with equivalent height
p1 <- ggplot(data=d,mapping=aes(x=drv,fill=fl)) + geom_bar(position="fill")
print(p1)

# best to use position = dodge for multiple bars
p1 <- ggplot(data=d,mapping=aes(x=drv,fill=fl)) + geom_bar(position="dodge",color="black",size=1)
print(p1)

# more typical "bar plot" has heights as the values themselves
dTiny <- tapply(X=d$hwy,INDEX=as.factor(d$fl),FUN=mean) #calculate the means
dTiny <- data.frame(hwy=dTiny) # create a single-column data frame
dTiny <- cbind(fl=row.names(dTiny),dTiny) # 

p2 <- ggplot(data=dTiny, mapping=aes(x=fl,y=hwy,fill=fl)) +
  geom_col()
print(p2)



#### Use a box plot instead of standard "means" bars!


```{r, warning=FALSE, message=FALSE}
# basic boxplot is simple and informative
p1 <- ggplot(data=d,mapping=aes(x=fl,y=hwy,fill=fl)) +
               geom_boxplot()
print(p1)

# now overlay the raw data
p1 <- ggplot(data=d,mapping=aes(x=fl,y=hwy)) +
               geom_boxplot(fill="thistle",outlier.shape=NA) + 
  geom_point()
 # geom_point(position=position_jitter(width=0.2,height=0.7),color="grey60")

print(p1)
```
### Using Colors

1. Aesthetics
  a. Colors that are attractive
    - large geoms (fills) - pale colors
    - small geoms(lines,points) - bright colors
  b. Colors that highlight elements
    - pale, grey to dee-emphasize
    - bright or saturated colors to emphasize
  c. Colors that are visible to the color blind

2. Information content
  a. Discrete scale
   - colors to group similar treatments
   - neutral colors (black,grey,white) to indicate control groups
   - Symbolic colors (heat=red, cool = blue, photosynthesis/growth=green, oligotrophic=blue, eutrophic=brown, infected=red)
   - Colors that map to chemical stains or gels, or even colors of organisms
 b. Continuous scale
  - monochromatic (differing shades of 1 color)
  - 2 tone chromatic scale (from color x to color y)
  - 3 tone divergent scale (from color x through color y to color z)
 c. Use color information with and between graphs
    



#### Preliminaries

```{r}
library(ggplot2)
library(ggthemes)
library(patchwork)
library(TeachingDemos)
library(wesanderson)
char2seed("Dark Star")
```
#### Working with color and grey

```{r,warning=FALSE,message=FALSE}
d <- mpg # use built in mpg data frame
p1 <- ggplot(data=d,mapping=aes(x=fl,y=hwy,group=fl))
p1 + geom_boxplot() # no color
p1 + geom_boxplot(fill="red") # single color (no info)

# create a palette
myColors <- c("red","green","pink","blue","orange") #no info
p1 + geom_boxplot(fill=myColors)

myColors <- c("red","green","green","blue","orange") # info
p1 + geom_boxplot(fill=myColors)



p1 + geom_boxplot(fill=gray(seq(from=0.1,to=0.9,length=5)))


# using alpha transparency for histograms
x1 <- rnorm(n=100,mean=0)
x2 <- rnorm(n=100,mean=3)
dFrame <- data.frame(v1=c(x1,x2))
lab <- rep(c("Control","Treatment"),each=100)
dFrame <- cbind(dFrame,lab)
str(dFrame)
h1 <- ggplot(data=dFrame,mapping=aes(x=v1,fill=lab))
h1 + geom_histogram(position="identity",alpha=0.5,color="black") 

# now go out to web and show them the palette page
p1 + geom_boxplot(fill=wes_palettes[["Royal2"]])
p1 + geom_boxplot(fill=c(grey(0.5),canva_palettes[[1]]))

# use scale fill manual

p2 <- ggplot(data=d, mapping=aes(x=fl,y=hwy,fill=fl)) + geom_boxplot()+ scale_fill_manual(values=wes_palettes[["Darjeeling1"]])
print(p2)


myCanvas <- canva_palettes[["Fresh and bright"]]
myCanvas <- c(gray(0.6),myCanvas)
p2 + geom_boxplot() + scale_fill_manual(values=myCanvas)
```

#### Colorbrewer page
```{r}
# visit colorbrewer page for demo
p2 + geom_boxplot() + scale_fill_brewer(palette="Blues")

# slightly different scheme for points, which have color

p3 <- ggplot(data=d, mapping=aes(x=displ,y=hwy,color=fl)) +
  geom_point()   +
   scale_color_brewer(palette="Spectral")
print(p3)
```

#### Continuous variables
```{r}
p3 <- ggplot(data=d, mapping=aes(x=displ,y=hwy,color=cty)) +
  geom_point()    
print(p3)

# use scale_color_gradient to change low and high colors
  # scale_color_gradient(
p3 +
  scale_color_gradient(low="red",high="blue")

# use scale_color_gradient2 for a 3 color gradient
z=mean(d$cty)
p3 +
  scale_color_gradient2(midpoint=z,low="red",mid="seagreen",high="cyan",space="Lab")


# use scale_color_gradientn for multicolored changes
p3+scale_color_gradientn(colours = rainbow(5))
p3+scale_color_gradientn(colours = myCanvas)
```

#### Making a heat map
```{r}
xVar <- 1:30
yVar <- 1:5
myData <- expand.grid(xVar=xVar,yVar=yVar)
head(myData)
zVar <- myData$xVar + myData$yVar + 2*rnorm(n=150)
myData <- cbind(myData,zVar)
head(myData)
p4 <- ggplot(data=myData,mapping=aes(x=xVar,y=yVar,fill=zVar))
p4 + geom_tile()

p4 <- ggplot(data=myData,mapping=aes(x=xVar,y=yVar,fill=zVar))
p4 + geom_tile() +
scale_fill_gradient2(midpoint=19,low="brown",mid=grey(0.8),high="darkblue")
```
#### New viridis color scale
```{r}
p4 + geom_tile() + scale_fill_viridis_c()
# options viridis, cividis, magma, inferno, plasma
p4 + geom_tile() + scale_fill_viridis_c(option="inferno")
```

```{r}
library(colorblindr) # devtools::install_github("clauswilke/colorblindr")
library(colorspace) # install.packages("colorspace", repos = "http://R-Forge.R-project.org") --- colorblindr requires the development version
# this also installs cowplot 
library(cowplot)
p4 <- p4 + geom_tile() + scale_fill_viridis_c() 
p4des<-edit_colors(p4, desaturate)
ggdraw(p4des)


```

