---
title: 'Lecture #4: Working With R Studio II'
author: "Nicholas J. Gotelli"
date: "September 6 2018"
output:
  html_document:
    highlight: tango
    keep_md: yes
    theme: united
  pdf_document: default
---

___

### Links  
- links [linking text](website address)
- footnotes^[Footnoted text, consecutively at bottom of page]
- images 
```
![image link](website address)
```

### Tables

```
First Header  | Second Header
------------- | -------------
Content Cell  | Content Cell
Content Cell  | Content Cell
```

- also the `kable` package for later
```
{r, echo=TRUE,results='asis'}
library(knitr)
kable(head(iris))
```

```{r, echo=TRUE, results='asis'}
library(knitr)
kable(head(iris))
```




### Equations

* in-line `$`
* centered `$$`

- basic math and text spacing handled by LateX

```
$$y = a + b$$
```

$$y = a + b$$

#### Subcripts

```
$$H_0 = Z_{a + b}$$
```

$$H_0 = Z_{a + b}$$

####Superscripts

```
$$S = cA^z$$

```

$$S = cA^z$$


Contrast this with the construction of superscripts in Markdown

```

```
```
Markdown z^x^
```

z^x^

```
LaTeX: $z^x$
```
$z^x$

```
Markdown: z^x + 1^
```

z^x + 1^

```
LaTeX: $z^{x + 1}$
```

$z^{x + 1}$

```
Markdown: z^x+1^
```
z^x+1^

```
LaTeX: $z^{x+1}$
```
$z^{x+1}$


- elements can be coupled and nested

$$S=cA^z_1 + z_{2 + x}$$


```
$$S=cA^z_1 + z_{2 + x}$$

```

#### Fractions and Greek Symbols

$$\alpha = \frac{\beta}{\delta + \gamma_x}$$
```
$$\alpha = \frac{\beta}{\delta + \gamma_x}$$

```

#### Summation signs

$$z = \sum_{i=1}^X{K}$$

```
$$z = \sum_{i=1}^X{K}$$

```

#### "Escaping" the \ special character in LaTeX

Use `\backslash`

```
$$\backslash \alpha \le b \backslash$$

```

$$\backslash \alpha \le b \backslash$$

#### Rendering plain text in a LaTex Equation

$$P(Occurrence Of Species A) = Z$$
```
$$P(Occurrence Of Species A) = Z$$
```

$$P(\mbox{Occurrence Of Species A}) = Z$$

```
$$P(\mbox{Occurrence Of Species A}) = Z$$
```

### R Code in Chunks

* `CTRL-ALT-I` inserts a new "chunk" of R code
```{r}
# Use comments extensively in ALL of your coding!
Pred <- seq(1,10)     # make a vector of integers from 1 to 10
Res <- runif(10)      # generate 10 draws from a random uniform (0,1) distribution

# print the random numbers
print(Res)

# plot the graph
plot(x=Pred,y=Res,type="b")
```

### Using chunk options to control printing of code and output
```{r}
plot(runif(50),cex=10*runif(50),col="goldenrod")
print(rnorm(n=10, mean=10, sd=2))
```

### Writing R scripts

- new script
- code executes and prints in console
- graphs in output window
- `CTRL - ENTER` to run a single line of code
- `CTRL - SHIFT - ENTER` to source the entire script
- code to illustrate:
```{r}
# First comment to explain what this program is doing.
# Be expansive and describe it in great detail. This may seem trivial, but will become increasingly important as you create complex programs.
# Simple script to examine the distribution of the product of two uniform variables
# Make sure it is readable. Use complete sentences, not cryptic phrases.
# 6 September 2018
# NJG

# Preliminaries
library(ggplot2)
set.seed(100)
library(TeachingDemos) # use this to set the random number seed from a character string
char2seed("green tea")
char2seed("green tea",set=FALSE)

#

# Global variables
nRep <- 10000

# Create or read in data
ranVar1 <- rnorm(nRep)
# print(ranVar1)
head(ranVar1)
tail(ranVar1)

ranVar2 <- rnorm(nRep)


# visualize data
qplot(x=ranVar1)
qplot(x=ranVar1,color=I("black"),fill=I("goldenrod"))
qplot(x=ranVar2,color=I("black"),fill=I("grey50"))


# create product vector
ranProd <- ranVar1*ranVar2
length(ranProd)
str(ranProd)
head(ranProd)

# visualize product vector
qplot(x=ranVar1,color=I("black"),fill=I(grey(0.5)))

# do other stuff..
```
### Converting `.R` files to `.html` (creating an R notebook)
- use notebook script
- roxygen comments ``#`


### `Purling` to strip and consolidate R chunks from a Markdown file
- commands executed from console

```
library(knitr)
purl("FileName.Rmd")
```
These commands will create a file `FileName.R` that has all of the R code and none of the markdown text. R comments are still retained.

### Bonus: A Cool New Markdown Editor!

- Demo of `Typora` a WYSIWYG text editor at https://typora.io/

