---
title: 'Lecture #23: Apply Functions'
author: "Nicholas J. Gotelli"
date: "April 20, 2017"
output:
  html_document:
    highlight: tango
    keep_md: yes
    theme: united
  pdf_document: default
---
### On-line resources

- [Advanced R](http://adv-r.had.co.nz/)
- [R Inferno](http://www.burns-stat.com/pages/Tutor/R_inferno.pdf)
- [Bioinformatics Zen](http://www.bioinformaticszen.com/faq/)
- [R grouping functions: sapply vs. lapply vs. apply vs. tapply vs. by vs. aggregate](http://stackoverflow.com/questions/3505701/r-grouping-functions-sapply-vs-lapply-vs-apply-vs-tapply-vs-by-vs-aggrega)
- [ The R apply function - a tutorial with examples](https://www.r-bloggers.com/the-r-apply-function-a-tutorial-with-examples/)
- [The infamous apply function](http://rforpublichealth.blogspot.com/2012/09/the-infamous-apply-function.html)
- [Using apply, sapply, lapply in R](https://www.r-bloggers.com/using-apply-sapply-lapply-in-r/)




### `Apply` Functions
sapply
lapply
apply
tapply
vapply
mapply
rapply
Map
by
aggregate
sweep
replicate
doBy
data.table

ase function   Input   Output   plyr function 

---------------------------------------
aggregate        d       d       ddply + colwise 
apply            a       a/l     aaply / alply 
by               d       l       dlply 
lapply           l       l       llply  
mapply           a       a/l     maply / mlply 
replicate        r       a/l     raply / rlply 
sapply           l       a       laply 


### The `apply` function

```
apply(x, margin, fun,...)
x : a vector, matrix, or array (all elements of same type)
margin: 1 for first dimension, 2 for second dimension, c(1,2) for both dimensions
fun: the function that is to be applied over the margin
...: additional optional arguments that can be applied to the function
```
```{r,eval=FALSE}
m <- matrix(1:12,3,4,byrow=TRUE) # short-hand for a 3 x 4 matrix; fill rows
print(m)


myCV <- function(x=runif(5)) {
  x <- sd(x)/mean(x)  # rescale proportions
  return(x)
}
myCV()

# get cvs of the rows with a for loop
myOut <- vector(mode="numeric",length=nrow(m))
for (i in 1:nrow(m)){
  myOut[i] <- myCV(m[i,]) # operate on all columns in row i
}
print(myOut)
# get cvs of the rows with apply; no need to create output structure!
mO <- apply(m,1,myCV)
print(mO) 

# get cvs of the columns
mO <- apply(m,2,myCV) 
print(mO)

# what happens applying the function over both margins?
mO <- apply(m,c(1,2),myCV)
print(mO)

# Not what we expected, but the input to myCV is now 
# one number:
myCV(3)
sd(3)

# so how do we get CV for the entire matrix?
myCV(m)

# no need to call the myCV function
# write an anonymous (=unnamed) function on the fly
mO <- apply(m,1, function(x) sd(x)/mean(x))
print(mO)        

# problems with dimension of output
# first, a simple rescaling of column totals:

mO <- apply(m,2,function(x) x/sum(x))
print(mO)

# but now try this for rows
mO <- apply(m,1,function(x) x/sum(x))
print(mO)

# what happened?
# output from apply is given in each column of output structure!

mO <- t(apply(m,1,function(x) x/sum(x)))
print(mO)

# how do we rescale to the matrix total?

mO <- apply(m,c(1,2),function(x) x/sum(x))
print(mO)
# because we are operating on elements, not rows or columns
# just use a vector operation

mO <- m/sum(m)
print(mO)

mO0 <- m/sum(m)
print(mO0)

# EcoSimR sim2 reshuffles row values
mO <- t(apply(m,1,function(x) sample(x)))
print(mO)
print(m)

# shuffle columns
mO <- apply(m,2,function(x) sample(x))
print(mO)

# shuffle all elements
mO <- sample(m)
print(mO)

mO <- matrix(sample(m),3,4,byrow=TRUE)
print(mO)

# alternatively, we could set up the structure for 
# the receiving variable

mO <- m
mO[,] <- sample(m)
print(mO)

# what if the output from apply is not always the same length?

print(m)
mO <- apply(m,c(1,2),function(x) runif(x))

print(mO)
str(mO)
# now we have a list, and each element is a vector of random numbers
mO[[2]]

```
### The `replicate` function

```
replicate(n,expression,simplify)
n : number of replications
expression : any r expression, including a function or operation
simplify : default="array", which will generate an array with dimensions 1 higher than output structure. simplify=TRUE will generate a matrix or vector. simplify=FALSE will generate a list)
```

```{r,eval=FALSE}

# first set this up as a matrix

myOut <- matrix(data=0,3,5)

# fill with a for loop
for (i in 1:nrow(myOut)) {
  for (j in 1:ncol(myOut)) {
   myOut[i,j] <- runif(1)
  }
}
print(myOut)

# silly to loop here. Just create myOut and fill directly
myOut <- matrix(data=runif(15),3,5)
print(myOut)

# each column of output is one of the replicates
mO <- replicate(5, 100 + runif(3),simplify=TRUE)
print(mO)

# try making random matrices
mO <- replicate(5,matrix(runif(6),3,2),simplify="array")
print(mO)

# what if the output is not all of the same length? 
#  in that case, replicate will always make a list, 
# regardless of setting for simplify
mO <- replicate(5, 100 + runif(sample.int(10,1)),simplify=TRUE)
print(mO)
```
### The `lapply` function

```
lapply(x,fun,...)
x : a vector or list (or an expression that generates this)
fun : a function that is applied to each element of x
... : optional arguments to the function
```
```{r,eval=FALSE}
# since a data frame is a list of vectors, this is a common
# application

d <- read.table(file="antcountydata.csv",
                header=TRUE,
                sep=",",
                stringsAsFactors=FALSE)
str(d)

# with a for loop, calculate the mean of variables 6 and 7

myMeans <- vector(mode="numeric",length=2)
z <- 0 # uggh, need to set up another counter!
for (i in 5:6) {
  z <- z + 1
  myMeans[z] <-mean(d[[i]])
}
print(myMeans)

myMeans <- lapply(d[c(5,6)],mean)
print(myMeans) # now a list with names!
print(unlist(myMeans)) # use unlist to put back in simple vector

```
### The `sapply` function

```
sapply(x,fun,...,simplify=TRUE,USE.NAMES=TRUE)

Behaves just like lapply, but the output is more user-friendly,
with labels. If simplify=FALSE and USE.NAMES=TRUE you have lapply, which gives you a named list as output.
```
```{r,eval=FALSE}

myMeans <- sapply(d[c(5,6)],mean)
print(myMeans) # now a vector with names!
```

### The `aggregate` function

```
aggregate(x,by,fun)
x : vector or list
by : list of grouping elements each as long as the data frame
fun : a function to apply to the data in each of the groups
```
```{r,eval=FALSE}
# Calculate group means for ecoregions
myMeans <- matrix(0,5,2) #set up with zeroes
myFac <- as.numeric(as.factor(d$ecoregion))

# brittle double for-loop
z <- 0
for (i in 5:6) {
    z <- z + 1
  for (j in 1:5) {
    myMeans[j,z] <- mean(d[myFac==j,i])
  }
}
print(myMeans)
    
# easier with aggregate, which gives a data frame    
myMeans <- aggregate(d[c(5,6)],by=list(d$ecoregion),mean)
print(myMeans)
str(myMeans)

```
### The `tapply` function

```
Behaves like aggregate but operates on vectors not lists
tapply(x,index,fun)
```

```{r,eval=FALSE}
myMeans <- tapply(d[,6],d$ecoregion,mean)
#myMeans <- tapply(d$ecoregion,as.matrix(d[,c(5,6)]),mean)
print(myMeans)
str(myMeans)

```


