---
title: 'Lecture #6: Working With R Studio'
author: "Nicholas J. Gotelli"
date: "January 30 2020"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---

___

### Three ways to use R
- from console (in R studio and in shell)
- from .Rmd markdown file (markdown)
- from .R script file

#### General Rstudio shortcuts
```
control 1 : move to script panel
control 2 : move to console panel
control shift 1 : move to script panel and expand
control shift 2 : move to console panel and expand
control l : to clear the console
```
#### Working with Rmarkdown
- render to view pane not to browser
- set up headers in YAML
- keep first chunk in place
- set options in chunks for `echo, eval`

#### Rmarkdown shortcuts
```
CTRL-SHIFT-k : knits the markdown file to html/pdf/word
CTRL-OPTION-i : inserts a new R chunk
```

#### Fencing
- for plain text (show a sample)
- chunk for R code  (show a few lines)
- even a chunk for bash (show pwd command)

#### Three languages in one file!
- rmarkdown (for text, links, images)
- r code (for computing)
- LaTeX for rendering math equations

#### All three in one sentence:

The value of $\pi$ is `r pi`.


#### Details of \LaTeX


### Equations

* in-line `$`
* centered `$$`

- basic math and text spacing handled by LateX

```
$$y = a + b$$
```

$$y = a + b$$

#### Subcripts

```
$$H_0 = Z_{a + b}$$
```

$$H_0 = Z_{a + b}$$

#### Superscripts

```
$$S = cA^z$$

```

$$S = cA^z$$


- elements can be coupled and nested

$$S=cA^z_1 + z_{2 + x}$$


```
$$S=cA^z_1 + z_{2 + x}$$

```

#### Fractions and Greek Symbols

$$\alpha = \frac{\beta}{\delta + \gamma_x}$$
```
$$\alpha = \frac{\beta}{\delta + \gamma_x}$$

```

#### Summation signs

$$z = \sum_{i=1}^X{K}$$

```
$$z = \sum_{i=1}^X{K}$$

```

#### "Escaping" the \ special character in LaTeX

Use `\backslash`

```
$$\backslash \alpha \le b \backslash$$

```

$$\backslash \alpha \le b \backslash$$

#### Rendering plain text in a LaTex Equation

$$P(Occurrence Of Species A) = Z$$
```
$$P(Occurrence Of Species A) = Z$$
```

$$P(\mbox{Occurrence Of Species A}) = Z$$

```
$$P(\mbox{Occurrence Of Species A}) = Z$$
```
### R Code in Chunks

* `CTRL-ALT-I` inserts a new "chunk" of R code
```{r}
# Use comments extensively in ALL of your coding!
Pred <- seq(1,10)     # make a vector of integers from 1 to 10
Res <- runif(10)      # generate 10 draws from a random uniform (0,1) distribution

# print the random numbers
print(Res)

# plot the graph
plot(x=Pred,y=Res,type="b")
```

### Using chunk options to control printing of code and output
```{r}
plot(runif(50),cex=10*runif(50),col="goldenrod")
print(rnorm(n=10, mean=10, sd=2))
```

### Working with R scripts




### Writing R scripts

- new script
- code executes and prints in console
- graphs in output window
- `CTRL - ENTER` to run a single line of code
- `CTRL - ENTER` to run a block of selected lines
- `CTRL - SHIFT - ENTER` to source the entire script and echo
- `CTRL - SHIFT - S` to sourc the script with no echo
- `CTRL - SHIFT - R` to enter a labeled or unlabeled section divider

- code to illustrate:

```{r}
# First comment to explain what this program is doing.
# Be expansive and describe it in great detail. This may seem trivial, but will become increasingly important as you create complex programs.
# Simple script to examine the distribution of the product of two uniform variables
# Make sure it is readable. Use complete sentences, not cryptic phrases.
# 6 September 2018
# NJG

# Preliminaries
library(ggplot2)
set.seed(100)
library(TeachingDemos) # use this to set the random number seed from a character string
char2seed("green tea")
char2seed("green tea",set=FALSE)

#

# Global variables
nRep <- 10000

# Create or read in data
ranVar1 <- rnorm(nRep)
# print(ranVar1)
head(ranVar1)
tail(ranVar1)

ranVar2 <- rnorm(nRep)


# visualize data
qplot(x=ranVar1)
qplot(x=ranVar1,color=I("black"),fill=I("goldenrod"))
qplot(x=ranVar2,color=I("black"),fill=I("grey50"))


# create product vector
ranProd <- ranVar1*ranVar2
length(ranProd)
str(ranProd)
head(ranProd)

# visualize product vector
qplot(x=ranVar1,color=I("black"),fill=I(grey(0.5)))

# do other stuff..
```
### Converting `.R` files to `.html` (creating an R notebook)
- use notebook script
- roxygen comments ``#`


### `Purling` to strip and consolidate R chunks from a Markdown file
- commands executed from console

```
library(knitr)
purl("FileName.Rmd")
```
These commands will create a file `FileName.R` that has all of the R code and none of the markdown text. R comments are still retained.


