---
title: 'Lecture #5: Markdown II'
author: "Nicholas J. Gotelli"
date: "31 January 2023"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---

___



- `.Rmd` is Rstudio's special blend of Markdown (a text rendering language)
- `.md` is the more generic Markdown file type. In RStudio, the intermediate `.md` files are not (in the default state) preserved.

### Languages
- `LaTeX` Typesetting language invented in the 1970s by computer programmer Donald Knuth for proper rendering of text and equations. Hundreds of commands. Precise, but somewhat complicated to use.
- `Markdown` Greatly simplified version of LaTeX with a small number of commands to master. Easy to use, text looks great.
- `yaml`^[Originally "Yet Another Markdown Language", but more recently "YAML Ain't no Markdown Language", which is a self-referencing acronym]. Header material for `.Rmd` files that can be changed manually or by selecting options.
- `R` Computer language for manipulating numbers, text, graphics, statistics.

### Using R-markdown
- pull up cheatsheet
- new `.Rmd` file

### Basic text formatting
- headers `#` to `######`
    *numbering from options
- text
- markup
    * italic `*<text>*`
    * bold-face `**<text>**`
    * subscript `~<text>~`
    * superscript `^<text>^`
    * strikethrough `~~<text>~~`
    * quotations `><text with no closing mark`
    
### Spacing, manual breaks, lines
- line spacing effects
- two extra spaces at the end of a line for a manual break
- lists
    * indented
    * numbered
    
### Links  
- links `[linking text](website address)`
- footnotes^[Footnoted text, consecutively at bottom of page]
- images 
```
![image link](website address)
```

### Tables

```
First Header  | Second Header
------------- | -------------
Content Cell  | Content Cell
Content Cell  | Content Cell
```

- also the `kable` package for later
```
{r, echo=TRUE,results='asis'}
library(knitr)
kable(head(iris))
```

```{r, echo=TRUE, results='asis'}
library(knitr)
kable(head(iris))
```


### Fencing
- in line fencing `to shade`
- in line R code for calculations `r 3 + pi`
- simple fencing to set off large chunks of plain text

### Links  
- links [linking text](website address)
- footnotes^[Footnoted text, consecutively at bottom of page]
- images 
```
![image link](website address)
```

### Tables

```
First Header  | Second Header
------------- | -------------
Content Cell  | Content Cell
Content Cell  | Content Cell
```

- also the `kable` package for later
```
{r, echo=TRUE,results='asis'}
library(knitr)
kable(head(iris))
```

```{r, echo=TRUE, results='asis'}
library(knitr)
kable(head(iris))
```




### Equations

* in-line `$`
* centered `$$`

- basic math and text spacing handled by LateX

```
$$y = a + b$$
```

$$y = a + b$$

#### Subcripts

```
$$H_0 = Z_{a + b}$$
```

$$H_0 = Z_{a + b}$$

#### Superscripts

```
$$S = cA^z$$

```

$$S = cA^z$$


- elements can be coupled and nested

$$S=cA^z_1 + z_{2 + x}$$


```
$$S=cA^z_1 + z_{2 + x}$$

```

#### Fractions and Greek Symbols

$$\alpha = \frac{\beta}{\delta + \gamma_x}$$
```
$$\alpha = \frac{\beta}{\delta + \gamma_x}$$

```

#### Summation signs

$$z = \sum_{i=1}^X{K}$$

```
$$z = \sum_{i=1}^X{K}$$

```

#### "Escaping" the \ special character in LaTeX

Use `\backslash`

```
$$\backslash \alpha \le b \backslash$$

```

$$\backslash \alpha \le b \backslash$$

#### Rendering plain text in a LaTex Equation

$$P(Occurrence Of Species A) = Z$$
```
$$P(Occurrence Of Species A) = Z$$
```

$$P(\mbox{Occurrence Of Species A}) = Z$$

```
$$P(\mbox{Occurrence Of Species A}) = Z$$
```

### R Code in Chunks

* `CTRL-ALT-I` inserts a new "chunk" of R code
```{r}
# Use comments extensively in ALL of your coding!
Pred <- seq(1,10)     # make a vector of integers from 1 to 10
Res <- runif(10)      # generate 10 draws from a random uniform (0,1) distribution

# print the random numbers
print(Res)

# plot the graph
plot(x=Pred,y=Res,type="b")
```

### Using chunk options `echo` and `eval` to control printing of code and output
```{r}
plot(runif(50),cex=10*runif(50),col="goldenrod")
print(rnorm(n=10, mean=10, sd=2))
```
### Setting Up A Project

#### Suggested Folders
- OriginalData
- CleanData
- Scripts
- Functions
- Outputs
- Plots

#### Using path statements
`getwd()` from console for current paths statement
- this is an absolute path
- starting point is always root of project directory, so we don't have to write the whole path
- `\path\to\my\projectfolder\Scripts\filename` absolute path
- `Scripts\filename` relative path
- use relative path both for opening a script and writing to an output or a plot
- keep .rmd files up in the root (too complicated)
### Writing R scripts

- new script
- code executes and prints in console
- graphs in output window
- `CTRL - ENTER` to run a single line of code
- `CTRL - SHIFT - ENTER` to source and echo the entire script
- `CTRL - SHIFT - S` to source the entire script without echoing console output
- code to illustrate:
```{r}
# First comment to explain what this program is doing.
# Be expansive and describe it in great detail. This may seem trivial, but will become increasingly important as you create complex programs.
# Simple script to examine the distribution of the product of two uniform variables
# Make sure it is readable. Use complete sentences, not cryptic phrases.
# 6 September 2018
# NJG

# Preliminaries
library(ggplot2)
set.seed(100)
library(TeachingDemos) # use this to set the random number seed from a character string
char2seed("green tea")
char2seed("green tea",set=FALSE)

#

# Global variables
nRep <- 10000

# Create or read in data
ranVar1 <- rnorm(nRep)
# print(ranVar1)
head(ranVar1)
tail(ranVar1)

ranVar2 <- rnorm(nRep)


# visualize data
qplot(x=ranVar1)


# create product vector
ranProd <- ranVar1*ranVar2
length(ranProd)
str(ranProd)
head(ranProd)


# do other stuff..
```
### Converting `.R` files to `.html` (creating an R notebook)
- use notebook script
- roxygen comments ``#`


### `Purling` to strip and consolidate R chunks from a Markdown file
- commands executed from console

```
library(knitr)
purl("FileName.Rmd")
```
These commands will create a file `FileName.R` that has all of the R code and none of the markdown text. R comments are still retained.

