---
title: 'Lecture #9: Data Structures in R'
author: "Nicholas J. Gotelli"
date: "February 9, 2017"
output:
  html_document:
    highlight: tango
    keep_md: yes
    theme: united
  pdf_document: default
---
## History of R
- Early languages FORTRAN, C
- S language in 1970s; programming plus stats, matrix algebra, graphics later
- Open-source R in 1992; first stable version in 2000

## Advantages of R
- interpreted (for interactive use)
- graphics, statistics
- very active community of contributors
- works on multiple platforms

## Disadvantages of R
- interpreted (slow speed)
- lazy evaluation
- functions hard to learn
- poorly documented
- unreliable packages
- problems with big data
   - subroutines coded for C or Fortran
   - Julia?

## R as a general programming language
- data structures and types
- writing functions
- for loops
- if..then, while statements
- structured programming
- annotation
- naming of variables
- using random numbers
- structured programming
- pseudocode

### Using the assignment operator

```{r, eval=FALSE}
# Using the assignment operator
x <- 5 # preferred
y = 4 # legal but not used except in function defaults
y = y + 1.1
print(y)
y <- y + 1.1
print(y)
```

# Variable names

```{r}
z <- 3 # Begin with lower case letter 
plant_height <- 3.3  # a bit long
plant.height <- 4.2 # avoid periods
plantHeight <- 10 # best option "camelCaseFormatting"
```

## R's Four Data Types

Dimensions  | Homogeneous | Heterogeneous
------------- | ------------- | -------------
1-dimension  | Atomic Vector | List
2-dimensions  | Matrix | Data Frame
n-dimenions   | (array) |          |

## Types of Atomic Vectors
- character strings
- integers
- double
- integers, doubles are "numeric"
- logical
- (factor)

### One dimensional atomic vectors
```{r, eval=FALSE}
# the combine function
z <- c(3.2, 5, 5, 6) 
print(z)
typeof(z)
is.numeric(z)

# c() always "flattens" to an atomic vector
z <- c(c(3,4),c(5,6)) 
print(z)

# character strings with single or double quotes
z <- c("perch","bass",'trout') 
print(z)

# use both with an internal quote
z <- c("This is only 'one' character string", 'a second')
print(z)
typeof(z)
is.character(z)

# building logicals
# Boolean, not with quotes, all caps
z <- c(TRUE,TRUE,FALSE) 
# avoid abbreviations T, F which will work
print(z)
typeof(z)
is.logical(z)
is.integer(z)

```
### Three Properties of a Vector

#### Type

```{r, eval=FALSE}
z <- c(1.1, 1.2, 3, 4.4)
typeof(z) # gives type
is.numeric(z) # is. gives logical
as.character(z) # as. coerces variable
print(z)
typeof(z)

```

#### Length
```{r, eval=FALSE}
length(z) # gives number of elements
length(y) # throws error if variable does not exist

```

#### Names
```{r, eval=FALSE}
z <- runif(5)
# optional attribute not initially assigned
names(z) 
print(z)
# add names later after variable is created
names(z) <- c("chow","pug","beagle","greyhound","akita")
print(z)

# add names when variable is built (with or without quotes)
 z2 <- c(gold=3.3, silver=10, lead=2)
print(z2)

# reset names
names(z2) <- NULL

# names can be added for only a few elements
# names do not have to be distinct, but often are
names(z2) <- c("copper","zinc")
print(z2)

```

### Three Notable Features of Atomic Vectors

#### Coercion

```{r, eval=FALSE}
# All atomics are of the same type
# if they are different, R coerces them
# logical -> integer -> double -> character

a <- c(2, 2.0)
print(a)
typeof(a)

b <- c("purple","green")
typeof(b)

d <- c(a,b)
print(d)
typeof(d)

# "Mistakes" in numeric variables convert to strings
# Very useful when working with logical variables

a <- runif(10)
print(a)

# Comparison operators yield a logical result
a > 0.5

# do math on a logical and it converts to an integer!

# How many elements are greater than 0.5?
sum(a > 0.5)

# What proportion of the vector elements are greater than 0.5?

mean(a > 0.5)

#Qualifying exam question! Approximately what proportion of observations drawn from a normal (0,1) distribution are larger than 2.0?

mean(rnorm(1000) > 2)
```


#### Vectorization

```{r, eval=FALSE}
# adding a constant to a vector
z <- c(10,20,30)
z + 1

# what happens when vectors are added?

y <- c(1,2,3)
z + y

# results is an "element by element" operation on the vector
# most vector operations can be done this way

z^2

```



#### Recycling
```{r, eval=FALSE}
# but what if vector lengths are not equal?
z <- c(10,20,30)
x <- c(1,2)
z + x

# warning is issued by calculation is still made
# shorter vector is always "recycled"
# works with scalars (= vector of length(1))

```

### Creating Vectors
#### Creating an empty vector and adding to it (avoid)
```{r,eval=FALSE}
# create an empty vector, specify mode & length
z <- vector(mode="numeric",length=0)  
print(z)

# now add elements to it

z <- c(z,5)
print(z)
# This "dynamic sizing" is very SLOW!
```

#### Creating a vector of predefined length (preferred)
```{r, eval=FALSE}
### Better to create a vector of pre-defined length

# has 100 0s in a numeric vector
z <- rep(0,100) 
head(z)

# but would be better to start with NAs
z <- rep(NA,100)
head(z)

# but what is its type?
typeof(z)

# take advantage of coercion!
z[1] <- "Washington"
head(z)
typeof(z)
```

#### Efficiently creating a vector with many names
```{r, eval=FALSE}
MyVector <- runif(100)
MyNames <- paste("Species",seq(1:length(MyVector)),sep="")
names(MyVector) <- MyNames
str(MyVector)
```

