---
title: 'Batch Processing'
author: "Nicholas J. Gotelli"
date: "6 April 2023"

output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---
## Batch processing

### Build a set of random files
```{r}
##################################################
# function: file_builder
# create a set of random files for regression
# input: file_n = number of files to create
#       : file_folder = name of folder for random files
#       : file_size = c(min,max) number of rows in file
#       : file_na = number on average of NA values per column
# output: set of random files
#------------------------------------------------- 
file_builder <- function(file_n=10,
                        file_folder="RandomFiles/",
                        file_size=c(15,100),
                        file_na=3){
  # check for folder with random files. If present, destroy it.
  # next create a fresh copy of the folder and new files
  if(dir.exists(file_folder)) unlink(file_folder,
                                     recursive=TRUE)
  dir.create(file_folder)
  
for (i in seq_len(file_n)) {
file_length <- sample(file_size[1]:file_size[2],size=1) # get number of rows
var_x <- runif(file_length) # create random x
var_y <- runif(file_length) # create randon y
df <- data.frame(var_x,var_y) # bind into a data frame
bad_vals <- rpois(n=1,lambda=file_na) # determine NA number
df[sample(nrow(df),size=bad_vals),1] <- NA # random NA in var_x
df[sample(nrow(df),size=bad_vals),2] <- NA # random NA in var_y

# create label for file name with padded zeroes
file_label <- paste(file_folder,
                       "ranFile",
                       formatC(i,
                       width=3,
                       format="d",
                       flag="0"),
                       ".csv",sep="")

# set up data file and incorporate time stamp and minimal metadata
write.table(cat("# Simulated random data file for batch processing","\n",
                    "# timestamp: ",as.character(Sys.time()),"\n",
                    "# NJG","\n",
                    "# ------------------------", "\n",
                    "\n",
                    file=file_label,
                    row.names="",
                    col.names="",
                    sep=""))

# now add the data frame
write.table(x=df,
            file=file_label,
            sep=",",
            row.names=FALSE,
            append=TRUE)


}
}
```
### Run regression model and extract stats

```{r}

##################################################
# function: reg_stats
# fits linear model, extracts statistics
# input: 2-column data frame (x and y)
# output: slope, p-value, and r2
#------------------------------------------------- 
reg_stats <- function(d=NULL) {
             if(is.null(d)) {
               x_var <- runif(10)
               y_var <- runif(10)
               d <- data.frame(x_var,y_var)
             }
  . <- lm(data=d,d[,2]~d[,1])
  . <- summary(.)
  stats_list <- list(slope=.$coefficients[2,1],
                    p_val=.$coefficients[2,4],
                    r2=.$r.squared)
  return(stats_list)

}
```

### Body of script for batch processing of regression models
```{r, warning=FALSE,eval=FALSE}

#--------------------------------------------
# Global variables
file_folder <- "RandomFiles/"
n_files <- 100
file_out <- "StatsSummary.csv"
#--------------------------------------------

# Create 100 random data sets
file_builder(file_n=n_files)
file_names <- list.files(path=file_folder,
                         pattern="*.csv")

# Create data frame to hold file summary statistics
ID <- seq_along(file_names)
file_name <- file_names
slope <- rep(NA,n_files)
p_val <- rep(NA,n_files)
r2 <- rep(NA,n_files)

stats_out <- data.frame(ID,file_name,slope,p_val,r2)

# batch process by looping through individual files
for (i in seq_along(file_names)) {
  data <- read.table(file=paste(file_folder,file_names[i],sep=""),
                     sep=",",
                     header=TRUE) # read in next data file
  
  d_clean <- data[complete.cases(data),] # get clean cases
  
  . <- reg_stats(d_clean) # pull regression stats from clean file
  stats_out[i,3:5] <- unlist(.) # unlist, copy into last 3 columns
  
}
# set up output file and incorporate time stamp and minimal metadata
  write.table(cat("# Summary stats for ",
                    "batch processing of regression models","\n",
                    "# timestamp: ",as.character(Sys.time()),"\n",
                    "# NJG","\n",
                    "# ------------------------", "\n",
                    "\n",
                    file=file_out,
                    row.names="",
                    col.names="",
                    sep=""))
  
# now add the data frame
  write.table(x=stats_out,
              file=file_out,
              row.names=FALSE,
              col.names=TRUE,
              sep=",",
              append=TRUE)

```

## Organizing Source Files

- don't write to .csv unless leaving R environment
- use saveRDS amd readRDS to store R objects
- create an uberscript

```
# typical contents of uberscript
source ('Functions.R')
source('ModelScript.R') # creates object with saveRDS
source('GraphicsScript.R') # loads object with readRDS

```
## Logging & Progress Bars

Use to keep a written record of the activities generated by a script.
- useful for debugging
- useful for annotation and keeping track of files

```{r}
library(log4r)
my_logfile='logfile.txt'
console_echo=FALSE

if (file.exists(my_logfile)) unlink(my_logfile)
my_console_appender = console_appender(layout = default_log_layout())
my_file_appender = file_appender(my_logfile, append = TRUE, 
                                 layout = default_log_layout())
output <- list(my_console_appender,my_file_appender)
if (!console_echo) output <- list(my_file_appender)

my_logger <- logger(threshold = "INFO", 
                           appenders=output)

l <- function(t="",index=NULL) {
  if(is.null(index)) info(my_logger, t) else {
    index <- paste0(" |index=",index)
    info(my_logger,t,index)
  }
}

l()
l('text message')

l('before the loop')
l()
for (i in 1:5) {
  l('in the loop',index=i)
  print(runif(1))
#  print(x) # check the log to see where the program broke
}
l()
l('after the loop')
```

##
Using a progress bar

```{r, eval=FALSE}
library(progress)
## Basic
pb <- progress_bar$new(total = 100)
for (i in 1:100) {
  pb$tick()
  Sys.sleep(0.1)
}

## ETA
pb <- progress_bar$new(
  format = "  downloading [:bar] :percent eta: :eta",
  total = 100, clear = FALSE, width= 60)
for (i in 1:100) {
  pb$tick()
  Sys.sleep(0.1)
}

## Elapsed time
pb <- progress_bar$new(
  format = "  downloading [:bar] :percent in :elapsed",
  total = 100, clear = FALSE, width= 60)
for (i in 1:100) {
  pb$tick()
  Sys.sleep(0.1)
}

## Spinner
pb <- progress_bar$new(
  format = "(:spin) [:bar] :percent",
  total = 30, clear = FALSE, width = 60)
for (i in 1:30) {
  pb$tick()
  Sys.sleep(0.1)
}


```

## "Old School" progress bar (no packages or formatting needed!)
```{r}
for (i in 1:100) {
  Sys.sleep(0.1)
  if(i%%10==0) cat(i) else cat('.')
  # if(i%%10==0) cat(i) else if(i%%5==0) cat('.')
}
```

## Measuring the elapsed time of parts of your code
```{r}


library(pracma)
tic()
for (i in 1:100) {
  Sys.sleep(0.1)
}
toc()
```

