---
title: 'Lecture #25: dplyr'
author: "Nicholas J. Gotelli"
date: "April 19, 2018"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---
### dplyr

#### Wrangling tools
##### Regular Expressions
- good for manipulating strings and cleaning data
- find and replace, but mostly for manipulating columns, not rows
#### Subsetting rows and columns
- create a boolean expression for selection
- apply to rows or columns to generate an n-element boolean vector
- use this vector to select and subset rows and columns

#### dplyr
- consistent set of grammar
- use to modify actual layout of the data

### tidy data
- need to be in the "long" form
- each cell is a value (= smallest object of measurement)
- each row is a unique observation
- each column is an attribute of the observation

### why not tidy?
- no consistent formatting structure
- alternate forms required for some programs
- ease of input in spreadsheets
- storage and size

### Preliminaries

```{r, message=FALSE,warning=FALSE}
library(reshape2)
library(tidyr)
library(ggplot2)
library(TeachingDemos)
char2seed("sharpei")
```
### Build species x site matrix

```{r, message=FALSE,warning=FALSE}
species <- 5
sites <- 8
abundanceRange <- 1:10
mFill <- 0.4

vec <- rep(0,species*sites)  # set up empty vector
abun <- sample(x=abundanceRange,
               size=round(mFill*length(vec)),
               replace=TRUE) # sample abundance counts
vec[seq_along(abun)] <- abun # fill initial cells with counts
vec <- sample(vec) # reshuffle the values
aMat <- matrix(vec,nrow=species)
rownames(aMat) <- rownames(aMat, 
                           do.NULL=FALSE, 
                           prefix="Species")
colnames(aMat) <- colnames(aMat, 
                           do.NULL=FALSE, 
                           prefix="Site")
print(aMat)
```
#### Convert to long form

```{r,warning=FALSE,message=FALSE}
# using melt function from reshape2 for matrices and arrays
. <- melt(aMat)
print(.)
. <- melt(aMat,varnames=c("Species","Site"),value.name="Abundance")
print(.)

# now convert to a data frame
aFrame <- data.frame(cbind(Species=rownames(aMat),aMat))
print(aFrame)

# use gather function from tidyr for data frames
 . <- gather(aFrame,Site1:Site8,key="Site",value="Abundance")
 print(.)
```
#### Build a Subject x Time Experimental Matrix

```{r}

# construct data frame; typical excel spreadsheet format
Treatment <- rep(c("Control","Treatment"),each=5)
Subject <- 1:10
T1 <- rnorm(10)
T2 <- rnorm(10)
T3 <- rnorm(10)
eFrame <- data.frame(Treatment=Treatment,Subject=Subject,T1=T1,T2=T2,T3=T3)
str(eFrame)
eFrame$Subject <- as.character(eFrame$Subject)
str(eFrame)
print(eFrame)

# now use gather to turn this into long format
. <- gather(eFrame,T1:T3,key="Time",value="Response")
print(.)
str(.)
# set up new variable "Time" as a factor
.$Time <- as.factor(.$Time)
str(.)
```
Abundance.Matrix <- cast(MyData.Long,Species~Site,value="Abundance")
print(Abundance.Matrix)
