---
title: 'Lecture #20: Randomization Tests'
author: "Nicholas J. Gotelli"
date: "April 3, 2018"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---
### General Steps in a Randomization Test
1. Define a metric $X$ as a single number to represent pattern
2. Calculate $X_{obs}$, the metric for the empirical (= observed) data
3. Randomize or reshuffle the data in a way that would uncouple association between the observed data and their assignment to treatments. Otherwise, the randomized data set should have the same dimensions and properties as the original data set.
4. For this single randomization, calculate $X_{sim}$, the metric for the simulated (= randomized) data.
5. Repeat steps (3) and (4) 1000 (or more times) to get a distribution for $H_{sim}$
6. Estimate the tail probability of the observed metric, given the null distribution $(p(X_{obs}|H_0))$.

#### Preliminaries
```{r}
library(ggplot2)
library(TeachingDemos)
#char2seed("Cruel April")
```
#### Read in data
```{r}

####### ###########################################
# function: readData
# read in (or generate) data set for analysis
# input: file name (or nothing, for this demo)
# output: 3 column data frame of observed data (ID,x,y)
#------------------------------------------------- 
readData <- function(z=NULL) {
                if(is.null(z)){
                  xObs <- 1:20
                  yObs <- xObs + 10*rnorm(20)
                  dF <- data.frame(ID=seq_along(xObs),xObs,yObs)} # set up data frame                 
# print(qplot(x=xObs,y=yObs)) # peek at input data
return(dF)
}
```

#### Calculate Metric
```{r}

##################################################
# function: getMetric
# calculate metric for randomization test
# input: 2-column data frame for regression
# output: regression slope
#------------------------------------------------- 
getMetric <- function(z=NULL) {
                if(is.null(z)){
                  xObs <- 1:20
                  yObs <-  xObs + 10*rnorm(20)
                  z <- data.frame(ID=seq_along(xObs),xObs,yObs)} # set up data frame                 
. <- lm(z[,3]~z[,2])
. <- summary(.)
. <- .$coefficients[2,1]

slope <- .
return(slope)
}
```
### Create randomization
```{r}

##################################################
# function: shuffleData
# randomize data for regression analysis
# input: 3-column data frame (ID,xVar,yVar)
# output: 3-column data frame (ID,xVar,yVar)
#------------------------------------------------- 
shuffleData <- function(z=NULL) {
                if(is.null(z)){
                  xObs <- 1:20
                  yObs <- xObs + 3*rnorm(20)
                  z <- data.frame(ID=seq_along(xObs),xObs,yObs)} # set up data frame                 
z[,3] <- sample(z[,3]) # use sample function with defaults to reshuffle column

return(z)
}
```

#### Calculate p value from simulated metrics

```{r}
##################################################
# function: getPVal
# calculate p value from simulation
# input: list of observed metric, and vector of simulated metrics
# output: lower, upper tail probability values
#------------------------------------------------- 
getPVal <- function(z=NULL) {
                    if(is.null(z)){
                      z <- list(xObs=runif(1),xSim=runif(1000))}
                      pLower <- mean(z[[2]]<=z[[1]])
                      pUpper <- mean(z[[2]]>=z[[1]])
return(c(pL=pLower,pU=pUpper))
                    }
```
#### Create histogram of simulated and observed metric

```{r}
##################################################
# function: plotRanTest
# create ggplot of histogram of simulated values
# input: list of observed metric and vector of simulated metrics
# output: saved ggplot graph
#------------------------------------------------- 
plotRanTest <- function(z=NULL) {
                if(is.null(z)){
                  z <- list(rnorm(1),rnorm(1000)) }
dF <- data.frame(ID=seq_along(z[[2]]),simX=z[[2]])
p1 <- ggplot(data=dF,mapping=aes(x=simX))
p1 + geom_histogram(mapping=aes(fill=I("goldenrod"),color=I("black"))) +
geom_vline(aes(xintercept=z[[1]],col="blue")) 

                  }
```
#### Now use functions to do the work

```{r}
nSim <- 1000
Xsim <- rep(NA,nSim)
dF <- readData()
Xobs <- getMetric(dF)

for (i in seq_len(nSim)) {
Xsim[i] <- getMetric(shuffleData(dF))
}

slopes <- list(Xobs,Xsim)
getPVal(slopes)
plotRanTest(slopes)
```

