---
title: "Lecture #10: Data curation"
author: "Nicholas J. Gotelli"
date: "9 March 2021"
output: 
  html_document: 
    highlight: tango
    theme: united
---
## Data curation
* GitHub repository   
* Student contract   
* Metadata   
* [Metadata template](https://gotellilab.github.io/Bio381/Scripts/Feb02/ExcelDataTemplate.xlsx)   
* Example of Lauren's data 

## Exporting and importing data

First create a tiny data set in Excel:
```
# comments at the top
# beaucoup metadata
ID, Treatment, Biomass, Notes
1, Control, 30.3, 
2, HighN, 13.0, 
3, HighN, NA, broken scale
4, Control, 35.3,
```
- Save as .csv
- Inspect in RStudio Editor

### Use `read.table` to bring in data
```{r, eval=FALSE}
my_data <- read.table(file="path/to/data.csv",
                    header=TRUE,
                    sep=",",
                    comment.char="#")

# inspect object
str(my_data)

# now add a column
my_data$newVar <- runif(4)
head(my_data)
```

### Use `write.table` to export to a data file

```{r, eval=FALSE}

write.table(x=my_data,
            file="Path/To/OutputFileName.csv",
            HEADER=TRUE,
            sep=",")
```



But this is not a good way to save or share data objects if we are working in R. Some researchers use the `save()` function, which preserves the whole environment, but once it is restored with `load()`, the variable names cannot be changed. It is better to use `saveRDS().

### `saveRDS()`: useful when you are working only in R
```{r, eval=FALSE}
saveRDS(my_data, file="Path/To/FileName.RDS") # .RDS suffix is not required, but good for clarity
```

This only saves a single R object as a binary, but remember, you can bundle up many things into a single list!

Use `readRDS()` to restore it.

### `readRDS()`
```{r, eval=FALSE}
data_in <-readRDS("FileName.RDS") 
```

