---
title: 'Lecture #12: Data Entry, Experimental Designs'
author: "Nicholas J. Gotelli"
date: "February 22, 2018"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---
### Data analysis recap
1. Begin with observed data
2. Choose an appropriate distribution
  - bounded versus unbounded distributions
  - continuous versus integer variables
3. Use `fitdistr` to get MLE parameters for data
4. Examine mean and variance
5. What would mean and variances and sample sizes be for groups in your experiment?
6. Simulate a data set with these values
7. Use simple ANOVA/regression tests for p value
8. Explore effects of increasing/decreasing sample sizes
9. Explore effects of increasing/decreasing differences among mean (= effect size)

### Data Curation
- GitHub repository
- Student contract
- Metadata
- Metadata template
- example of Lauren's data

### Data Read Statement
```{r,eval=FALSE}
data <-read.table(file="FileName.csv",row.names=1,header=TRUE,sep=",",stringsAsFactors=FALSE)
```

### Omitting NA values
```{r, eval= FALSE}
dataClean <- data[complete.cases(data),] # removes all rows with NA values

# to get complete cases for just certain columns 
dataCleanCols <-data[complete.cases(data[,5:6]),]
                                   
```
 
