---
title: 'Lecture #11: Complete Cases, Tribbles, Long/Wide Data, and Functions'
author: "Nicholas J. Gotelli"
date: "October 2, 2018"
output:
  html_document:
    highlight: tango
    keep_md: yes
    theme: united
  pdf_document: default
---
### Eliminating missing values with `complete.cases()`

```{r, eval=TRUE}

# preliminaries
library(tidyverse)
set.seed(99)


z <- 1:10 # simple integer sequence
z <- sample(z) # reshuffle the ordering
print(z)

z < 4 # create logical vector
z[z < 4] # use as index call
which(z < 4) # use to get indices for logical
z[which(z < 4)] # does same as above

zD <- c(z,NA,NA) # contaminate it
zD[zD < 4] # NA values carried along!
zD[which(zD < 4)] # NA values dropped

# use complete.cases with atomic vector
print(zD)

complete.cases(zD)

zD[complete.cases(zD)] # clean them out

which(!complete.cases(zD)) # find NA slots

# use with a matrix

m <- matrix(1:20,nrow=5)
m[1,1] <- NA
m[5,4] <- NA
print(m)

m[complete.cases(m),] 

# now get complete cases for only certain columns!
m[complete.cases(m[,c(1,2)]),] # drops row 1
m[complete.cases(m[,c(2,3)]),] # no drops
m[complete.cases(m[,c(3,4)]),] # drops row 4
m[complete.cases(m[,c(1,4)]),] # drops 1&4

```



### Converting between long and wide forms of data
```{r}
# Introducing the "tribble", a transposed tibble
# that is easy to type directly in to your own code

repMeasDat <- tribble(
  ~Subject, ~Treatment, ~Time1, ~Time2, ~Time3,
  #-----------------------------
  1, "Control", 1,1,2,
  2, "Control", 1,4,1,
  3, "Control", 2,2,2,
  4, "Heated", 5,6,6,
  5, "Heated", 7,7,5,
  6, "Heated", 7,5,5
  #------------------------------
)

print(repMeasDat)
```
This is a very common data format, in which each row is a subject. There are two treatments, and the three columns give the measurements on each subject at 3 different times. 

Unfortunately, this is not in the "long" format, which requires that each row represents a different observation. In this case, the observations should be the individual measurements, not the subjects.

Here is how we convert this to a correct format:



```{r}
repMeasDat %>%
gather(Time1:Time3, key="Time",value="Response")  
```

We first list the set of variables that need to be gathered (Time1:Time3). Next, we provide a `key`, which is the name of the new variable (in this case "Time"). Finally, we add a value, which is the measure that is associated with each row (in this case "Response"). Notice that we did not reference "Subject" or "Treatment". Their values are properly repeated in each of the appropriate rows. Thus, each row corresponds to a unique observation in the experiment. In this case, there were 3 subjects in each of 2 treatments, and each were measured 3 times. Thus, there are 3 x 3 x 2 rows in the new data frame.

Let's add one other refinement, using dplyr's `arrange` function to re-order the rows so that the 3 measurements from a single subject appear on successive lines in the data frame:

```{r}
longData <- repMeasDat %>%
gather(Time1:Time3, key="Time",value="Response") %>%
arrange(Subject)
print(longData)
```
The reason the long format is important, is that most statistical models and summaries in R operate this way. For example, to calculate averages for different times and treatments, we would use

```{r}
longData %>%
group_by(Time,Treatment) %>%
summarize(AvgRes = mean(Response)) %>%
arrange(Treatment,Time)
```


Now let's convert this back to wide format, using the `spread` function which is roughly the inverse of `gather`

```{r}
wideData <- longData %>%
spread(key=Time,value=Response)
print(wideData)
```

