---
title: 'Probability Distributions'
author: "Nicholas J. Gotelli"
date: "March 11, 2021"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

### Probability distributions in R
#### Discrete distributions

- Poisson
    * Range: [0,$\infty$]
    * Parameters: size = number of events (trials), rate = $\lambda$
    * Interpretation: Distribution of events that occur during a fixed time interval or sampling effort with a constant rate of independent events
    
- Binomial
    * Range: [0, # of trials]
    * Parameters: size= number of trials; p = probability of positive outcome
    * Interpretation: Distribution of number of successful independent dichotomous trials, with constant p; resembles normal with large $\lambda$, or exponential with small $\lambda$
    
- Negative Binomial
    * Range: [0, $\infty$]
    * Parameters: size=number of successes; p = probability of success
    * Interpretation: Distribution of number of failures in a series of independent Bernouli trials, each with p = probability of a success. Generates a discrete distribution that is more heterogeneous ("overdispersed") than Poisson
    
#### Continuous distributions

- Uniform
    * Range: [min,max]
    * Parameters: min = minimum boundary; max = maximum boundary
    * Interpretation: Distribution of a value that is equally likely within a specified range
    
- Normal
    * Range: [$-\infty,\infty$]
    * Parameters: mean = central tendency; SD = standard deviation
    * Interpretation: Symmetric bell-shaped curve with unbounded tails
    
- Gamma $\Gamma$
    * Range: [0,$\infty$]
    * Parameters: shape, scale
    * Interpretation: mean=$shape*scale$, variance=$shape*scale^2$; generates a variety of shapes (including normal and exponential) for positive continuous variables
    
- Beta $\beta$
    * Range: [0,1] (can be rescaled to any range by simple multiplication and addition)
    * Paramters: shape1, shape2
    * Interpretation: if shape1 and shape 2 are integers, interpret as a coin toss, with shape1 = # of successes + 1, shape2 = # of failures + 1. Gives distribution of value of p, estimated from data, which can range from exponential through uniform through normal (but all are bounded). Setting shape1 and shape2 <1 yields u-shaped distributions.

### The "grammar" of probability distributions in R
- `d` gives probability density function
- `p` gives cumulative distribution function
- `q` gives quantile function (the inverse of `p`)
- `r` gives random number generation

Combine these with the base name of the function. For example `rbinom` gives a set of random values drawn from a binomial, whereas `dnorm` gives the density function for a normal distribution. There are many probability distributions available in R, but we will discuss only 7 of them.

#### Poisson distribution

```{r}
library(ggplot2)
library(MASS)
#-------------------------------------------------
# Poisson distribution
# Discrete X >= 0
# Random events with a constant rate lambda
# (observations per time or per unit area)
# Parameter lambda > 0

# "d" function generates probability density
hits <- 0:10
myVec <- dpois(x=hits,lambda=1)
qplot(x=hits,y=myVec,geom="col",color=I("black"),fill=I("goldenrod"))

myVec <- dpois(x=hits,lambda=2)
qplot(x=hits,y=myVec,geom="col",color=I("black"),fill=I("goldenrod"))

hits <- 0:15
myVec <- dpois(x=hits,lambda=6)
qplot(x=hits,y=myVec,geom="col",color=I("black"),fill=I("goldenrod"))


hits <- 0:15
myVec <- dpois(x=hits,lambda=0.2)
qplot(x=hits,y=myVec,geom="col",color=I("black"),fill=I("goldenrod"))

sum(myVec)  # sum of density function = 1.0 (total area under curve)

# for a Poisson distribution with lambda=2, 
# what is the probability that a single draw will yield X=0?

dpois(x=0,lambda=2)

# "p" function generates cumulative probability density; gives the 
# "lower tail" cumulative area of the distribution

hits <- 0:10
myVec <- ppois(q=hits,lambda=2)
qplot(x=hits,y=myVec,geom="col",color=I("black"),fill=I("goldenrod"))


# for a Poisson distribution with lambda=2, 
# what is the probability of getting 1 or fewer hits?

ppois(q=1, lambda=2)


# We could also get this through dpois
p_0 <- dpois(x=0,lambda=2)
p_0
p_1 <- dpois(x=1,lambda=2)
p_1
p_0 + p_1


# The q function is the inverse of p
# What is the number of hits corresponding to 50% of the probability mass
qpois(p=0.5,lambda=2.5)
qplot(x=0:10,y=dpois(x=0:10,lambda=2.5),geom="col",color=I("black"),fill=I("goldenrod"))

# but distribution is discrete, so this is not exact
ppois(q=2,lambda=2.5)

# finally, we can simulate individual values from a poisson
ranPois <- rpois(n=1000,lambda=2.5)
qplot(x=ranPois,color=I("black"),fill=I("goldenrod"))


# for real or simulated data, we can use the quantile
# function to find the empirical  95% confidence interval on the data

quantile(x=ranPois,probs=c(0.025,0.975))
```

#### Binomial distribution

```{r}
#-------------------------------------------------
# Binomial distribution
# p = probability of a dichotomous outcome
# size = number of trials
# x = possible outcomes
# outcome x is bounded between 0 and number of trials

# use "d" binom for density function
hits <- 0:10
myVec <- dbinom(x=hits,size=10,prob=0.5)
qplot(x=0:10,y=myVec,geom="col",color=I("black"),fill=I("goldenrod"))



# and how does this compare to an actual simulation of 50 tosses of 100 coins?

myCoins <- rbinom(n=50,size=100,prob=0.5)
qplot(x=myCoins,color=I("black"),fill=I("goldenrod"))
quantile(x=myCoins,probs=c(0.025,0.975))

```

#### Negative Binomial distribution

```{r}
#-------------------------------------------------
# negative binomial: number of failures (values of MyVec)
# in a series of (Bernouli) with p=probability of success 
# before a target number of successes (= size)
# generates a discrete distribution that is more 
# heterogeneous ("overdispersed") than Poisson
hits <- 0:40
myVec <- dnbinom(x=hits, size=5, prob=0.5)
qplot(x=hits,y=myVec,geom="col",color=I("black"),fill=I("goldenrod"))

# geometric series is a special case where N= 1 success
# each bar is a constant fraction 1 - "prob" of the bar before it
myVec <- dnbinom(x=hits, size=1, prob=0.1)
qplot(x=hits,y=myVec,geom="col",color=I("black"),fill=I("goldenrod"))


# alternatively specify mean = mu of distribution and size, 
# the dispersion parameter (small is more dispersed)
# this gives us a poisson with a lambda value that varies
# the dispersion parameter is the shape parameter in the gamma
# as it increases, the distribution has a smaller variance
# just simulate it directly

nbiRan <- rnbinom(n=1000,size=10,mu=5)
qplot(nbiRan,color=I("black"),fill=I("goldenrod"))

nbiRan <- rnbinom(n=1000,size=0.1,mu=5)
qplot(nbiRan,color=I("black"),fill=I("goldenrod"))
```
