---
title: 'Lecture #25: dplyr'
author: "Nicholas J. Gotelli"
date: "April 19, 2018"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---
### tidyr, dplyr

#### Wrangling tools
##### Regular Expressions
- good for manipulating strings and cleaning data
- find and replace, but mostly for manipulating columns, not rows
#### Subsetting rows and columns
- create a boolean expression for selection
- apply to rows or columns to generate an n-element boolean vector
- use this vector to select and subset rows and columns

#### dplyr
- consistent set of grammar
- use to modify actual layout of the data

### tidy data
- need to be in the "long" form
- each cell is a value (= smallest object of measurement)
- each row is a unique observation
- each column is an attribute of the observation

### why not tidy?
- no consistent formatting structure
- alternate forms required for some programs
- ease of input in spreadsheets
- storage and size

### Preliminaries

```{r, message=FALSE,warning=FALSE}
library(reshape2)
library(tidyr)
library(dplyr)
library(ggplot2)
library(TeachingDemos)
char2seed("Sharpei")
```

### Build species x site matrix

```{r, message=FALSE,warning=FALSE}
species <- 5
sites <- 8
abundanceRange <- 1:10
mFill <- 0.4

vec <- rep(0,species*sites)  # set up empty vector
abun <- sample(x=abundanceRange,
               size=round(mFill*length(vec)),
               replace=TRUE) # sample abundance counts
vec[seq_along(abun)] <- abun # fill initial cells with counts
vec <- sample(vec) # reshuffle the values
aMat <- matrix(vec,nrow=species)
rownames(aMat) <- rownames(aMat, 
                           do.NULL=FALSE, 
                           prefix="Species")
colnames(aMat) <- colnames(aMat, 
                           do.NULL=FALSE, 
                           prefix="Site")
print(aMat)
```

#### Convert to long form

```{r,warning=FALSE,message=FALSE}
# using melt function from reshape2 for matrices and arrays
. <- melt(aMat)
print(.)
. <- melt(aMat,varnames=c("Species","Site"),value.name="Abundance")
print(.)

# now convert to a data frame
aFrame <- data.frame(cbind(Species=rownames(aMat),aMat))
print(aFrame)

# use gather function from tidyr for data frames
 . <- gather(aFrame,Site1:Site8,key="Site",value="Abundance")
 print(.)
 str(.)
 .$Abundance <- as.numeric(.$Abundance)
 str(.)
 aFrameL <- .
# now it is ready for a barplot in ggplot
 
 ggplot(aFrameL,aes(x=Site,y=Abundance,fill=Species)) +
   geom_bar(position="dodge",stat="identity",color="black") 
 
 
```

#### Build a Subject x Time Experimental Matrix

```{r, warning=FALSE,message=FALSE}

# construct data frame; typical excel spreadsheet format
Treatment <- rep(c("Control","Treatment"),each=5)
Subject <- 1:10
T1 <- rnorm(10)
T2 <- rnorm(10)
T3 <- rnorm(10)
eFrame <- data.frame(Treatment=Treatment,Subject=Subject,T1=T1,T2=T2,T3=T3)
str(eFrame)
print(eFrame)

# now use gather to turn this into long format
. <- gather(eFrame,T1:T3,key="Time",value="Response")
print(.)
str(.)
# set up new variable "Time" as a factor
.$Time <- as.factor(.$Time)
str(.)
eFrameL <- .
# now a quick boxplot

ggplot(eFrameL,aes(x=Treatment,y=Response,fill=Time)) +
  geom_boxplot()
```

#### Going from long back to wide formats: `recast` and `spread`

```{r,message=FALSE,warning=FALSE}
. <- dcast(aFrameL,Species~Site,value="Abundance")
str(.)
print(.)

. <- spread(aFrameL,key=Site,value=Abundance)
str(.)
print(.)

. <- spread(aFrameL,key=Species,value=Abundance)
str(.)
print(.)

. <- spread(eFrameL,key=Time,value=Response)
str(.)
print(.)
```

#### Using dplyr functions to summarize and group data

```{r, warning=FALSE,MESSAGE=FALSE}

# summarize gives column means for an entire data frame
as.data.frame(summarize(mpg,ctyM=mean(cty), ctySD=sd(cty)))

# not helpful, but almost always used with group_by function
. <- group_by(mpg,fl)
summarize(.,ctyM=mean(cty), ctySD=sd(cty))

# can also group by multiple classifications
. <- group_by(mpg,fl,class)
summarize(.,ctyM=mean(cty), ctySD=sd(cty))

# let's add in the sample size 
. <- group_by(mpg,fl,class)
summarize(.,ctyM=mean(cty), ctySD=sd(cty),n=length(cty))

# finally, let's filter and remove all of the suvs
. <- filter(mpg,class!="suv")
. <- group_by(.,fl,class)
summarize(.,ctyM=mean(cty), ctySD=sd(cty),n=length(cty))

```

#### Introducing the apply family

```
replicate(n,expression,simplify)
n : number of replications
expression : any r expression, including a function or operation
simplify : default="array", which will generate an array with dimensions 1 higher than output structure. simplify=TRUE will generate a matrix or vector. simplify=FALSE will generate a list)
```

```{r, message=FALSE,warning=FALSE}
# first set this up as a matrix

myOut <- matrix(data=0,nrow=3,ncol=5)
print(myOut)

# fill with a for loop
for (i in 1:nrow(myOut)) {
  for (j in 1:ncol(myOut)) {
   myOut[i,j] <- runif(1)
  }
}
print(myOut)

# silly to loop here. Just create myOut and fill directly
myOut <- matrix(data=runif(15),nrow=3,ncol=5)
print(myOut)

# each column of output is one of the replicates
mO <- replicate(n=5, 
                100 + runif(3),
                simplify=TRUE)
print(mO)

# try making random matrices
mO <- replicate(n=5,
                matrix(runif(6),3,2),
                simplify="array")
print(mO)       # print 3-D array
print(mO[,,2]) # print all rows and columns of second matrix
print(mO[2,1,3]) # print row 2, column1 of matrix 3

# what if the output is not all of the same length? 
#  in that case, replicate will always make a list, 
# regardless of setting for simplify
mO <- replicate(n=5, 
                100 + runif(sample.int(10,1)),
                simplify=TRUE)
print(mO)
print(mO[[3]][4]) # print the 4th element from the third vector
```

#### The `tapply` function

```
Behaves like summarize() and by_group but operates on vectors not data frames

tapply(x,index,fun)
X = vector of values to operate on
INDEX = grouping factor (elements coerced to factor)
FUN = function to be applied to the elements from x in each of the index groups
```

```{r, warning=FALSE,message=FALSE}
myMeans <- tapply(X=mpg$cty,
                  INDEX=mpg$class,
                  FUN=mean)
print(myMeans)
str(myMeans)

# can also use with implicit functions, but not expressions

myMeans <- tapply(X=mpg$cty,
                  INDEX=mpg$class,
                  FUN= function(x) mean(x)/100)
print(myMeans)
str(myMeans)
```

