---
title: 'Lecture #12: Data Entry, Experimental Designs'
author: "Nicholas J. Gotelli"
date: "February 22, 2018"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---
### Data analysis recap
1. Begin with observed data
2. Choose an appropriate distribution
  - bounded versus unbounded distributions
  - continuous versus integer variables
3. Use `fitdistr` to get MLE parameters for data
4. Examine mean and variance
5. What would mean and variances and sample sizes be for groups in your experiment?
6. Simulate a data set with these values
7. Use simple ANOVA/regression tests for p value
8. Explore effects of increasing/decreasing sample sizes
9. Explore effects of increasing/decreasing differences among mean (= effect size)

### Data Curation
- GitHub repository
- Student contract
- Metadata
- Metadata template
- example of Lauren's data

### Data Read Statement
```{r,eval=FALSE}
data <-read.table(file="FileName.csv",row.names=1,header=TRUE,sep=",",stringsAsFactors=FALSE)
```

### Omitting NA values
```{r, eval= FALSE}
dataClean <- data[complete.cases(data),] # removes all rows with NA values

# to get complete cases for just certain columns 
dataCleanCols <-data[complete.cases(data[,5:6]),]
                                   
```
 
### Archetype Data Designs
- independent versus dependent variables
- discrete versus continuous variables
- continuous variables (integer and real)
- direction of cause and effect, x axis is independent
- continuous versus discrete (natural or arbitrary or statistical bins)

### Regression (dependent: continuous, independent: continuous)
- linear model of $y = a + bx$
- statistical tests for null of hypothesis of slope and/or intercept = 0
- confidence and prediction intervals of uncertainty
- goodness of fit tests for linearity

### Set-up
```{r}
library(ggplot2)
library(TeachingDemos)
temp <- char2seed('espresso',set=FALSE)
print(temp)
set.seed(temp)
```


### Data Frame construction for Regression Data

```{r}
n = 50  # number of observations (rows)

varA <- runif(n) # random uniform values (independent)
varB <- runif(n) # a second random column (dependent)
varC <- 5.5 + varA*10 # a noisy linear relationship with varA
ID <- seq_len(n) # creates a sequence from 1:n (if n > 0!)
regData <- data.frame(ID,varA,varB,varC)
head(regData)
str(regData)
```

# Basic regression analysis in R
```{r}
# model
regModel <- lm(varB~varA,data=regData)

# model output
str(regModel)
print(regModel)
print(summary(regModel))
head(regModel$residuals)
str(summary(regModel))
z <- unlist(summary(regModel))
str(z)

```

