---
title: "Lecture #10: Data curation and dplyr (cont'd)"
author: "Lauren V. Ash"
date: "September 27, 2018"
output: 
  html_document: 
    highlight: tango
    theme: united
---
<font size="3"> 

## Data curation
* GitHub repository   
* Student contract   
* Metadata   
* [Metadata template](https://gotellilab.github.io/Bio381/Scripts/Feb02/ExcelDataTemplate.xlsx)   
* Example of Lauren's data 

## R general review
+ installing packages    
    + `install.packages("packagename")` only once, then just load library `library(packagename)`   
    + know where your library is located on your computer: `.libPaths()`     
+ projects
    + useful for organization   
    + do not have to set working directory each time    
    + create on in the top right corner of RStudio
+ writing code in R vs Rmarkdown    
    + code runs in an R Chunk in R Markdown; text is markdown language    
    + useful for creating documents that have code in them    
+ running code 
    + you can run line by line or highlight simultaneous lines
    + make sure you run the lines above first    
+ use the console when you don't need the code saved
    + i.e. `install.packages()`   
    + you can use up arrow while in the console to access previous code easily


## Exporting and importing data
### `write.table()`: useful when you want to work outside of R or with a csv file
```{r warning=F, message=F}
library(dplyr)
data(starwars)

starwars1<-select(starwars, name:species) # Let's say we want to create a new csv file with our R object
write.table(starwars1, file = "StarwarsNamesInfo.csv", row.names = F, sep = ",") # creating a csv with write.table()

```

### `read.table()`: useful when you have metadata (show example)
#### Data Read Statement Format
```{r eval=F}
data <-read.table(file="FileName.csv",row.names=1,header=TRUE,sep=",",stringsAsFactors=FALSE)

# compare to read.csv; include comment.char if you have metadata
data2<-read.csv(file="FileName.csv", row.names=1, header=TRUE, sep=",",stringsAsFactors=FALSE, comment.char = "#")
```

```{r}
data<-read.table("StarwarsNamesInfo.csv", header=T, sep = ",", stringsAsFactors = F) #our dataset has a header and the character variables are not factors
head(data) # metadata doesn't show up # use comment.char = "#" if you use read.csv()
class(data)
data<-as_tibble(data) 
glimpse(data)

```

### `saveRDS()`: useful when you are working only in R
```{r}
saveRDS(starwars1, file="StarWarsTibble") # writes a single R object to a file
```

### `readRDS()`
```{r}
sw<-readRDS("StarWarsTibble") # restores the R object
class(sw) # stays as a tibble
```


## Further into `dplyr`

### Counts 
```{r}
glimpse(sw)

# count of NAs and nonmissing values (base R)
sum(is.na(sw)) # count NAs
sum(!is.na(sw)) # count nonmissing data

swSp <- sw %>%
    group_by(species) %>%
    arrange(desc(mass))
    
# determine sample size   
swSp %>%
    summarize(avgMass = mean(mass, na.rm=T), avgHeight=mean(height, na.rm=T), n=n())

# filter out low sample size
swSp %>%
    summarize(avgMass = mean(mass, na.rm=T), avgHeight=mean(height, na.rm=T), n=n()) %>%
    filter(n >= 2)

# put in descending order
swSp %>%
    summarize(avgMass = mean(mass, na.rm=T), avgHeight=mean(height, na.rm=T), n=n()) %>%
    filter(n >= 2) %>%
    arrange(desc(n))

## Using the count helper
swSp %>%
    count(eye_color) # gives number of individuals with a given eye color (already grouped by species)

swSp %>%
    count(wt=height) # gives 'weight' (sum)

```

### Useful summary functions 
```{r}
# you can use a lot of R functions to summarize
starwarsSummary<- swSp %>%
    summarize(avgHeight=mean(height, na.rm=T), medHeight= median(height, na.rm=T), height_sd=sd(height, na.rm=T), height_IQR= IQR(height, na.rm=T), min_height=min(height, na.rm=T), first_height=first(height), n=n(), n_eyecolors = n_distinct(eye_color)) %>% #also can use last() and nth(x,#)
    filter(n >= 2) %>%
    arrange(desc(n))

starwarsSummary


```

### Grouping by Multiple variables/ ungrouping 
```{r}
# clean up
sw2<-sw[complete.cases(sw),]

sw2groups <- group_by(sw2, species, hair_color)
summarize(sw2groups, n = n())

sw3groups <- group_by(sw2, species, hair_color, gender)
summarize(sw3groups, n = n())

# ungrouping
sw3groups %>%
    ungroup() %>%
    summarize(n=n())
```

### Grouping with mutate
```{r}
# Example: standardize within groups
sw3 <- sw2 %>%
    group_by(species) %>%
    mutate(prop_height= height/sum(height)) %>%
    select(species, height, prop_height)
    
sw3

sw3 %>%
    arrange(species) # alphabetical order

```


</font>