---
title: 'Atomic Vectors I'
author: "Nicholas J. Gotelli"
date: "1 February 2024"
output:
  html_document:
    highlight: tango
    keep_md: no 
    theme: united
  pdf_document: default
---
## History of R
- Early languages FORTRAN, C
- S language in 1970s; programming plus stats, matrix algebra, graphics later
- Open-source R in 1992; first stable version in 2000

## Advantages of R
- interpreted (for interactive use)
- graphics, statistics
- very active community of contributors
- works on multiple platforms

## Disadvantages of R
- interpreted (slow speed)
- lazy evaluation
- functions hard to learn
- poorly documented
- unreliable packages
- problems with big data
   - subroutines coded for C or Fortran
   - Julia?

## R as a general programming language
- data structures and types
- writing functions
- for loops
- if..then, while statements
- annotation
- naming of variables
- using random numbers
- structured programming
- pseudocode

### Using the assignment operator

```{r, eval=FALSE}
# Using the assignment operator
x <- 5 # preferred
y = 4 # legal but not used except in function defaults
y = y + 1.1
print(y)
y <- y + 1.1
print(y)
```

# Variable names

```{r}
z <- 3 # Begin with lower case letter 
plantHeight <- 10 # option "camelCaseFormatting"
plant.height <- 4.2 # avoid periods
plant_height <- 3.3  # optimal "snake_case_formatting"
. <- 5.5 # reserve this for a generic temporary variable (more later)
```

## R's Four Data Types

Dimensions  | Homogeneous | Heterogeneous
------------- | ------------- | -------------
1-dimension  | Atomic Vector | List
2-dimensions  | Matrix | Data Frame
n-dimenions   | (array) |          |

## Types of Atomic Vectors
- character strings
- integers
- double
- integers, doubles are "numeric"
- logical
- (factor)
- vector of lists!

### One dimensional atomic vectors
```{r, eval=FALSE}
# the combine function
z <- c(3.2, 5, 5, 6) 
print(z)
typeof(z)
is.numeric(z)

# c() always "flattens" to an atomic vector
z <- c(c(3,4),c(5,6)) 
print(z)

# character strings with single or double quotes
z <- c("perch","bass",'trout') 
print(z)

# use both with an internal quote
z <- c("This is only 'one' character string", 'a second')
print(z)
typeof(z)
is.character(z)

# building logicals
# Boolean, not with quotes, all caps
z <- c(TRUE,TRUE,FALSE) 
# avoid abbreviations T, F which will work
print(z)
typeof(z)
is.logical(z)
is.integer(z)

```

### Specifying the elements of a matrix []

```{r,eval=FALSE}
# vector of character strings
dogs <- c("chow","pug","beagle","greyhound","akita")

# use number in brackets to refer to a single element in vector. First slot is "1"
dogs[1] # picks first element
dogs[5] # picks last element
dogs[6] # NA, but not an error

# pass the brackets a group of elements (= a vector) to subset the vector
dogs[c(3,5)]

# works fine with multiple repeats
dogs[c(1,1,1,4)]

# grab the entire list by leaving brackets blank (very useful with 2-D objects)
dogs[]

# pass function to calculate the element needed
length(dogs)
dogs[5]
dogs[length(dogs)]

# use negative numbers to EXCLUDE elements
dogs[-1]

# ok to use multiple negations
dogs[c(-2,-4)]
dogs[-c(2,4)]

# can't mix positive and negative elements in brackets
dogs[c(1,-5)]
```

### Three Properties of a Vector

#### Type

```{r, eval=FALSE}
z <- c(1.1, 1.2, 3, 4.4)
typeof(z) # gives type
is.numeric(z) # is. gives logical
as.character(z) # as. coerces variable
print(z)
typeof(z)

```

#### Length
```{r, eval=FALSE}
length(z) # gives number of elements
length(y) # throws error if variable does not exist

```

#### Names
```{r, eval=FALSE}
z <- runif(5)
# optional attribute not initially assigned
names(z) 
print(z)
# add names later after variable is created
names(z) <- c("chow","pug","beagle","greyhound","akita")
print(z)

# add names when variable is built (with or without quotes)
 z2 <- c(gold=3.3, silver=10, lead=2)
print(z2)

# reset names
names(z2) <- NULL

# names can be added for only a few elements
# names do not have to be distinct, but often are
names(z2) <- c("copper","zinc")
print(z2)

```

#### Special Data Types
```{r, eval=FALSE}

# NA values for missing data
z <- c(3.2,3.3,NA) # NA is a missing value
typeof(z)
length(z)
typeof(z[3]) # what is the type of third element

z1 <- NA
typeof(z1) #different NA types

is.na(z) # logical operator to find missing values
mean(z) # won't work because of NA
is.na(z)# evaluate to find missing values
!is.na(z) # use ! for NOT missing values
mean(!is.na(z)) # wrong answer based on TRUE FALSE!!
mean(z[!is.na(z)]) # correct use of indexing
#-----------------------------

# NaN, -Inf, and Inf from numeric division
z <-  0/0   # NaN
typeof(z)
z <- 1/0   # Inf
-1/0  # - Inf
#-------------------------------
# NULL is an object that is nothing!
# a reserved word in R
z <- NULL
typeof(z)
length(z)
is.null(z) # only operation that works on a null
```
### Three Notable Features of Atomic Vectors

#### Coercion

```{r, eval=FALSE}
# All atomics are of the same type
# if they are different, R coerces them
# logical -> integer -> double -> character

a <- c(2, 2.0)
print(a)
typeof(a) # technically integer coerced to numeric

b <- c("purple","green")
typeof(b)

d <- c(a,b)
print(d)
typeof(d)

# "Mistakes" in numeric variables convert to strings
# Very useful when working with logical variables

a <- runif(10)
print(a)

# Comparison operators yield a logical result
a > 0.5

# do math on a logical and it coerces to an integer!

# How many elements are greater than 0.5?
sum(a > 0.5)

# What proportion of the vector elements are greater than 0.5?

mean(a > 0.5)

#Qualifying exam question! Approximately what proportion of observations drawn from a normal (0,1) distribution are larger than 2.0?

mean(rnorm(1000) > 2)
```


#### Vectorization

```{r, eval=FALSE}
# adding a constant to a vector
z <- c(10,20,30)
z + 1

# what happens when vectors are added?

y <- c(1,2,3)
z + y

# results is an "element by element" operation on the vector
# most vector operations can be done this way

z^2

```



#### Recycling
```{r, eval=FALSE}
# but what if vector lengths are not equal?
z <- c(10,20,30)
x <- c(1,2)
z + x

# warning is issued by calculation is still made
# shorter vector is always "recycled"
# works with scalars (= vector of length(1))

```

