---
title: 'Lecture #5: Beamer Slide Presentations'
author: "Nicholas J. Gotelli"
date: "January 31, 2017"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---
### Typora
- Plain text editing demonstration
- illustrate `<CTRL \>` for hiding and showing code
- illustrate saving file to `.md` file type
- illustrate saving file to `.pdf`
- illustrate preferences and set up of `flow` code for diagram

### Using Pieces Of Different Tools

#### Pulling a style sheet from Typora for Rmarkdown
- `.css` stylesheet in Typora
- copy preferred style and subfolder to desktop
- create Rstudio new markdown and link to new `.css`

#### Creating the `.md` file from RStudio for Typora 
- now show saving of `.Rmd` to generate intermediate `.md` markdown file.
- open the new markdown file and play with it in Typora

#### "Sharing" the `.md` file between RStudio and Typora 
- open the `.md` file in both editors and move back and forth

### Beamer Slides - A Preview
- Better fonts and styles
- Avoids corporate odor of Microsoft
- Frequency dependent selection
- Cachet of "hard science" LaTeX users (computer science, physics, chemistry, stats)
- Completely stable and reliable appearance from pdf
- No need to bring your own computer

### Creating Basic Beamer Slides
- always show bullets incrementally
- choose among styles, colors, and fonts
- show `Madrid`, `beaver`, `bold` on fonts
- use two `##` to denote a new slide
- use `{r, echo=FALSE}` to just show your graphs

### Demonstrate Colors And Bullets
- Use `> - ` to override the incremental bullets
- Use `\color{gray}` to color a bullet point
- Illustrate building of red/gray bullet points in a list

### Demonstrate Adjustment of Text Placement
- Illustrate use of `\bigskip` to adjust spacing of words

### Dealing with images
- must be placed in folder and references
- use image code that is posted 

___

## History of R
- Early languages FORTRAN, C
- S language in 1970s; programming plus stats, matrix algebra, graphics later
- Open-source R in 1992; first stable version in 2000

## Advantages of R
- interpreted (for interactive use)
- graphics, statistics
- very active community of contributors
- works on multiple platforms

## Disadvantages of R
- interpreted (slow speed)
- lazy evaluation
- functions hard to learn
- poorly documented
- unreliable packages
- problems with big data
   - subroutines coded for C or Fortran
   - Julia?

## R as a general programming language
- data structures and types
- writing functions
- for loops
- if..then, while statements
- structured programming
- annotation
- naming of variables
- using random numbers
- structured programming
- pseudocode

### Using the assignment operator

```{r, eval=FALSE}
# Using the assignment operator
x <- 5 # preferred
y = 4 # legal but not used except in function defaults
y = y + 1.1
print(y)
y <- y + 1.1
print(y)
```

# Variable names

```{r}
z <- 3 # Begin with lower case letter 
plant_height <- 3.3  # a bit long
plant.height <- 4.2 # avoid periods
plantHeight <- 10 # best option "camelCaseFormatting"
```

## R's Four Data Types

Dimensions  | Homogeneous | Heterogeneous
------------- | ------------- | -------------
1-dimension  | Atomic Vector | List
2-dimensions  | Matrix | Data Frame
n-dimenions   | (array) |          |

## Types of Atomic Vectors
- character strings
- integers
- double
- integers, doubles are "numeric"
- logical
- (factor)

### One dimensional atomic vectors
```{r, eval=FALSE}
# the combine function
z <- c(3.2, 5, 5, 6) 
print(z)
typeof(z)
is.numeric(z)

# c() always "flattens" to an atomic vector
z <- c(c(3,4),c(5,6)) 
print(z)

# character strings with single or double quotes
z <- c("perch","bass",'trout') 
print(z)

# use both with an internal quote
z <- c("This is only 'one' character string", 'a second')
print(z)
typeof(z)
is.character(z)

# building logicals
# Boolean, not with quotes, all caps
z <- c(TRUE,TRUE,FALSE) 
# avoid abbreviations T, F which will work
print(z)
typeof(z)
is.logical(z)
is.integer(z)

```
### Three Properties of a Vector

#### Type

```{r, eval=FALSE}
z <- c(1.1, 1.2, 3, 4.4)
typeof(z) # gives type
is.numeric(z) # is. gives logical
as.character(z) # as. coerces variable
print(z)
typeof(z)

```

#### Length
```{r, eval=FALSE}
length(z) # gives number of elements
length(y) # throws error if variable does not exist

```

#### Names
```{r, eval=FALSE}
z <- runif(5)
# optional attribute not initially assigned
names(z) 
print(z)
# add names later after variable is created
names(z) <- c("chow","pug","beagle","greyhound","akita")
print(z)

# add names when variable is built (with or without quotes)
 z2 <- c(gold=3.3, silver=10, lead=2)
print(z2)

# reset names
names(z2) <- NULL

# names can be added for only a few elements
# names do not have to be distinct, but often are
names(z2) <- c("copper","zinc")
print(z2)

```

#### Special Data Types
```{r, eval=FALSE}

# NA values for missing data
z <- c(3.2,3.3,NA) # NA is a missing value
typeof(z)
length(z)
typeof(z[3]) # what is the type of third element

z1 <- NA
typeof(z1) #different NA types

is.na(z) # logical operator to find missing values
mean(z) # won't work because of NA
is.na(z)# evaluate to find midding values
!is.na(z) # use ! for NOT missing values
mean(!is.na(z)) # wrong answer based on TRUE FALSE!!
mean(z[!is.na(z)]) # correct use of indexing
#-----------------------------

# NaN, -Inf, and Inf from numeric division
z <-  0/0   # NaN
typeof(z)
z <- 1/0   # Inf
-1/0  # - Inf
#-------------------------------
# NULL is an object that is nothing!
# a reserved word in R
z <- NULL
typeof(z)
length(z)
is.null(z) # only operation that works on a null
```
### Three Notable Features of Atomic Vectors

#### Coercion

```{r, eval=FALSE}
# All atomics are of the same type
# if they are different, R coerces them
# logical -> integer -> double -> character

a <- c(2, 2.0)
print(a)
typeof(a) # technically integer coerced to numeric

b <- c("purple","green")
typeof(b)

d <- c(a,b)
print(d)
typeof(d)

# "Mistakes" in numeric variables convert to strings
# Very useful when working with logical variables

a <- runif(10)
print(a)

# Comparison operators yield a logical result
a > 0.5

# do math on a logical and it coerces to an integer!

# How many elements are greater than 0.5?
sum(a > 0.5)

# What proportion of the vector elements are greater than 0.5?

mean(a > 0.5)

#Qualifying exam question! Approximately what proportion of observations drawn from a normal (0,1) distribution are larger than 2.0?

mean(rnorm(1000) > 2)
```


#### Vectorization

```{r, eval=FALSE}
# adding a constant to a vector
z <- c(10,20,30)
z + 1

# what happens when vectors are added?

y <- c(1,2,3)
z + y

# results is an "element by element" operation on the vector
# most vector operations can be done this way

z^2

```



#### Recycling
```{r, eval=FALSE}
# but what if vector lengths are not equal?
z <- c(10,20,30)
x <- c(1,2)
z + x

# warning is issued by calculation is still made
# shorter vector is always "recycled"
# works with scalars (= vector of length(1))

```

### Creating Vectors
#### Creating an empty vector and adding to it (avoid)
```{r,eval=FALSE}
# create an empty vector, specify mode & length
z <- vector(mode="numeric",length=0)  
print(z)

# now add elements to it

z <- c(z,5)
print(z)
# This "dynamic sizing" is very SLOW!
```

#### Creating a vector of predefined length (preferred)
```{r, eval=FALSE}
### Better to create a vector of pre-defined length

# has 100 0s in a numeric vector
z <- rep(0,100) 
head(z)

# but would be better to start with NAs
z <- rep(NA,100)
head(z)

# but what is its type?
typeof(z)

# take advantage of coercion!
z[1] <- "Washington"
head(z)
typeof(z)
```

#### Efficiently creating a vector with many names
```{r, eval=FALSE}
MyVector <- runif(100)
MyNames <- paste("Species",seq(1:length(MyVector)),sep="")
names(MyVector) <- MyNames
str(MyVector)
```

#### Using `rep` to repeat elements and create vectors
```{r, eval=FALSE}
# rep for repeating elements
rep(0.5,6) # give the element (or vector!) and number of times to repeat
rep(x=0.5,times=6) # using the argument names is always prudent
rep(times=6,x=0.5) # with argument names, order is not important
myVec <- c(1,2,3)
rep(x=myVec,times=2) # applies to entire vectors
rep(x=myVec,each=2) # repeat each element individually
rep(x=myVec,times=myVec) # what does this do?
rep(x=myVec,each=myVec) # and this?
```

#### Using `seq` to create regular sequences
```{r,eval=FALSE}
seq(from=2, to=4) # set limits for integer sequences
2:4 # very common short cut with no explicit function wrappers
seq(from=2,to=4,by=0.5) # use a by function can generate real numbers
x <- seq(from=2, to=4,length=7) # sometimes easier to just specify the length
myVec <- 1:length(x) # commonly used, but actually slow
print(myVec)
seq_along(myVec) # much faster for models and big data!
seq_len(5) # this is also faster than 1:5
```

#### Using `rnorm` and `runif` to create vectors of random numbers
```{r,eval=FALSE}
runif(5) # 5 random uniform values betweeon 0 and 1
runif(n=3,min=100, max=101) # 3 random uniform values between 100 and 101

rnorm(6) # 6 random normal values with mean 0 and standard deviation 1
rnorm(n=5, mean=100, sd=30) # 5 random normal values with mean 100 and sd 30

## Explore distributions by sampling and plotting
library(ggplot2) # do this at the very start
z <- runif(1000) #default uniform (0,1)
qplot(x=z)
z <- rnorm(1000) #default normal (0,1)
qplot(x=z)
```

#### Using `sample` to draw random values from an existing vector

```{r,eval=FALSE}
longVec <- seq_len(10)
typeof(longVec)
str(longVec)

sample(x=longVec) # with no other params, this reorders the vector
sample(x=longVec, size=3) # specify a number (sampling without replacement)
sample(x=longVec,size=16,replace=TRUE) # can generate duplicates
myWeights <- c(rep(20,5),rep(100,5)) # create a set of non-zero positive weights (integer or real)
print(myWeights)
sample(x=longVec,replace=TRUE,prob=myWeights) # sampling with replacement and weights
sample(x=longVec,replace=FALSE,prob=myWeights) # sampling without replacement and weights
sample(x=longVec,replace=TRUE,prob=longVec) # what does this do?
```

