---
title: 'Matrices, Lists, & Data Frames'
author: "Nicholas J. Gotelli"
date: "6,11 February 2025"
output:
  html_document:
    highlight: tango
    keep_md: no
    theme: united
  pdf_document: default
---

### Matrices

```{r, eval=FALSE}
# a matrix is an atomic vector that is organized into rows and columns

my_vec <- 1:12

m <- matrix(data=my_vec,nrow=4)
print(m)

m <- matrix(data=my_vec,ncol=3)
print(m)

m <- matrix(data=my_vec,ncol=3,byrow=TRUE)
print(m)
```
### Lists

```{r,eval=FALSE}
# Lists are atomic vectors but each element 
# can hold things of different types and different sizes

my_list <- list(1:10, matrix(1:8,nrow=4,byrow=TRUE), letters[1:3],pi)
str(my_list)
print(my_list)


# using [] gives you a single item, which is of type list
my_list[4]
my_list[4] - 3 # no, can't subtract a number from a list!


# single brackets gives you only the element in that slot, which is always of type list

# to grab the object itself, use [[]]
my_list[[4]]
my_list[[4]] - 3 # now we have the contents

# if a list has 10 elements it is like a train with 10 cars
# [[5]] gives you the contents of car #5
##[c(4,5,6)] gives you a little train with cars 4, 5, and 6

# once the double bracket is called, you can access individual elements as before

my_list[[2]]
my_list[[2]][4,1]

# name list items when they are created

my_list2 <- list(tester=FALSE,little_m=matrix(1:9,nrow=3))

# named elements can be accessed with dollar sign

my_list2$little_m[2,3] # get row 2, column3
my_list2$little_m # show whole matrix
my_list2$little_m[2,] # show second row, all columns
my_list2$little_m[2] # what does this give you?

# The "unlist" strings everything back into
# a single atomic vector; coercion is used if there are mixed data types
unrolled <- unlist(my_list2)
print(unrolled)


# The most common use of list: output from a linear model is a list

library(ggplot2)
y_var <- runif(10)
x_var <- runif(10)
my_model <- lm(y_var~x_var)
qplot(x=x_var,y=y_var)

# look at output in myModel
print(my_model)

# full results are in summary
print(summary(my_model))

# drill down into summary to get what we want
str(summary(my_model))
summary(my_model)$coefficients
summary(my_model)$coefficients["x_var","Pr(>|t|)"]
summary(my_model)$coefficients[2,4]

# use unlist instead

u <- unlist(summary(my_model))
print(u)

my_slope <- u$coefficients2
my_pval <- u$coefficients8

```

### Data Frames

```{r,eval=FALSE}
# a data frame is a list of equal-lengthed vectors, each of which is a column

var_a <- 1:12
var_b <- rep(c("Con","LowN","HighN"),each=4)
var_c <- runif(12)
d_frame <- data.frame(var_a,var_b,var_c)
print(d_frame)
str(d_frame)

# add another row with rbind
# make sure you add a list, with each item corresponding to a column

# newData <- data.frame(list(varA=13,varB="HighN",varC=0.668),stringsAsFactors=FALSE)
new_data <- list(var_a=13,var_b="HighN",var_c=0.668)
print(new_data)
str(new_data)

# now bind them
d_frame <- rbind(d_frame,new_data)
str(d_frame)
tail(d_rame)


# adding another column is a little easier

#newVar <- data.frame(varD=runif(13))
new_var <- runif(13)
d_frame <- cbind(d_frame,new_var)
head(d_frame)

```

### Important Distinctions Between Lists and Matrices

```{r, eval=FALSE}

# create a matrix and data frame with same structures
z_mat <- matrix(data=1:30,ncol=3,byrow=TRUE)
z_dframe <- as.data.frame(z_mat) # coerce it

str(z_mat)    # an atomic vector with 2 dimensions
str(z_dframe) # note horizontal layout of variabes!

head(z_dframe) # note automatic variable names
head(z_mat) # note identical layout

# element referencing is the same in both
z_mat[3,3]
z_dframe[3,3]

# so is column referencing

z_mat[,3]
z_dframe[,3]
z_dframe$V3 # note use of $ and named variable column
# and row referencing
z_mat[3,]
z_dframe[3,] # note variable names and row number shown

# what happens if we reference only one dimension?

z_mat[2] # takes the second element of atomic vector (column fill)
z_dframe[2] # takes second atomic vector (= column) from list
z_dframe["V2"]
z_dframe$V2
```

### Eliminating missing values
```{r, eval=FALSE}
# use complete.cases with atomic vector
zd <- runif(10)
zd[c(5,7)] <- NA
print(zd)

complete.cases(zd)

zD[complete.cases(zd)] # clean them out

which(!complete.cases(zd)) # find NA slots

# use with a matrix

m <- matrix(1:20,nrow=5)
m[1,1] <- NA
m[5,4] <- NA
print(m)

m[complete.cases(m),] 

# now get complete cases for only certain columns!
m[complete.cases(m[,c(1,2)]),] # drops row 1
m[complete.cases(m[,c(2,3)]),] # no drops
m[complete.cases(m[,c(3,4)]),] # drops row 4
m[complete.cases(m[,c(1,4)]),] # drops 1&4

```



### Techniques for assignments and subsetting matrices and data frames
```{r,eval=FALSE}
# same principle applied to both dimensions of a matrix
m <- matrix(data=1:12,nrow=3)
dimnames(m) <- list(paste("Species",LETTERS[1:nrow(m)],sep=""),paste("Site",1:ncol(m),sep=""))
print(m)

# subsetting based on elements
m[1:2,3:4]
# same subsetting based on character strings (but no negative elements)
m[c("SpeciesA","SpeciesB"), c("Site3","Site4")]

# use blanks before or after comma to indicate full rows or columns
m[1:2, ]    

m[ ,3:4]

# use logicals for more complex subsetting

# e.g. select all columns for which the totals are > 15

# first try this logical
colSums(m) > 15
m[ , colSums(m) > 15]


# e.g. select all rows for which the row total is 22
m[rowSums(m)==22, ]

# note == for logical equal and != for logical NOT equal
m[rowSums(m)!=22, ]

# e.g., choose all rows for which numbers for site 1 are less than 3
# AND choose all columns for which the numbers for species A are less than 5

# first, try out this logical for rows
m[ ,"Site1"]<3

# add this in and select with all columns
m[m[ ,"Site1"]<3, ]

# and try this logical for columns
m["SpeciesA", ]<5

# add this in and select with all rows
m[ ,m["SpeciesA", ]<5]

# now combine both
m[m[ ,"Site1"]<3,m["SpeciesA", ]<5]

# and compare with full m
print(m)


# caution! simple subscripting to a vector changes the data type!
z <- m[1, ]
print(z)
str(z)

# to keep this as a matrix, must add the drop=FALSE option

z2 <- m[1, ,drop=FALSE]
print(z2)
str(z2)

# caution #2, always use both dimensions, or you will select a single matrix element

m2 <- matrix(data=runif(9),nrow=3)
print(m2)
m2[2, ]

# but now this will just pull the second element
m2[2]

# probably should specify row and column indicators
m2[2,1]
# also use logicals for assignments, not just subsetting
m2[m2>0.6] <- NA
print(m2)


# A few changes for working with data frames:

data <-read.csv(file="antcountydata.csv",header=TRUE,sep=",")
str(data)

# the data frame is a list of vectors, so it is set up like a matrix
data[3,2]

# you can specify just the column names

data_names <- data[c("state","county")]
str(dataNames)

# or in matrix style
data_names <- data[ ,c("county", "ecoregion")]
str(data_names)


# as before, with matrices, selecting only a single column changes it
# from a data frame to a vector
data_names <- data[ ,"county"]
str(data_names)

```
## Data curation
* GitHub repository   
* Metadata   
* [Metadata template](https://gotellilab.github.io/Bio381/Scripts/Feb02/ExcelDataTemplate.xlsx)   
* Example of Lauren's data 

## Exporting and importing data

First create a tiny data set in Excel:
```
# comments at the top
# beaucoup metadata
ID, Treatment, Biomass, Notes
1, Control, 30.3, 
2, HighN, 13.0, 
3, HighN, NA, broken scale
4, Control, 35.3,
```
- Save as .csv
- Inspect in RStudio Editor

### Use `read.table` to bring in data
```{r, eval=FALSE}
my_data <- read.table(file="path/to/data.csv",
                    header=TRUE,
                    sep=",",
                    comment.char="#")

# inspect object
str(my_data)

# now add a column
my_data$newVar <- runif(4)
head(my_data)
```

### Use `write.table` to export to a data file

```{r, eval=FALSE}

write.table(x=my_data,
            file="Path/To/OutputFileName.csv",
            HEADER=TRUE,
            sep=",")
```



But this is not a good way to save or share data objects if we are working in R. Some researchers use the `save()` function, which preserves the whole environment, but once it is restored with `load()`, the variable names cannot be changed. It is better to use `saveRDS().

### `saveRDS()`: useful when you are working only in R
```{r, eval=FALSE}
saveRDS(my_data, file="Path/To/FileName.RDS") # .RDS suffix is not required, but good for clarity
```

This only saves a single R object as a binary, but remember, you can bundle up many things into a single list!

Use `readRDS()` to restore it.

### `readRDS()`
```{r, eval=FALSE}
data_in <-readRDS("FileName.RDS") 
```

