---
title: 'Probability Distributions'
author: "Nicholas J. Gotelli"
date: "6, 18 March 2025"
output:
  html_document: 
    highlight: tango
    theme: united
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

### Probability distributions in R
#### Discrete distributions

- Poisson
    * Range: [0,$\infty$]
    * Parameters: size = number of events, rate = $\lambda$
    * Interpretation: Distribution of events that occur during a fixed time interval or sampling effort with a constant rate of independent events; resembles normal with large $\lambda$, or exponential with small $\lambda$
    
- Binomial
    * Range: [0, # of trials]
    * Parameters: size= number of trials; p = probability of positive outcome
    * Interpretation: Distribution of number of successful independent dichotomous trials, with constant p
    
- Negative Binomial
    * Range: [0, $\infty$]
    * Parameters: size=number of successes; p = probability of success
    * Interpretation: Distribution of number of failures in a series of independent Bernouli trials, each with p = probability of a success. Generates a discrete distribution that is more heterogeneous ("overdispersed") than Poisson
    
#### Continuous distributions

- Uniform
    * Range: [min,max]
    * Parameters: min = minimum boundary; max = maximum boundary
    * Interpretation: Distribution of a value that is equally likely within a specified range
    
- Normal
    * Range: [$-\infty,\infty$]
    * Parameters: mean = central tendency; SD = standard deviation
    * Interpretation: Symmetric bell-shaped curve with unbounded tails
    
- Gamma $\Gamma$
    * Range: [0,$\infty$]
    * Parameters: shape, scale
    * Interpretation: mean=$shape*scale$, variance=$shape*scale^2$; generates a variety of shapes (including normal and exponential) for positive continuous variables
    
- Beta $\beta$
    * Range: [0,1] (can be rescaled to any range by simple multiplication and addition)
    * Paramters: shape1, shape2
    * Interpretation: if shape1 and shape 2 are integers, interpret as a coin toss, with shape1 = # of successes + 1, shape2 = # of failures + 1. Gives distribution of value of p, estimated from data, which can range from exponential through uniform through normal (but all are bounded). Setting shape1 and shape2 <1 yields u-shaped distributions.

### The "grammar" of probability distributions in R
- `d` gives probability density function
- `p` gives cumulative distribution function
- `q` gives quantile function (the inverse of `p`)
- `r` gives random number generation

Combine these with the base name of the function. For example `rbinom` gives a set of random values drawn from a binomial, whereas `dnorm` gives the density function for a normal distribution. There are many probability distributions available in R, but we will discuss only 7 of them.

#### Build a plotting function for simulated values (`my_histo()`) and pdfs (`my_pdf()`)

```{r}
############################
# Function: my_histo
# Purpose: creates a ggplot histogram
# Requires: ggplot
# Input: x = a numeric vector
#        data_type= "cont" or "disc"
# Output: a ggplot histogram
############################
library(ggplot2)
my_histo <- function(x=NULL,data_type="cont") {
  if(is.null(x)) x=runif(1000)
  df <- data.frame(x=x) 

# if data are continuous bounded (0,1), adjust bins for histogram  
  if (data_type=="cont" & min(df$x) > 0 & max(df$x) < 1) {
  p1 <- ggplot(data=df) +
    aes(x=x) +
    geom_histogram(boundary=0,
                   binwidth=1/30,
                   color="black",
                   fill="goldenrod") +
    scale_x_continuous(limits=c(0,1))}  

  
# if data are continuous, but not bounded (0,1), use
# ggplot default bins
  if (data_type=="cont" & (min(df$x) < 0 | max(df$x) > 1)) {
  p1 <- ggplot(data=df) +
    aes(x=x) +
    geom_histogram(color="black",
                   fill="goldenrod")}

     
  

# if data are discrete integers, 
#  use geom_bar to create a histogram
if (data_type=="disc") {
  p1 <- ggplot(data=df) + 
    aes(x=x) +
    geom_bar(color="black",fill="goldenrod") }
  
print(p1)
} 
my_histo()
my_histo(data_type="disc",x=rpois(1000,lambda=0.2))
my_histo(data_type="cont",x=runif(1000))
my_histo(data_type="cont",x=rnorm(n=1000,mean=0,sd=1))
############################
# Function: my_pdf
# Purpose: creates a ggplot probability density function
# Requires: ggplot
# Input: x = a numeric vector of x values
#        y = pdf values calculated for each value of x
#        data_type= "cont" or "disc"
# Output: a ggplot pdf
############################
my_pdf <- function(x=NULL,y=NULL,data_type="cont") {
  if(is.null(x) | is.null(y)) {
    x=seq(from=-3,to=3,length.out=100)
    y=dnorm(x) }
  
    df <- data.frame(x=x,y=y) 
    
    # for continuous distributions, 
    # plot the line for the pdf
    if(data_type=="cont") {
      p1 <- ggplot(data=df) +
        aes(x=x,y=y) +
      geom_line() +
        geom_area(fill = "cornflower blue") } 
    
    # for discrete distributions,
    # plot a bar for the probability at each value
    if (data_type=="disc") {
      p1 <- ggplot(data=df) + 
        aes(x=x,y=y) +
        geom_col(color="black",fill="cornflower blue") }
    print(p1)
}
my_pdf()
my_x=seq(from=0,to=1,length.out=100)
my_pdf(x=my_x,y=dbeta(x=my_x,shape1=15,shape2=10))
my_pdf(x=0:10,y=dpois(x=0:10,lambda=1.1),data_type="disc")

```


#### Poisson distribution

```{r}
library(ggplot2)
library(MASS)
#-------------------------------------------------
# Poisson distribution
# Discrete X >= 0
# Random events with a constant rate lambda
# (observations per time or per unit area)
# Parameter lambda > 0

# "d" function generates probability density
hits <- 0:10
my_vec <- dpois(x=hits,lambda=1)
my_pdf(x=hits,y=my_vec,data_type="disc")

my_vec <- dpois(x=hits,lambda=2)
my_pdf(x=hits,y=my_vec,data_type="disc")

sum(my_vec)  # sum of density function = 1.0 (total area under curve)

# sum is not quite 1 because we need more elements in hits:
hits <- 0:15
my_vec <- dpois(x=hits, lambda=2)
my_pdf(x=hits,y=my_vec,data_type="disc")
sum(my_vec)
# now the sum is correct

# for a Poisson distribution with lambda=2, 
# what is the probability that a single draw will yield X=0?

dpois(x=0,lambda=2)

# "p" function generates cumulative probability density; gives the 
# "lower tail" cumulative area of the distribution

hits <- 0:10
my_vec <- ppois(q=hits,lambda=2)
my_pdf(x=hits,y=my_vec,data_type="disc")


# for a Poisson distribution with lambda=2, 
# what is the probability of getting 1 or fewer hits?

ppois(q=1, lambda=2)


# We could also get this through dpois
p_0 <- dpois(x=0,lambda=2)
p_0
p_1 <- dpois(x=1,lambda=2)
p_1
p_0 + p_1


# The q function is the inverse of p
# What is the number of hits corresponding to 50% of the probability mass
qpois(p=0.5,lambda=2.5)
my_pdf(x=0:10,y=dpois(x=0:10,lambda=2.5),data_type="disc")

# but distribution is discrete, so this is not exact
ppois(q=2,lambda=2.5)

# finally, we can simulate individual values from a poisson
ran_pois <- rpois(n=1000,lambda=2.5)
my_histo(x=ran_pois,data_type="disc")


# for real or simulated data, we can use the quantile
# function to find the empirical  95% confidence interval on the data

quantile(x=ran_pois,probs=c(0.025,0.975))
```

#### Binomial distribution

```{r}
#-------------------------------------------------
# Binomial distribution
# p = probability of a dichotomous outcome
# size = number of trials
# x = possible outcomes
# outcome x is bounded between 0 and number of trials

# use "d" binom for density function
hits <- 0:10
my_vec <- dbinom(x=hits,size=10,prob=0.5)
my_pdf(x=0:10,y=my_vec,data_type="disc")



# and how does this compare to an actual simulation of 50 tosses of 100 coins?

my_coins <- rbinom(n=50,size=100,prob=0.5)
my_histo(x=my_coins,data_type="disc")
quantile(x=my_coins,probs=c(0.025,0.975))

```

#### Negative Binomial distribution

```{r}
#-------------------------------------------------
# negative binomial: number of failures (values of MyVec)
# in a series of (Bernouli) with p=probability of success 
# before a target number of successes (= size)
# generates a discrete distribution that is more 
# heterogeneous ("overdispersed") than Poisson
hits <- 0:40
my_vec <- dnbinom(x=hits, size=5, prob=0.5)
my_pdf(x=hits,y=my_vec,data_type="disc")

# geometric series is a special case where N= 1 success
# each bar is a constant fraction 1 - "prob" of the bar before it
my_vec <- dnbinom(x=hits, size=1, prob=0.1)
my_pdf(x=hits,y=my_vec,data_type="disc")


# alternatively specify mean = mu of distribution and size, 
# the dispersion parameter (small is more dispersed)
# this gives us a poisson with a lambda value that varies
# the dispersion parameter is the shape parameter in the gamma
# as it increases, the distribution has a smaller variance
# just simulate it directly

nbi_ran <- rnbinom(n=1000,size=10,mu=5)
my_histo(x=nbi_ran,data_type="disc")


nbi_ran <- rnbinom(n=1000,size=0.1,mu=5)
my_histo(x=nbi_ran,data_type="disc")
```

#### Uniform distribution

```{r}
#-------------------------------------------------
# uniform
# params specify minimum and maximum

#runif for random data
my_histo(x=runif(n=100),data_type="cont")
my_histo(x=runif(n=1000),data_type="cont")

#-------------------------------------------------
```
#### Normal distribution

```{r}
# normal 
my_norm <- rnorm(n=100,mean=100,sd=2)
my_histo(x=my_norm,data_type="cont")


# problems with normal when mean is small but zero is not allowed.
my_norm <- rnorm(n=100,mean=2,sd=2)
my_histo(x=my_norm,data_type="cont")

summary(my_norm)
toss_zeroes <- my_norm[my_norm>0]
my_histo(x=toss_zeroes,data_type="cont")

summary(toss_zeroes)
```

#### Gamma distribution

``` {r}
#-------------------------------------------------
# gamma distribution, continuous positive values, but bounded at 0

my_gamma <- rgamma(n=100,shape=1,scale=10)
my_histo(x=my_gamma,data_type="cont")

# gamma with shape= 1 is an exponential with scale = mean

# shape <=1 gives a mode near zero; very small shape rounds to zero
my_gamma <- rgamma(n=100,shape=0.1,scale=1)
my_histo(x=my_gamma,data_type="cont")

# large shape parameters moves towards a normal
my_gamma <- rgamma(n=100,shape=20,scale=1)
my_histo(x=my_gamma,data_type="cont")

# scale parameter changes mean- and the variance!
my_histo(x=rgamma(n=100,shape=2,scale=100),data_type="cont")
my_histo(x=rgamma(n=100,shape=2,scale=10),data_type="cont")
my_histo(x=rgamma(n=100,shape=2,scale=1),data_type="cont")
my_histo(x=rgamma(n=100,shape=2,scale=0.1),data_type="cont")



# unlike the normal, the two parameters affect both mean and variance

# mean = shape*scale
# variance= shape*scale^2


```
#### Beta distribution 

```{r}
#-------------------------------------------------

# beta distribution 
# bounded at 0 and 1
# analagous to a binomial, but result is a continuous distribution of probabilities
# parameter shape1 = number of successes + 1
# parameter shape2 = number of failures + 1
# interpret these in terms of a coin you are tossing

# shape1 = 1, shape2 = 1 = "no data"
my_beta <- rbeta(n=1000,shape1=1,shape2=1)
my_histo(x=my_beta,data_type="cont")

# shape1 = 2, shape1 = 1 = "1 coin toss, comes up heads!"
my_beta <- rbeta(n=1000,shape1=2,shape2=1)
my_histo(x=my_beta,data_type="cont")
# two tosses, 1 head and 1 tail
my_beta <- rbeta(n=1000,shape1=2,shape2=2)
my_histo(x=my_beta,data_type="cont")
# two tosses, both heads
my_beta <- rbeta(n=1000,shape1=2,shape2=1)
my_histo(x=my_beta,data_type="cont")
# let's get more data
my_beta <- rbeta(n=1000,shape1=20,shape2=20)
my_histo(x=my_beta,data_type="cont")
my_beta <- rbeta(n=1000,shape1=500,shape2=500)
my_histo(x=my_beta,data_type="cont")

# if the coin is biased
my_beta <- rbeta(n=1000,shape1=1000,shape2=500)
my_histo(x=my_beta,data_type="cont")
my_beta <- rbeta(n=1000,shape1=10,shape2=5)
my_histo(x=my_beta,data_type="cont")


# shape parameters less than 1.0 give us a u-shaped distribution
my_beta <- rbeta(n=1000,shape1=0.1,shape2=0.1)
my_histo(x=my_beta,data_type="cont")

my_beta <- rbeta(n=1000,shape1=0.5,shape2=0.2)
my_histo(x=my_beta,data_type="cont")
```

### estimating paramaters from data
 `p(data|hypothesis)` is the standard approach for null hypothesis tests.

Model paramaters can be thought of as a hypothesis about the distribution of the data:  `p(data|paramaters)`

This would be a goodness of fit test for data to a particular statistical distribution.

More useful might be `p(paramaters|data)`. For a given set of data, what is the probability that a particular statistical distribution (such as a normal with a specified mean and standard deviation), provides a fit to these data?

Clearly, we want paramaters that *maximize* this probability. The paramaters that fit this definition are *maximum likelihood estimators*. 

Here is a simple example:

```{r}
library(MASS)
data <- c(100,100,104,99)
z <- fitdistr(data,"normal")
print(z)
``` 

For this procedure:

1. Select a model that is appropriate for the data
2. Use `fitdistr` to find maximum likelihood estimators for the paramaters.
3. Use those paramaters to either plot the density function or simulate new data.

Here are a few data from Lauren Ash's dissertation: snout-vent length (SVL) measurements for 6 specimens of green frogs (*Lithobates clamitans*) collected from Berlin Pond in 2016.

```{r}

frog_data <- c(30.15,26.3,27.5,22.9,27.8,26.2)

# get and print model parameters assuming a normal distribution
z<- fitdistr(frog_data,"normal")
print(z)

# plot the density function for the normal and annotate the plot with the original data
x <- 1:100
g_density <- dnorm(x=x,mean=z$estimate["mean"],sd=z$estimate["sd"])
qplot(x,g_density,geom="line") + annotate(geom="point",x=frog_data,y=0.001)   
 
# now do the same for a gamma distribution, which has a shape and rate parameter as outputs from fitdistr
z<- fitdistr(frog_data,"gamma")
print(z)

# plot the density function for the gamma and annotate the plot with the original data
x <- 1:100
g_density <- dgamma(x=x,shape=z$estimate["shape"],rate=z$estimate["rate"])
qplot(x,g_density,geom="line") + annotate(geom="point",x=frog_data,y=0.001,color="red")   
```

There is not much difference here.But now let's add a weird outlier of a tiny value and see what happens
```{r}

frog_data <- c(30.15,26.3,27.5,22.9,27.8,26.2)
outlier <- 0.050
frog_data <- c(frog_data,outlier)

# get and print model parameters assuming a normal distribution
z<- fitdistr(frog_data,"normal")
print(z)

# plot the density function for the normal and annotate the plot with the original data
x <- 1:100
g_density <- dnorm(x=x,mean=z$estimate["mean"],sd=z$estimate["sd"])
qplot(x,g_density,geom="line") + annotate(geom="point",x=frog_data,y=0.001)   
 
# now do the same for a gamma distribution, which has a shape and rate parameter as outputs from fitdistr
z<- fitdistr(frog_data,"gamma")
print(z)

# plot the density function for the gamma and annotate the plot with the original data
x <- 1:100
g_density <- dgamma(x=x,shape=z$estimate["shape"],rate=z$estimate["rate"])
qplot(x,g_density,geom="line") + annotate(geom="point",x=frog_data,y=0.001,color="red")   
```

