---
title: "Strategic Coding Practices"
author: "Nicholas J. Gotelli"
date: "2025-02-02"
output: 
  html_document: 
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval=FALSE)
```
# Code folding
* create script
* use markdown headings with 4 dashes ----
* show headers and subheaders
* show folding and outline

# Code snippets
* Tools -> edit code snippets
* m.simple_script

```
# Script description
# `r Sys.Date()`
# Nicholas J. Gotelli
```

# Organize Project: `add_folders()`

```{r}
# add_folder(folder_names=NULL)
add_folder() # create typical folders for workflow
add_folder("MyUniqueFolder") # use for your own groupings
```

* will not overwrite existing folders
* will include hidden `.keep` file for github functionality

## OriginalData 

* start with primary `.csv` files or text files 
* for large files, rasters or images give links and meta data
* never ever edit anything in original data folder!

### Embed Your MetaData In Your DataFile:` metadata_template()`

```{r}
# metadata_template(file=NULL)
metadata_template(file="OriginalData/MyData.csv")
```

* Show real-world example with Lauren Ash amphibian data (2017)
* Update with Gwen Ellis genetic analysis (2025)

## CleanedData

* copy everything from Original Data folder and only do editing here
* “dirty” data cannot be opened or read
* not all data cleaning can be easily documented
* additional cleaning and wrangling after texts are created

## Scripts
*  contains standard R scripts
*  may contain only a single MainScript.R with global variables and function calls

## Functions
*  contains scripts for user-defined functions
*  these functions may be contained in a single script
*  more portability and ease of programming to create a single script for each function

## Plots
*  contains `.jpg`, `.tiff`, other image files created with code
*  do not save images by using RStudio GUI interface
*  learn to use `ggsave()` command in ggplot2 for saving graphical output
*  use letters, not numbers, to name consecutive images or tables
*  figure_a, figure_b, table_a, table_b
*  these will be later converted to numbered tables and figures in your final manuscript

## Outputs
*  should be used only to create and store summary `.csv` files that contain contents of final tables or stats numbers to be used in manuscript
*  do not use `.csv` files to "pass data" to other parts of project
*  first set up output as a data frame before passing to function

### Store summary results in table: `data_table_template()`

```{r}
# data_table_template(data_frame=NULL,file_name=NULL)
data_table_template(data_frame=NULL,file_name="Outputs/TableA.csv")
```

### Pad zeroes in file names or variable names: `create_padded_labels()`

```{r}
# create_padded_labels <- function(n=6,
#                                 string="Toy", 
#                                 suffix=NULL)
create_padded_labels(n=10,string="Species",suffix=".txt")
```

## DataObjects

* Folder to hold a serialized data object
* Use to store intermediate results that may be difficult or time-consuming to repeat
* Do not use `.csv` files for this purpose
* Do not use `save()` or `load()` for this purpose

```{r}
x <- runif(10) # an object to save
saveRDS(object=x,
        file="DataObjects/x.rds") # save to disk
restored_x <- readRDS(file="DataObjects/x.rds") # reopen to new name

y <- rnorm(3)
z <- pi
bundle <- list(x=x,y=y,z=z) # save multiple objects in a single list
saveRDS(object=bundle,
        file="DataObjects/bundle.rds") 
restored_bundle <- readRDS(file="DataObjects/bundle.rds")
restored_bundle$y # reference named list items
restored_bundle[[3]] # reference content of item number in a list
```
## Markdown

* Use this folder to store `.Rmd` markdown scripts, local image files they may call, and markdown outputs (`.html`, `.pdf` files)

# Use Logging System

* Creates a `logfile.txt` plain text file in project root
* Store meta-data about your system
* Organize and decorate output
* Use consecutive logs to probe errors

## Create Log File 

```{r}
set_up_log()
# set_up_log(my_logfile='logfile.txt',
#            user_seed=NULL,
#            console_echo=FALSE,
#            overwrite_log=TRUE)
```

* inspect initial log for system information and seed

## Supply user-defined random number seed

```{r}
set_up_log(user_seed=100)
# inspect log then restore default set-up
set_up_log()
```



## Toggle the log console to echo log messages to screen

```{r}
echo_log_console(TRUE) # show log messages also on screen
# echo_log_console(toggle=FALSE)
```

## Basic log function `l()`

```{r}
l() # plain log entry


l('log message that is echoed to the screen')
echo_log_console(FALSE)
l('this message only shows in the log file')
l()
```

## Retain logs for debugging purposes
```{r}
set_up_log(overwrite_log=FALSE)
l('new message')
l()

set_up_log()
```

# Add an 'old school' progress bar to your `for loop`: `show_progress(bar)`

```{r}
#show_progress_bar(index=i,counter=10,increment=1)
for (k in 1:100) {
  show_progress_bar(k)
  Sys.sleep(0.075)
}
l('end of loop')

# Note that the progress bar also pinpoints errors
for (k in 1:100) {
  show_progress_bar(k)
  Sys.sleep(0.075)
  # if(k==52)print(ghost) # this throws an error!
}
l('end of loop with error')

# Adjust parameters of progress bar for longer loops
for (k in 1:1000) {
  show_progress_bar(index=k,counter=50,dot=5)
  Sys.sleep(0.0075)
}
l('end of long loop')

# Add a timer for long loops (from pracma package)
tic()
for (k in 1:10) {
  show_progress_bar(k)
  Sys.sleep(1)
}
toc()
l('end of timed loop')
```

## Use the Log Message to Interrogate Objects for Debugging
```{r}
library(pryr)
set_up_log(overwrite=FALSE)
for (i in 1:100) {
  show_progress_bar()
  l(paste('memory_used=',trunc(mem_used()/10^6),
          " MB;"," i=",i,sep=''))
  z <- runif(n=10^i)
}
```

# Coding with User-Defined Functions

## A template for user-defined functions:`build_function()`
```{r}
# build_function <- function (function_name=NULL,
#                            file_prefix=NULL,
#                            file_suffix=NULL)
build_function("fit_regression") # creates an R script template for the function
source("Functions/FitRegression.R") # sources that script to load function in memory
```

## Punctuation conventions for names
* snake_case
* camelCase
* PascalCase
* kebab-case
* SCREAMING_SNAKE_CASE

```{r}
# Open the file in Functions folder: FitRegression.R
fit_regression() # runs the new function with its default values
```

##  Anatomy of A User-Defined Function
* function name
* named input parameters
* function body
* function output (optional return() statement)

## Features of a Good Function
* Has a verb-based descriptive name
* Has few inputs (< 3)
* Does one thing in isolation
* Is short (no scrolling)
* Returns one thing (could be a list)
* Uses only data from input parameters and/or locally created variables
* Does not use global variables
* Does not (usually) create or change global variables ->>
* Sets up default values, ideally based on random number generator

## Functional Programming
* Step 1: Create Pseudocode: describe project with a list of major steps (<6)
   + Select Recipes
   + Write Shopping List
   + Buy Groceries
   + Cook Meal
   + Serve Meal
   + Clean Up

* Step 2: Each list item becomes a function
    + select_recipes()
    + write_shopping_list()
    + cook_meal()
    + serve_meal()
    + clean_up()

* Step 3: Create function templates as a batch operation

```{r}
build_function(c("select_recipes",
                 "write_shopping_list",
                 "buy_groceries",
                 "cook_meal",
                 "serve_meal",
                 "clean_up"))
```
* Step 4: Source all function templates as a batch operation

```{r}
source_batch("Functions")
```

* Step 5: Run each function template

```{r}
select_recipes()
write_shopping_list()
buy_groceries()
cook_meal()
serve_meal()
clean_up()
```

* Step 6: Create inputs and outputs for each function
* Step 7: Code and test functions separately
* Step 8: Link functions through shared inputs and outputs
* Step 9: In main program call functions, create outputs, pass inputs

```{r}
# recipe_list <- select_recipe(cook_books,websites)
# shopping_list <- write_shopping_list(recipe_list,pantry_items)
# groceries <- buy_groceries(shopping_list)
# meal <- cook_meal(recipe_list,groceries)
# dirty_dishes <- serve_meal(meal,clean_dishes,utensils)
# clean_up(dirty_dishes,utensils)
```