---
title: 'Homework #8'
author: "Nicholas J. Gotelli"
date: 'March 8, 2017'
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---

We could easily spend an entire useful semester doing nothing but learning how to program statistical analyses in R. Instead, we will quickly cover in this one homework exercise the implementation in R of 4 core kinds of statistical analyses. We will use these models to gain pratice building functions.

Variables that we measure in the world can be measured either on *continuous* or *discrete* scales. Examples of continuous variables are body mass, leaf area, and CO~2~ concentration, which can take on values that are rational numbers. Examples of discrete variables are sex, species, and genotype, in which individuals are classified into different groups or categories.

A second distinction for variables is whether they are being analyzed as the *dependent* variable, or the *independent* variable. Dependent variables are often the "Y" variable or the response variable, and independent variables are the "X" variable or predictor variable.

Note that designations of discrete/continuous and dependent/independent are classifications that we impose on nature in order to analyze data and make inferences about mechanism.

These dichotomies give us a two-way table that defines four kinds of statistical analysis:

Variable  | Dependent Continuous | Dependent Discrete
------------- | -------------| -------------
Independent Continuous  | Regression | Logistic Regression
Independent Discrete  | ANOVA | Contingency Table

Here are some minimal scraps of R code to carry out these analyses. I have not annotated them, so you may have to do a bit of reading from the help system to figure some of them out. 

### Regression
```{r, eval=FALSE}
# data
xVar <- 1:10
yVar <- runif(10)
dataFrame <- data.frame(xVar,yVar)

# model
regModel <- lm(yVar~xVar,data=dataFrame)

# model output
print(regModel)
print(summary(regModel))

# plot
plot(y=dataFrame$yVar,x=dataFrame$xVar,pch=21,bg="lightblue",cex=2)
abline(regModel)
```

### ANOVA
```{r, eval=FALSE}
# data
xVar <- as.factor(rep(c("Control","Heated","Cooled"),each=5))
yVar <- c(rgamma(10,shape=5,scale=5),rgamma(5,shape=5,scale=10))
dataFrame <- data.frame(xVar,yVar)

# model
anovaModel <- aov(yVar~xVar,data=dataFrame)

# model output
print(anovaModel)
summary(anovaModel)

# plot
boxplot(yVar~xVar,data=dataFrame,col=c("grey","thistle","orchid"))

```

### Contingency Table Analysis
```{r, eval=FALSE}
# data
vec1 <- c(50,66,22)
vec2 <- c(120,22,30)
dataMatrix <- rbind(vec1,vec2)
rownames(dataMatrix) <- c("Cold","Warm")
colnames(dataMatrix) <-c("Aphaenogaster",
                         "Camponotus",
                         "Crematogaster")



# model + model output
print(chisq.test(dataMatrix))

# plot
mosaicplot(x=dataMatrix,
           col=c("goldenrod","grey","black"),
           shade=FALSE)
barplot(height=dataMatrix,
        beside=TRUE,
        col=c("cornflowerblue","tomato"))



```

### Logistic Regression
```{r, eval=FALSE}
# data
xVar <- rgamma(n=20,shape=5,scale=5)
yVar <- rbinom(n=20,size=1,p=0.5)
dataFrame <- data.frame(xVar,yVar)

# model
logRegMod <- glm(yVar ~ xVar,
                 data=dataFrame,
                 family=binomial(link="logit"))
# model output
print(logRegMod)
summary(logRegMod)

# plot
plot(x=dataFrame$xVar, y=dataFrame$yVar,pch=21,bg="tan",cex=2.5)
curve(predict(logRegMod,data.frame(xVar=x),type="response"),add=TRUE,lwd=2)
              

```
1. Set up a new markdown file for this homework. For each of the 4 models create a function to run the model. You will need to think carefully about the formal parameters for the input, the default values, and the output from your function. The output should just include model results, not any graphics
    - create the function in a single R chunk
    - immediately after the function run it with its default values
    - create a tiny fake data set (use some of the random number functions you now know about) and run your code on the fake data.

2. Now, for each of the 4 statistical models, write a graphics function that will generate a nice plot of the results. The formal parameters for input to your graphics function should be the same as the input for your corresponding stats function. Again, illustrate the graphics function for your default settings and for the tiny fake data set you created in 1.
