---
title: 'Functional Programming'
author: "Nicholas J. Gotelli"
date: "18 April 2024"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      warning=FALSE)
```

### Different function types in R
```{r}

z <- 1:10


# built-in functions ("prefix" functions)
mean(z)

# "in-fix" functions
z + 100

# equivalent "pre-fix" functions
`+`(z,100)

# user-defined functions
# --------------------------------------
# FUNCTION my_fun
# description: calculate maximum of sin of x + x
# inputs: numeric vector
# outputs: 1-element numeric vector
########################################
my_fun <- function(x=runif(5)) {

z <- max(sin(x) + x)
return(z)

} # end of my_fun
# --------------------------------------
my_fun()
my_fun(z)

# anonymous functions
# unnamed, used for simple calculations, usually with a single input, by convention called x

function(x) x + 3 # anonymous function
function(x) x + 3 (10) # try to provide input
(function(x) x + 3) (10) # use of parentheses to call
```

### Functions that call functions

```{r}
# first create some short user-defined functions
my_sum <- function(a,b) a + b
my_dif <- function(a,b) a - b
my_mult <- function(a,b) a*b

# we already know that built in functions can be called directly from within a function

funct_1 <- function(a=3,b=2) sum(a,b)
funct_1()

funct_2 <- function(a=3,b=2) my_sum(a,b)
funct_2()

funct_3 <- function(a=3,b=2) my_mult(a,b)
funct_3()
# each time we want to use a different one of the "my" functions, we have to create a new function to call it. 

# now pass data AND another function into a function as parameters:

algebra <- function(x=my_sum,a=3,b=2) x(a,b)

algebra(x=my_sum)
algebra(x=my_dif)
algebra(x=my_mult)


# This also works for passing in built-in functions
algebra(x=sum)

# but this will not work
# algebra(x=mean)

# problem: all the functions called must use same inputs
# most r functions work on a single vector
algebra_vec <- function(x=mean,a=1:10) x(a)
  algebra_vec(x=sqrt)
  algebra_vec(x=mean)
  algebra_vec(x=sd)
```



### First Task: Apply a function to each row (or column of a matrix


```{r}
m <- matrix(1:20,nrow=5,byrow=TRUE)
print(m)
```

#### `for loop` solution
```{r}
# create a vector of lists to hold the output
output <- vector("list",nrow(m))
str(output)
print(output)

# run the function in a for loop for each
# row of the matrix

for (i in seq_len(nrow(m))) {
  output[i] <- my_fun(m[i,])
}

print(output)
```

#### `apply` solution
```{r}
# using the apply function to do the same thing
# apply(X,MARGIN,FUN,...)
# X = vector or an array (= matrix)
# MARGIN 1=row,2=column, c(1,2)=rows and columns
# ... optional arguments to FUN

# apply function my_fun to each row of m
row_out <- apply(X=m,
                 MARGIN=1,
                 FUN=my_fun)
print(row_out)

# apply function my_fun to each column of m
apply(m,2,my_fun)

# apply function my_fun to each element of m
apply(m,c(1,2),my_fun)

```

#### `apply` solutions with anonymous function
```{r}
apply(m,1,function(x) max(sin(x) + x))
apply(m,2,function(x) max(sin(x) + x))
apply(m,c(1,2),function(x) max(sin(x) + x))

```

#### What happens to output of variable length?
```{r}

# first, modification of code to simply reshuffle the order of the elements
apply(m,1,sample)
# caution! array output is each vector is in a column! to preserve original matrix dimensions, we need to transpose
t(apply(m,1,sample))
  
# function to choose a random number of elements
# from each row and pick them in random order
apply(m,1,function(x) x[sample(seq_along(x),size=sample(seq_along(x),size=1))])

# thus, output from apply can be a matrix, vector, or a list!
```

### Second Task: Apply a function to every column of a data frame

```{r}
df <- data.frame(x=runif(20),y=runif(20),z=runif(20))
```

### `for loop` solution
```{r}
output <- vector("list",ncol(df))
print(output)


for (i in seq_len(ncol(df))) {
  output[i] <- sd(df[,i])/mean(df[,i])
}

print(output)
```

#### `lapply` solution
```{r}
# using lapply to do the same thing
# lapply(X,FUN,...)
# X is a vector (atomic or list)
# FUN is a function applied to each element of the list
# (note that a data frame is a list of vectors!)
# ... additional inputs to FUN

# note that output is always a list!
# note that names are retained from columns of data frame
summary_out <- lapply(X=df,
                      FUN=function(x) sd(x)/mean(x))
print(summary_out)

# sapply tries to simplify output to vector or matrix (s(implify)apply)
# vapply requires specified output formats (v(erify)apply)
# these are special cases of lapply
```

### Challenge: what if not all dataframe columns are of same type?
```{r}
treatment <- rep(c("Control","Treatment"),each=(nrow(df)/2))
print(treatment)
df2 <- cbind(treatment,df)
head(df2)
```

#### `for loop` solution
```{r}
output2 <- vector("list",ncol(df2))
for (i in seq_len(ncol(df2))) {
  if(!is.numeric(df2[,i])) next
    output2[i] <- sd(df2[,i])/mean(df2[,i])
}
print(output2)
```

#### `lapply` solution
```{r}
lapply(df2,function(x) if(is.numeric(x)) sd(x)/mean(x))
# if you wanted the output as a vector, you could 
# just unlist it:
z <- lapply(df2,function(x) if(is.numeric(x)) sd(x)/mean(x))
z <- unlist(z)

print(z) # note difference in output length!
```

### Third Task: split/apply/combine for groups in a data frame

#### `for loop` solution
```{r}
# use df2 for this, and split over two groups
print(df2)
g <- unique(df2$treatment)
print(g)
out_g <- vector("list",length(g))
names(out_g) <- g
print(out_g)
for (i in seq_along(g)){
  df_sub <- df2[df2$treatment==g[i],]
  out_g[i] <- sd(df_sub$x)/mean(df_sub$x)
}
print(out_g)
```

#### `tapply` solution
```{r}
# using tapply to do the same thing (t(agged)apply)
# tapply(X,INDEX,FUN...)
# X is a vector (atomic or list) to be subset
# index is a list of factors (or character strings) # with one or more groups
# FUN is a function applied to each element of the different subsetted groups
# ... additional inputs to FUN

z <- tapply(X=df2$x,
            INDEX=df2$treatment,
            FUN= function(x) sd(x)/mean(x))
print(z)
```

### Fourth Task: Replicate a stochastic process

```{r}
# --------------------------------------
# FUNCTION pop_gen
# description: generate a stochastic population track of varying length
# inputs: number of time steps
# outputs: population track
########################################
pop_gen <- function(z=sample.int(n=10,size=1)) {

n <- runif(z)

return(n) # note returns a numeric vector of stochastic length

} # end of pop_gen
# --------------------------------------
pop_gen()
```

#### `for loop` solution
```{r}
n_reps <- 20
list_out <- vector("list",n_reps)
for(i in seq_len(n_reps)){
  list_out[i] <- pop_gen()
}
head(list_out)
list_out[[1]]
```

#### `replicate` solution

```{r}
# using replicate to do the same thing
# replicate(n,expr)
# n is the number of times the operation is to be repeated
# expr is a function (base, or user-defined), or an expression (like an anonymous function, but without the function(x) header; just the bare code for execution).

z_out <- replicate(n=5,
                   expr=pop_gen())
print(z_out)
```

### Fifth Task: Sweep a function with all parameter combinations

```{r}
# use previous example of parameter sweep for 
# species area function S=cA^z
# this has parameters c, z, and A as inputs

# first, let's set up a data frame
# with all parameter combiinations
a_pars <- 1:10
c_pars <- c(100,150,125)
z_pars <- c(0.10,0.16,0.26,0.30)
df <- expand.grid(a=a_pars,c=c_pars,z=z_pars)
head(df)
```

#### `for loop` solution
```{r}
df_out <-cbind(df,s=NA)

for (i in seq_len(nrow(df))) {
  df_out$s[i] <- df$c[i]*(df$a[i]^df$z[i])
}
head(df_out)
```
#### `mapply` solution
```{r}
# using mapply to do the same thing (m(ultiple)apply)
# mapply(FUN,...,MoreArgs)
# FUN is the function to be used (note it is listed first!)
#...arguments to vectorize over(vectors or lists)
#MoreArgs list of additional arguments that are constant in all of the different runs
df_out$s <- mapply(FUN=function(a, c, z) c*(a^z), df$a,df$c,df$z)

head(df_out)
```

#### the "correct" solution
```{r}
# no need for loops or mapply for this simple 
# function. We can just vectorize it with a single line of code!

df_out$s <- df_out$c*(df_out$a^df_out$z)
head(df_out)