---
title: 'ggplot I'
author: "Nicholas J. Gotelli"
date: "4 April 2024"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---
### Introduction to ggplot

#### Components of a ggplot layer
1. data (in a data frame)
2. aesthetic mappings (variables are mapped to aesthetics)
3. geom (the geometric used to draw the layer; has specific aesthetics)
4. stat (takes raw data and transforms it for useful plotting)
5. position (adjustment to prevent overplotting)

#### Template for ggplot components
```
p1 <- ggplot(data= <DATA>) +
      aes(<MAPPINGS>) + 
      <GEOM_FUNCTION>(aes(<MAPPINGS>),
                      stat=<STAT>,
                      position=<POSITION>) +
                      <COORDINATE_FUNCTION> +
                      <FACET_FUNCTION>

print(p1)
ggsave(plot=p1, 
       filename="MyPlot",
       width=5,
       height=3,
       units="in",
       device="pdf")
```

#### Preliminaries

Use package [Inkscape](https://inkscape.org/)
```{r}
library(ggplot2)
library(ggthemes)
library(patchwork)

d <- mpg # use built in mpg data frame
str(d)
table(d$fl)

```

#### basic plotting

```{r,warning=FALSE,message=FALSE}

# basic histogram plot
 ggplot(data=d) +
      aes(x=hwy) +
      geom_histogram()

ggplot(data=d) +
  aes(x=hwy) +
  geom_histogram(fill="khaki",color="black")


# # basic density plot
ggplot(data=d) +
  aes(x=hwy) +
  geom_density(fill="mintcream",color="blue")


# basic scatter plot
ggplot(data=d) +
  aes(x=displ,y=hwy) +
  geom_point() +
  geom_smooth() +
  geom_smooth(method="lm",col="red")

# add a smoother
ggplot(data=d) +
  aes(x=displ,y=hwy) +
  geom_point() +
  geom_smooth()

# add a linear regression line
ggplot(data=d) +
  aes(x=displ,y=hwy) +
  geom_point() +
  geom_smooth(method = "lm",col="red")


# basic boxplot
ggplot(data=d) +
  aes(x=fl, y=cty) +
  geom_boxplot()

# basic boxplot
ggplot(data=d) +
  aes(x=fl, y=cty) +
  geom_boxplot(fill="thistle")





# qplot(x=d$fl, y=d$cty,geom="boxplot",fill=I("green"))       

# basic barplot (long format)
# qplot(x=d$fl,geom="bar",fill=I("green"))
# qplot(x=d$fl,geom="bar",fill="green")

# bar plot with specified counts or meansw
# xTreatment <- c("Control","Low","High")
# yResponse <- c(12,2.5,22.9)
# qplot(x=xTreatment,y=yResponse,geom="col",fill=I(c("red","green","blue")))
# basic curves and functions
# myVec <- seq(1,100,by=0.1)
# myFun <- function(x) sin(x) + 0.1*x
# qplot(x=myVec,y=sin(myVec),geom="line") # built in functions
# qplot(x=myVec,y=dgamma(myVec,shape=5, scale=3),geom="line") # stats d functions
# qplot(x=myVec,y=myFun(myVec),geom="line") # user-defined functions

```

### Themes and fonts

```{r,warning=FALSE,message=FALSE}
 p1 <- ggplot(data=d, mapping=aes(x=displ,y=cty)) + geom_point()
 print(p1)
 
 p1 + theme_classic() # no grid lines
 p1 + theme_linedraw() # black frame
 p1 + theme_dark() # good for brightly colored points
 p1 + theme_base() # mimics base R
 p1 + theme_par() # matches current par settings in base
 p1 + theme_void() # shows data only
 p1 + theme_solarized() # good for web pages
 p1 + theme_economist() # many specialized themes
 p1 + theme_grey() # ggplots default theme
``` 

### Major theme modifications 

```{r, warning=FALSE,message=FALSE}
 # use theme parameters to modify font and font size
 p1 + theme_classic(base_size=40,base_family="serif")
 # defaults: theme_grey, base_size=16,base_family="Helvetica")
 # font families (Mac): Times, Ariel, Monaco, Courier, Helvetica, serif,sans
 
# use coordinate_flip to invert entire plot
 p2 <- ggplot(data=d, mapping=aes(x=fl,fill=fl)) + geom_bar()
 print(p2)
 p2 + coord_flip() + theme_grey(base_size=20,base_family="sans")
```

### Minor theme modifications
```{r, warning=FALSE, message=FALSE}

 p1 <- ggplot(data=d, mapping=aes(x=displ,y=cty)) + 
 geom_point(size=7,shape=21,color="black",fill="steelblue") +
 labs(title="My graph title here",
      subtitle="An extended subtitle that will print below the main title",
      x="My x axis label",
      y="My y axis label") +
      xlim(0,4) + ylim(0,20)
 print(p1)
 
```

