---
title: 'Lecture #21: ggplot I'
author: "Nicholas J. Gotelli"
date: "20 April 2021"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---
### Introduction to ggplot

#### Components of a ggplot layer
1. data (in a data frame)
2. aesthetic mappings (variables are mapped to aesthetics)
3. geom (the geometric used to draw the layer; has specific aesthetics)
4. stat (takes raw data and transforms it for useful plotting)
5. position (adjustment to prevent overplotting)

#### Template for ggplot components
```
p1 <- ggplot(data= <DATA>,mapping=aes(<MAPPINGS>) + 
<GEOM_FUNCTION>(aes(<MAPPINGS>),
                stat=<STAT>,
                position=<POSITION>) +
                <COORDINATE_FUNCTION> +
                <FACET_FUNCTION>

print(p1)
ggsave(plot=p1, filename="MyPlot",width=5,height=3,units="in",device="pdf")
```

#### Preliminaries
```{r}
library(ggplot2)
library(ggthemes)
library(patchwork)

d <- mpg # use built in mpg data frame
str(d)
table(d$fl)

```

#### use `qplot` for fast plotting while coding

```{r,warning=FALSE,message=FALSE}
# basic histogram
qplot(x=d$hwy)
qplot(x=d$hwy,fill=I("khaki"),color=I("black"))

# make your own custom histogram function
my_histo <- function(x_var,fil_col="goldenrod"){
qplot(x=x_var,color=I("black"),fill=I(fil_col))}
my_histo(d$hwy)
my_histo(d$hwy,"thistle")
my_histo(x_var=d$hwy,fil_col="sienna")
# basic density plot
qplot(x=d$hwy,geom="density")

# basic scatter plot
qplot(x=d$displ,y=d$hwy,geom=c("smooth","point"))
qplot(x=d$displ,y=d$hwy,geom=c("smooth","point"),method="lm")

# basic boxplot
qplot(x=d$fl, y=d$cty,geom="boxplot",fill=I("green"))       

# basic barplot (long format)
qplot(x=d$fl,geom="bar",fill=I("green"))
qplot(x=d$fl,geom="bar",fill="green")

# bar plot with specified counts or meansw
x_treatment <- c("Control","Low","High")
y_response <- c(12,2.5,22.9)
qplot(x=x_treatment,y=y_response,geom="col",fill=I(c("red","green","blue")))
# basic curves and functions
my_vec <- seq(1,100,by=0.1)
my_fun <- function(x) sin(x) + 0.1*x
qplot(x=my_vec,y=sin(my_vec),geom="line") # built in functions
qplot(x=my_vec,y=dgamma(my_vec,shape=5, scale=3),geom="line") # stats d functions
qplot(x=my_vec,y=my_fun(my_vec),geom="line") # user-defined functions

```

### Themes and fonts

```{r,warning=FALSE,message=FALSE}
 p1 <- ggplot(data=d, mapping=aes(x=displ,y=cty)) + geom_point()
 print(p1)
 
 p1 + theme_bw() # good with grid lines
 p1 + theme_classic() # no grid lines
 p1 + theme_linedraw() # black frame
 p1 + theme_dark() # good for brightly colored points
 p1 + theme_base() # mimics base R
 p1 + theme_par() # matches current par settings in base
 p1 + theme_void() # shows data only
 p1 + theme_solarized() # good for web pages
 p1 + theme_economist() # many specialized themes
 p1 + theme_grey() # ggplots default theme
``` 

### Major theme modifications 

```{r, warning=FALSE,message=FALSE}
 # use theme parameters to modify font and font size
 p1 + theme_classic(base_size=40,base_family="serif")
 # defaults: theme_grey, base_size=16,base_family="Helvetica")
 # font families (Mac): Times, Ariel, Monaco, Courier, Helvetica, serif,sans
 
# use coordinate_flip to invert entire plot
 p2 <- ggplot(data=d) +
       aes(x=fl,fill=fl) + 
       geom_bar()
 print(p2)
 p2 + coord_flip() + theme_grey(base_size=20,base_family="sans")
```

### Minor theme modifications
```{r, warning=FALSE, message=FALSE}

 p1 <- ggplot(data=d) +
       aes(x=displ,y=cty) + 
       geom_point(size=7,
                  shape=21,
                  color="black",
                  fill="steelblue") +
       labs(title="My graph title here",
            subtitle="An extended subtitle that will print below the main title",
            x="My x axis label",
            y="My y axis label") +
            xlim(0,4) + ylim(0,20)
 print(p1)
 
```

