---
title: 'Lecture #7: Data Structures'
author: "Nicholas J. Gotelli"
date: "September 18, 2018"
output:
  html_document:
    highlight: tango
    keep_md: yes
    theme: united
  pdf_document: default
---
### Scheduling changes for next week

### List serves
- R-bloggers
- Sharp Sight Labs
- Data Camp

### Creating Vectors
#### Creating an empty vector and adding to it (avoid)
```{r,eval=FALSE}
# create an empty vector, specify mode & length
z <- vector(mode="numeric",length=0)  
print(z)

# now add elements to it

z <- c(z,5)
print(z)
# This "dynamic sizing" is very SLOW!
```

#### Creating a vector of predefined length (preferred)
```{r, eval=FALSE}
### Better to create a vector of pre-defined length

# has 100 0s in a numeric vector
z <- rep(0,100) 
head(z)

# but would be better to start with NAs
z <- rep(NA,100)
head(z)

# but what is its type?
typeof(z)

# take advantage of coercion!
z[1] <- "Washington"
head(z)
typeof(z)
```

#### Efficiently creating a vector with many names
```{r, eval=FALSE}
MyVector <- runif(100)
MyNames <- paste("Species",seq(1:length(MyVector)),sep="")
names(MyVector) <- MyNames
str(MyVector)
```

#### Using `rep` to repeat elements and create vectors
```{r, eval=FALSE}
# rep for repeating elements
rep(0.5,6) # give the element (or vector!) and number of times to repeat
rep(x=0.5,times=6) # using the argument names is always prudent
rep(times=6,x=0.5) # with argument names, order is not important
myVec <- c(1,2,3)
rep(x=myVec,times=2) # applies to entire vectors
rep(x=myVec,each=2) # repeat each element individually
rep(x=myVec,times=myVec) # what does this do?
rep(x=myVec,each=myVec) # and this?
```

#### Using `seq` to create regular sequences
```{r,eval=FALSE}
seq(from=2, to=4) # set limits for integer sequences
2:4 # very common short cut with no explicit function wrappers
seq(from=2,to=4,by=0.5) # use a by function can generate real numbers
x <- seq(from=2, to=4,length=7) # sometimes easier to just specify the length
myVec <- 1:length(x) # commonly used, but actually slow
print(myVec)
seq_along(myVec) # much faster for models and big data!
seq_len(5) # this is also faster than 1:5
```

#### Using `rnorm` and `runif` to create vectors of random numbers
```{r,eval=FALSE}
runif(5) # 5 random uniform values betweeon 0 and 1
runif(n=3,min=100, max=101) # 3 random uniform values between 100 and 101

rnorm(6) # 6 random normal values with mean 0 and standard deviation 1
rnorm(n=5, mean=100, sd=30) # 5 random normal values with mean 100 and sd 30

## Explore distributions by sampling and plotting
library(ggplot2) # do this at the very start
z <- runif(1000) #default uniform (0,1)
qplot(x=z)
z <- rnorm(1000) #default normal (0,1)
qplot(x=z)
```

#### Using `sample` to draw random values from an existing vector

```{r,eval=FALSE}
longVec <- LETTERS[1:10]
typeof(longVec)
str(longVec)

sample(x=longVec) # with no other params, this reorders the vector
sample(x=longVec, size=3) # specify a number (sampling without replacement)
sample(x=longVec,size=16,replace=TRUE) # can generate duplicates
myWeights <- c(rep(20,5),rep(100,5)) # create a set of non-zero positive weights (integer or real)
print(myWeights)
sample(x=longVec,replace=TRUE,prob=myWeights) # sampling with replacement and weights
sample(x=longVec,replace=FALSE,prob=myWeights) # sampling without replacement and weights
sample(x=longVec,replace=TRUE,prob=longVec) # what does this do?
```


### Matrices

```{r,eval=FALSE}
library(ggplot2)
# Create a matrix from an atomic vector
m <-matrix(data=1:12,nrow=4,ncol=3)
print(m)

# Must specify at least one of the two dimensions
m <- matrix(data=1:12,nrow=4)
print(m)

# use byrow=TRUE to fill one row at a time
m <-matrix(data=1:12,nrow=4,byrow=TRUE)
print(m)

# use dim() to see dimensions
dim(m)

# can also change dimensions of matrix (must be correct size)
dim(m) <- c(6,2)
print(m)
dim(m) <- c(4,3)
print(m)
# to get separate components of dim
nrow(m)
ncol(m)

# note that length still applies to original atomic vector
length(m)

# to add names 
rownames(m) <- c("a","b","c","d")
print(m)
colnames(m) <- LETTERS[1:ncol(m)]
print(m)
rownames(m) <- letters[nrow(m):1]
print(m)

# specify particular elements within matrix with brackets and subscripts

# print a single element
# print row 2, column 3 
print(m[2,3])
# print an entire row (or column) by leaving entry blank
# print row 2, showing all of the columns
print(m[2,])

#print column 3, showing all of the rows
print(m[,3])

#print everything
print(m)
print(m[,])
print(m[])
# (check help on LETTERS)
rownames(m) <- paste("Species",LETTERS[1:nrow(m)],sep="")
colnames(m) <- paste("Site",1:ncol(m),sep="")
print(m)

# also add names through dimnames function

dimnames(m) <- list(paste("Site",1:nrow(m),sep=""),
                    paste("Species",LETTERS[1:ncol(m)],sep=""))
print(m)

# transpose a matrix with t
m2 <- t(m)
print(m2)

# add a row to m with rbind
m2 <- rbind(m2,c(10,20,30,40))
print(m2)

# now fix the label
rownames(m2)
print(m2)
rownames(m2)[4] <- "myfix"
print(m2)

# Access individual rows and columns with names as well as indices

m2["myfix","Site3"] # equivalent to m2[4,3]
m2[c("SpeciesC","myfix"),c("Site3","Site4")] # note use of c function here!

# use cbind to add columns to the matrix

# can always change this back to a vector at any time:

myVec <- as.vector(m)
print(myVec)
```

