---
title: 'Lecture #14: Structured Programming'
author: "Nicholas J. Gotelli"
date: "February 27th, 2020"
output:
  html_document:
    highlight: tango
    theme: united
  pdf_document: default
---
### Building a Function Template

Save the following as `FunctionTemplate.txt`

```
##################################################
# FUNCTION: FunctionName
# one line description
# input: x
# output: x 
#------------------------------------------------- 
FunctionName <- function() {


 return("testing..............FunctionName")
}
```
### Pseudocode 

*pseudocode* is a simple list of instructions that gives the overall flow of the program without (yet) specifying the exact language and functions. At the most basic, it is just a list of what a program will do. Here, for example is generic pseudocode that describes about 95% of the programming chores we face:

```
Get data
Calculate stuff
Summarize output
Graph results
```
### Structured Programming With PseudoCode And Functions

Let's use the pseudocode to build a program. 

```
GetData()
CalculateStuff()
SummarizeOutput()
GraphResults()
```

Now we use our template to build 4 functions and then run them

```{r, eval=FALSE}
# Sample program to illustrate structured programming with functions
# 9 March 2017
# NJG

# All functions must be declared at the top before they are used
##################################################
# FUNCTION: GetData
# one line description
# input: x
# output: x 
#------------------------------------------------- 
GetData <- function() {


 return("testing..............GetData")
}
##################################################
# FUNCTION: CalculateStuff
# one line description
# input: x
# output: x 
#------------------------------------------------- 
CalculateStuff <- function() {


 return("testing..............CalculateStuff")
}
##################################################
# FUNCTION: SummarizeOutput
# one line description
# input: x
# output: x 
#------------------------------------------------- 
SummarizeOutput <- function() {


 return("testing..............SummarizeOutput")
}
##################################################
# FUNCTION: GraphResults
# one line description
# input: x
# output: x 
#------------------------------------------------- 
GraphResults <- function() {


 return("testing..............GraphResults")
}
##################################################
library(ggplot2)
GetData()
CalculateStuff()
SummarizeOutput()
GraphResults()

```
### Sourcing Functions In A Separate Script File
1. Cut and copy all functions to new Rscript `MyFunctions.R`
2. Create new file called `MyShortProgram.R`
3. Add following code to the new program:

```
source("MyFunctions.R")
GetData()
CalculateStuff()
SummarizeOutput()
GraphResults()
```
the `source` function runs a specified R script file automatically. It is equivalent to running `<Control><Shift><Enter>` from within a script file.

#### Advantages of Sourcing

1. Let's you build a library of reusable functions
2. Corrections can be made once and then applied to all programs that use them

#### Disadvantages of Sourcing

1. Breaks code into two separate files, both of which must be present.
2. Discourages minor tweaks to functions because they will affect other programs.

#### Compromise solution

1. Maintain a single R script that contains the functions you have built.
2. Cut and paste functions that you want from your function file to your script file.
3. When you create new functions that are useful, append a copy to your function file.

#### Building the `CalculateStuff()` Function
```{r, eval=FALSE}

##################################################
# FUNCTION: FitRegressionModel
# Fits an ordinary least squares regression model
# input: x and y vectors of numeric. Must be same length
# output: entire model summary from lm 
#------------------------------------------------- 
FitRegressionModel <- function(xVar=runif(10),yVar=runif(10)) {
dataFrame <- data.frame(xVar,yVar)
regModel <- lm(yVar~xVar,data=dataFrame)
return(summary(regModel))
}
FitRegressionModel()
```

#### Building the `SummarizeOutput()` Function

```{r, eval=FALSE}
##################################################
# FUNCTION: SummarizeOutput
# pull elements from model summary list
# input: list from summary call of lm
# output: vector of regression residuals 
#------------------------------------------------- 
SummarizeOutput <- function(z=NULL) {
if(is.null(z)) {
  z <- summary(lm(runif(10)~runif(10)))
}  
  return(z$residuals)
}
SummarizeOutput()
```

#### Building the `GraphResults()` Function

```{r, eval=FALSE}
##################################################
# FUNCTION: GraphResults
# graph data and fitted OLS line
# input: x and y vectors of numeric. Must be same length
# output: creates graph 
#------------------------------------------------- 
GraphResults <- function(xVar=runif(10), yVar=runif(10)) {
dataFrame <- data.frame(xVar,yVar)
# #plot(y=dataFrame$yVar,
#      x=dataFrame$xVar,
#      pch=21,
#      bg="lightblue",
#      cex=2)
p1 <- qplot(data=dataFrame,x=xVar,y=yVar,geom=c("smooth","point"))
print(p1)
regModel <- lm(yVar~xVar,data=dataFrame)
# abline(regModel)
message("Message: Regression graph created")
}
GraphResults()
str(GraphResults())
```

#### Building the `GetData()` Function
```{r, eval=FALSE}
##################################################
# FUNCTION: GetData
# read in a .csv file
# input: .csv file
# output: data frame
#------------------------------------------------- 
GetData <- function(fileName=NULL) {
if(is.null(fileName)) {
  dataFrame <- data.frame(ID=101:110,
                          varA=runif(10),
                          varB=runif(10))
}  else {
  dataFrame <- read.table(file=fileName,
                          header=TRUE,
                          sep=",",
                          stringsAsFactors=FALSE) 
  }
  return(dataFrame)
}


GetData()

```

#### Linking each of the functions
Now each function works, but they are all using their defaults and are not "talking" to each other. To link them, we first must create a very small number of **global variables** which can be passed down to some of our functions as parameters. In this case, we need 3 of them:

1. The name of the inputdata file
2. The column of data in the data frame that contains the x variable
3. The column of data in the data frame that contains the y variable

```{r, eval=FALSE}
# Global variables
#----------------------------------------------
antFile <- "antcountydata.csv" # New England ant county data
xCol <- 7 # column 7, latitude center of each county
yCol <- 5 # column 5, number of ant species
#----------------------------------------------

# Program body
temp1 <- GetData(fileName=antFile) # construct the data frame

x <- temp1[,xCol] # extract the predictor variable
y <- temp1[,yCol] # extract the response variable

temp2 <- FitRegressionModel(xVar=x,yVar=y) # fit the model
temp3 <- SummarizeOutput(temp2) # extract the residuals

GraphResults(xVar=x, yVar=y) # create graph

print(temp3) # show the residuals
print(temp2) # show the model summary
```

